### 2013-05-12 draft.tex
adding the most recent version to github for the first time

\documentclass{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{amsmath}
\pagestyle{empty}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{lineno}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{}{,}

\newenvironment{my_enumerate}{
\begin{enumerate}
  \setlength{\itemsep}{5pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}}{\end{enumerate}
}

\newcommand{\deltao}{$\delta^{18}$}
\newcommand{\deltac}{$\delta^{13}$}
\newcommand{\degrees}{$^{\circ}$}

\begin{document}

\subsection*{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}

\noindent Cummins, Renata C.$^1$; Finnegan, Seth$^2$; Fike, David A.$^3$; Eiler, John M.$^1$; Fischer, Woodward W.$^1$

\vspace{10pt}

\footnotesize

\noindent $^1$California Institute of Technology, Geological and Planetary Sciences, Pasadena, CA 91125

\noindent $^2$Department of Integrative Biology, University of California, 1005 Valley Life Sciences Bldg \#3140, Berkeley, CA 94720

\noindent $^3$Department of Earth and Planetary Sciences, Washington University, St. Louis, Missouri 63130

\normalsize

\linenumbers

\vspace{0.7cm} \noindent \large \textbf{Abstract} \normalsize

Much of what we know about the history of Earth's climate derives from the geologic record of carbonates. The oxygen isotopic composition (\deltao O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the \deltao O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the \deltao O of the water as well. The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set. These highly pristine samples ultimately provide a more reliable record of paleoclimate.

We tested the ability of carbonate clumped isotopes to constrain paleoclimate by measuring the clumped isotope composition of apparently well-preserved calcitic fossils from several Silurian-age (ca. 427 Ma) stratigraphic sections on the island of Gotland, Sweden. Variability in clumped isotope temperatures suggests differential preservation, complicating a resolved temperature history through these sections. Micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm ($30\pm 2$\degrees C) and similar in oxygen isotopic composition to a modern ice-free world (\deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time.

\section{Introduction}

Despite the importance of paleoclimate and more than half a century of investigations on the topic, the temperature history of Earth's oceans remains poorly constrained. The most widely-used proxy for past ocean temperatures is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate \deltao O thermometer, which was developed by Harold Urey in the 1940s \citep{Urey1947}. This carbonate \deltao O proxy has two significant sources of complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary envrionmental signature. Second, the measured \deltao O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the \deltao O of the seawater. The \deltao O of seawater can be affected by several factors, including evaporation, mixing with freshwater, and changes in the size of continental glaciers, all of which are difficult to quantify in the geologic record. The 
\deltao O of the ocean has also been hypothesized to change gradually but substantially through Earth's history \citep{Jaffres2007,Veizer1999}.

Gradual changes in the \deltao O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the \deltao O of seawater. The long-term average \deltao O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the \deltao O of ocean water by equilibrating the water with the average mantle \deltao O of 5.7$\permil$ \citep{Gregory1991}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the \deltao O of seawater by preferentially moving $^{18}$O from water into rocks. The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995,Tajika1993,Wallmann2001} and the magnitude of contintental weathering, which would lead to slow shifts in the \deltao O of the ocean. The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controvery in geochemistry.

This controversy arises from the complications associated with the carbonate \deltao O proxy, which allow conflicting interpretations of the $\sim 8\permil$ increase in the average \deltao O of marine carbonates from the Cambrian to today. Several studies have proposed that early Paleozoic carbonates may have low \deltao O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959,Keith1964,Land1995}, which tends to decrease the \deltao O of carbonate \citep{Marshall1992}. In response to this concern, \cite{Veizer1999} measured the \deltao O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics. These screened fossils also record increasing \deltao O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low \deltao O values in the early Paleozoic. An alternative hypothesis holds that oceans could have been much warmer in the early Paleozoic \citep{Karhu1986,Knauth1976}. Assuming that the \deltao O of the oceans was the same as today, the early Paleozoic oceans must have been $\sim 56$\degrees C to produce carbonates with a \deltao O$_{\text{PDB}}$ of $-8\permil$. These warm temperatures seem incompatible with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading some studies to suggest a third hypothesis, that the \deltao O$_{\text{SMOW}}$ of seawater has increased from $-8\permil$ in the Cambrian to its present value of $\sim 0\permil$ \citep{Jaffres2007,Veizer1999}.

Carbonate clumped isotope thermometry provides a unique insight into the controversy described above because it resolves the two main complications of the carbonate \deltao O proxy. The clumped isotope composition of a carbonate records a temperature that is independent of the \deltao O of water and highly sensitive to diagenetic alteration. The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates. For a shelly marine organism that lives in shallow water, the degree of clumping in the shell calcite will reflect the sea surface temperature \citep{Ghosh2006}. Measurement of the \deltao O of the shell calcite in addition to the temperature allows calculation of the \deltao O of the seawater, using the experimentally-determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}. If the calcite shell is then buried and fossilized, it is often altered in the presence of high temperatures. This alteration can take the form of recrystallization or thermally-activated re-ordering \citep{Eiler2011}. In either case, the altered material will take on a new clumped-isotope temperature that can be much higher than the temperature of the pristine fossil calcite. If the altered material is mixed with pristine fossil material during sample processing, the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived. We demonstrate here that carbonate clumped isotopes can be a more sensitive indicator of diagenetic alteration than the \deltao O or the trace metal content of a carbonate.

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected. In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease. In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the water is diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.

The present study investigates the implications of diagenetic alteration for the resolution of small temperature changes recorded in relatively well-preserved Paleozoic rocks. This study also provides a test of established proxies for diagenetic alteration as filters for clumped isotope data. Finally, our data add to the clumped isotope record of past ocean temperatures and \deltao O, including samples from a different continent and a different paleocean basin than previous clumped isotope studies.

\section{Geological setting and materials}

Samples were collected from a section exposed on the island of Gotland, Sweden, that spans the latest Llandovery through the Wenlock in the Early Silurian (Figure \ref{map}). Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}. The section records shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was within 30$^{\circ}$ of the equator during the Silurian \citep{Torsvik1992}.

The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones and were likely deposited below storm wave base (Figure \ref{column}). Fossils are present but not abundant \citep{Calner2004a,Samtleben1996}. The contact between Lower and Upper Visby beds is marked by a bed with a dense population of \textit{Phaulactis}, a species of solitary rugose coral \citep{Jeppsson1997,Jeppsson2006,Munnecke2003,Samtleben1996}. This contact is just above the Llandovery-Wenlock boundary, based on conodont stratigraphy \citep{Aldridge1993,Jeppsson1983,Mabillard1985}. During sampling, we defined this contact as a stratigraphic height of zero (Figure \ref{column}). In addition to the bed of abundant \textit{Phaulactis} corals, the contact between Lower and Upper Visby Beds is also marked by a facies change that represents a maximum flooding surface \citep{Calner2004b}. Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of detritic limestones and reef mounds. Ripple marks indicate deposition above storm wave base. Fossils are abundant in the Upper Visby Beds \citep{Calner2004a,Samtleben1996}.

Above the Upper Visby Beds, the H\"{o}gklint Formation consists of algal and crinoidal limestones with some reef mounds, deposited above storm wave base \citep{Riding1991,Samtleben1996,Watts2000}. The contact between the H\"{o}gklint Formation and the overlying Tofta Formation represents a hypothesized sequence boundary \citep{Calner2004b}. The Tofta Formation consists of oncoid-rich bedded limestone, and likely represents a shallow restricted setting, such as a backreef lagoon \citep{Riding1991,Samtleben1996}. The overlying Hangvar Formation consists of interbedded marls and limestones with some reef mounds, indicating deposition in a less restricted, deeper-water setting. The youngest strata sampled for this study were from the Slite Formation, which transitions from limestone-rich to marl-rich interbedded limestones and marls, likely representing deposition in deeper water. Fossils are present throughout the section \citep{Calner2004a}.

The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons. First, this section has long been recognized for its excellent preservation. The strata on Gotland are very nearly flat-lying, with a dip of less than 1\degrees, and display only minimal evidence of tectonic disturbance \citep{Calner2004a,Jeppsson1983}. Conodont fossils from this section are pale yellow in color, giving them a conodont alteration index (CAI) of one, which corresponds to a maximum temperature of $\sim$100\degrees C \citep{Epstein1976,Jeppsson1983,Wenzel2000}.

The Silurian section on Gotland is also attractive due to the relative abundance of fossils, especially brachiopods, throughout the section. By sampling fossil material for clumped isotope measurements, we target calcite that was first precipitated directly from seawater. All brachiopod samples were taken from the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998,Samtleben2001}. The fibrous calcite fabric characteristic of the secondary layer has been well-characterized in modern and fossil brachiopods \citep{PerezHuerta2007,Samtleben2001,Schmahl2004}, providing an expectation for pristine samples in our dataset. The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998,Brand2003,Brand2012,Mii1994,Grossman1996,Shields2003}. The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been used by other clumped isotope studies of paleotemperature \citep{Came2007,Dennis2013,Finnegan2011}. Measurements of several fossil taxa allow comparison of the preservation quality of different calcite materials.

A third attractive feature of the Silurian section on Gotland is the coupled positive excursion in \deltac C and \deltao O across the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). This excursion is called the Ireviken event, and is recognized globally \citep{Munnecke2003}. In Gotland, the 3$\permil$ excursion in \deltac C and 0.6$\permil$ excursion in \deltao O reach a maximum rate of change across the contact between the Lower and Upper Visby Beds \citep{Munnecke2003}. The excursions coincide with a shift to deposition in shallower water, but it is difficult to distinguish the relative contributions of progradation and eustatic sea level fall to this facies change \citep{Calner2004a}. The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.

The coincident isotopic excursions and stratigraphic change suggest that the Ireviken represents a climate phenomenon, the nature of which is still debated. The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998,Kaljo2003,Brand2006,Calner2008}. This hypothesis is supported by the widespread presence of glacial tillites in Brazil, dated to the latest Llandovery or earliest Wenlock \citep{Grahn1992}. Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics \citep{Bickert1997,Munnecke2003,Samtleben1996}. According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would lower the \deltao O of surface waters and also stimulate the upwelling of deep water with low \deltac C onto the shelves. Increased precipitation would also cause intense weathering, which explains the dominance of marls over carbonate reefs during this time. Then during the arid period beginning in the Upper Visby Beds, evaporation would increase the \deltao O of surface waters, and downwelling of saline water would bring only surface water with high \deltac C onto the shelves. Low terrigenous influxes would allow the growth of carbonate reefs, observed in the H\"{o}gklint Formation.

In the present study, we test the ability of clumped isotopes to resolve the changes in ocean temperature and seawater \deltao O that contributed to the 0.6$\permil$ shift in the \deltao O of carbonates through the Ireviken event. From first principles, it will be difficult for the carbonate clumped isotope thermometer to resolve these changes, because a shift of 0.6$\permil$ in \deltao O corresponds to a change in temperature of approximately 2\degrees C, which is near the limit of resolution of carbonate clumped isotopes.

\section{Methods}

\subsection{Collection and preparation of samples}

Samples were collected from the island of Gotland, Sweden during the summer of 2011. Each sample was labeled with a location, as shown in Figure \ref{map}, and a stratigraphic height. For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby Beds (Figure \ref{column}). Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}.

Most of the fossils were easily exposed by breaking the surrounding layers of poorly-lithified marl. For fossils in more strongly lithified beds, fossil material was accessed by cutting away the rock with a saw or microrotary drill. After fossils were separated from the rock, they were sonicated for two hours to remove residual sand, silt and mud from the surface of the fossils. Brachiopods, bryozoans, crinoids, gastropods and ostracods were sampled by flaking the surface of the skeleton or shell using a dental chisel. Every flake was visually inspected under a microscope at circa $25\times$ power. Flakes with visible pyrite grains or rough-textured recrystallized regions were rejected and not included in the material for analysis. For brachiopods, only the flakes preserving the characteristic fabric of the fibrous secondary layer were chosen for analysis (see Figure \ref{flakes}). Flakes were then sonicated briefly, rinsed, dried in a 50\degrees C oven, and powdered using a mortar and pestle. Only the apical ends of rugose corals were sampled, after removal of any micrite from the surface using an abrasive bit. The apical end was then cut off using a diamond microsaw, and powdered using a mortar and pestle. The apical ends of the corals were targeted because the coral septa are closely spaced together there, leaving less room for cement to precipitate between the septa. However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite. To assess diagenetic alteration, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.

\subsection{Clumped isotope analysis and standardization}

Each sample was measured between one and three times. Approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90\degrees C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate. The extraction and subsequent cleaning of the CO$_2$ followed procedures outlined by \cite{Ghosh2006}. The cleaned CO$_2$ was analyzed on a Finnigan MAT253 gas source mass spectrometer configured to collect masses 44 through 49. The sample gas was measured in eight separate acquisitions, each of which was bracketed by a measurement of reference CO$_2$ with known \deltac C and \deltao O. The \deltac C, \deltao O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample using the reference CO$_2$ as the reference material. Acquisition-to-acquisition standard deviations were on average 0.03$\permil$ for $\delta_{47}$, 0.04$\permil$ for \deltao O, and 0.02$\permil$ for \deltac C. Preliminary values for $\Delta_{47}$ and $\Delta_{48}$ of each sample were also calculated using the reference CO$_2$ as the reference material and the following equations:

\begin{align}
\Delta_{47\text{ prelim}}&=\left[\left(\frac{\text{R}^{47}_{\text{sa}}}{\text{R}^{47*}_{\text{sa}}}\frac{\text{R}^{47*}_{\text{ref}}}{\text{R}^{47}_{\text{ref}}}-1\right)-\left(\frac{\text{R}^{46}_{\text{sa}}}{\text{R}^{46*}_{\text{sa}}}\frac{\text{R}^{46*}_{\text{ref}}}{\text{R}^{46}_{\text{ref}}}-1\right)-\left(\frac{\text{R}^{45}_{\text{sa}}}{\text{R}^{45*}_{\text{sa}}}\frac{\text{R}^{45*}_{\text{ref}}}{\text{R}^{45}_{\text{ref}}}-1\right)\right]\times 1000\\
\Delta_{48\text{ prelim}}&=\left[\left(\frac{\text{R}^{48}_{\text{sa}}}{\text{R}^{48*}_{\text{sa}}}\frac{\text{R}^{48*}_{\text{ref}}}{\text{R}^{48}_{\text{ref}}}-1\right)-2\left(\frac{\text{R}^{46}_{\text{sa}}}{\text{R}^{46*}_{\text{sa}}}\frac{\text{R}^{46*}_{\text{ref}}}{\text{R}^{46}_{\text{ref}}}-1\right)\right]\times 1000
\end{align}

\noindent where R* refers to the stochastic distribution of masses calculated from the bulk isotopic composition of the sample (sa) or reference gas (ref).

Measurements were standardized by comparison to heated gases, which are CO$_2$ with two different bulk isotopic compositions that was heated at 1000\degrees C for three hours, and then cooled rapidly to room temperature. At least one heated gas was measured each day, and each sample was compared only to the heated gases that were run during the seven days before and after the date on which the sample was measured. The $\delta_{48}$ and $\Delta_{48\text{ prelim}}$ of each sample was compared to a line of best fit through a plot of $\delta_{48}$ versus $\Delta_{48}$ of the heated gases. All samples whose $\Delta_{48}$ fell more than 1$\permil$ off the line were discarded due to likely contamination. The $\Delta_{47}$ values of the remaining samples were corrected so that the $\Delta_{47}$ reflects a comparison to the stochastic heated gases, rather than the reference gas, using the following equation:

\begin{align}
\Delta_{47}=\Delta_{47\text{ prelim}}-\left(\delta^{47}\cdot\text{M}_{\text{HG}}+\text{b}_{\text{HG}}\right)\cdot\text{K}_{\text{stretch}}+\text{K}_{\text{acid}}
\end{align}

\noindent where M$_{\text{HG}}$ is the slope of the best fit line through a plot of $\delta_{47}$ versus $\Delta_{47}$ of heated gases, b$_{\text{HG}}$ is the intercept of the same line, K$_{\text{stretch}}=-0.8453$ and K$_{\text{acid}}=0.08$. Finally, the $\Delta_{47}$ of each sample was corrected based on the average offset from the accepted value of the $\Delta_{47}$ of two lab standards measured during the same week as the sample.

$\Delta_{47}$ values were converted to water temperature using the following equation \citep{Ghosh2006}:

\begin{align}
\Delta_{47}=50.0592\left(10^6\text{T}^{-2}\right)-0.02
\end{align}

\noindent where T is temperature in degrees Kelvin.

The \deltao O of water was calcuated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite \citep{Kim1997}:

\begin{align}
1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42
\end{align}

\noindent where T is temperature in degrees Kelvin.

\subsection{Electron microscopy}

Rinsed and dried brachiopod flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick. The coated flakes were imaged using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under high vacuum using the secondary electron detector. The working distance was $\sim 6$ mm and the electron high tension was 10 kV.

For electron backscatter diffraction (EBSD), 30 nm-thick uncovered, polished thin sections were obtained from cuts through the fossils. The thin sections were analyzed using the same SEM as above, under variable pressure vacuum, with an electron high tension of 20 kV. The stage was tilted 70$^{\circ}$. The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame. The step size was 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod. The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite. The diffracted beam then interacts with the EBSD detector, depdending on the orientation of the calcite crystal. The data were analyzed using software from Oxford instruments.

\subsection{Wavelength-dispersive spectroscopy}

The same thin sections as above were first coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer. Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite. Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel. Quantitative spot measurements were made using the JEOL quantitiative analysis program.

\subsection{Bulk trace metal analysis}

Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$). Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects. Solutions were sampled using a Cetac ASX 260 autosampler, aspirated into the Ar plasma using a peristaltic pump, and then analyzed on a Thermo Scientific iCAP 6000 series inductively-coupled plasma optical emission spectrometer (ICP-OES). Between each set of ten unknown samples, three standard solutions (10 ppb, 100 ppb, and 1 ppm) were measured. Reproducibility was on average 3\% for Fe, 2\% for Mn and 3\% for Sr.

\section{Results and discussion}

\subsection{Stratigraphic trends}

The \deltac C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}). We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters. The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C values of approximately -0.5$\permil$. The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.

The \deltao O values of fossil material is consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}). We were not able to resolve a \deltao O excursion from our sample set, but the range in \deltao O values of fossil material is comparable to the range observed by \cite{Bickert1997} and \cite{Munnecke2003}.

The clumped isotope temperature of samples is highly variable within beds, often covering a range of 10\degrees C within a single bed (Figure \ref{section}). The lowest temperature sample, a brachiopod, and the highest temperature sample, a rugose coral, are both from the Slite Formation, deposited well after the hypothesized climate phenomenon associated with the isotope excursions in the Lower and Upper Visby Beds. Due to the scatter in temperature, we were not able to resolve a change in temperature with time.

The large scatter in temperatures also suggests variable preservation quality among the samples. Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C. Diagenetic alteration can occur at elevated temperatures deep underground, which can increase the clumped isotope temperature of altered samples, but alteration can also occur near the Earth's surface, in the presence of cool, likely meteoric water. Consequently, low-temperature samples cannot be assumed to be more pristine than high-temperature samples. However, alteration in the presence of cool, meteoric water will deplete the \deltao O of cabonate, because meteoric water usually has a very light \deltao O. Most of our fossil samples are very similar in \deltao O, making alteration by meteoric water unlikely. This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature. We tested this hypothesis using independent methods to assess the preservation quality of representative samples.

\subsection{Independent assessments of preservation quality}

The concentrations of trace metals in fossil calcite have been widely used as indicators of recrystallization and contamination from cements and sediments. Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980,Shields2003}, so we measured the concentrations of these three elements in a large subset of our samples. The Mn and Sr concentrations of almost all of the brachiopods analyzed in this study fall within the most strictly defined limits for the Mn and Sr concentrations of pristine brachiopods (Figure \ref{Brand}). The rugose corals tend to have more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while the cements and micrites tend to have the most highly altered trace metal signatures.

The implications of the brachiopod trace metal compositions are inconsistent with the clumped isotope results. The brachiopods in this study have a range of temperatures of more than 15\degrees C, indicating variable preservation, but they would all be judged pristine or nearly pristine based on their trace metal compositions. We hesitate to extend this interpretation to rugose corals, because their trace metal composition is less well-studied than that of brachiopods. Regardless, the results in Figure \ref{Brand} imply that the carbonate clumped isotope thermometer is a more sensitive indicator of diagenetic alteration than are bulk trace metal compositions.

We further tested the sensitivity of trace metal composition to diagenetic alteration by investigating possible relationships between trace metal composition and temperature (Figure \ref{metals}). We tested brachiopods and rugose corals separately and combined with micrites and cements for the robustness of a linear correlation between trace element composition and temperature. We also tested for a linear correlation between the log of trace element composition and temperature. We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero. Only one pair of variables, the Sr concentration and temperature of rugose corals, has a p value of less than 0.05, meaning we can be 95\% confident that the correlation is real.

There are several possible explanations for the scarcity of strong correlations between trace metal concentrations and temperature. Our sample set may not include a sufficient number of highly altered samples to capture the full range in severity of alteration. Indeed, the brachiopods in this study have a much smaller range in trace metal concentrations than the brachiopods analyzed by \cite{Brand2012} (Figure \ref{Brand}). The variable temperatures of cements, depending on whether the cements formed early during shallow burial or later in the presence of high-temperature fluids, may also contribute to the lack of correlation between trace metal concentrations and temperatures.

An additional explanation for the lack of strong correlations between temperature and trace metal concentrations is provided by electron microprobe (E-probe analysis) of the trace metal signatures of recrystallization. A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B). However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D). This is only one example, but it suggests that variations in trace metal concentrations may be dominated by contamination from cements and sediments, rather than recrystallization. If recrystallization is an important contributor to the elevated temperatures of fossil material, trace metals may not be expected to be reliable indicators of diagenetic alteration.

We therefore turn to micro-scale studies of calcite textures as indicators of diagenetic alteration. Reflected-light and EBSD images of rugose corals reveal substantial amounts of late-phase diagenetic calcite, including cements and recrystallized regions (Figure \ref{EBSD_coral}A \& B). As discussed above, recrystallized regions are visible as dark colors in reflected light and aligned calcite crystals in EBSD images. Cements in Figure \ref{EBSD_coral}A \& B are characterized by large, subequant, interlocking calcite crystals. The fraction of late-phase calcite in the rugose coral samples may have been lower than suggested by the thin section in Figure \ref{EBSD_coral}, because samples were taken from the apical end of the corals where the void space between the septa was smaller, but sample preparation probably included some late-phase calcite along with pristine coral skeletal material.

In contrast to rugose corals, the EBSD image of a low-temperature brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Figure \ref{EBSD_brach}A). Within the brachiopod shell, calcite crystals are elongated approximately parallel to the surface of the shell, as expected for the fibrous secondary layer. This preferential orientation is represented by the vertical band in the stereographic projection of crystallographic planes (Figure \ref{EBSD_brach}B). No regions of recrystallization are visible within the shell. There are large interlocking cement crystals in the upper left hand corner of the EBSD image, but these formed in the void space between the two valves, and would not have been included in the sample for clumped isotope analysis.

Scanning electron microscope (SEM) images of low-temperature brachiopod flakes also demonstrate excellent preservation of original fibrous calcite textures (Figure \ref{flakes}A, B \& C). The original fibrous texture clearly contrasts with the rough recrystallized texture covering part of a flake that would not have been chosen for analysis (Figure \ref{flakes}D). The difference between these two textures is visible under the microscope used for picking brachiopod flakes, so flakes with large recrystallized regions were easily identified and excluded from the sample. However, small recrystallized regions, especially if they occurred between layers of fibrous texture, may not have been identified during picking. Therefore, although the representative flakes chosen for inspection with the SEM appear pristine, this does not guarantee that all recrystallized material was excluded from the samples.

The micro-scale studies of calcite textures in Figures \ref{EBSD_coral}--\ref{flakes} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite. Variable fractions of late-phase calcite likely contributed to the range in temperature among fossil samples. We only performed SEM and EBSD analysis on a small subset of samples, so we will not attempt to use this evidence to compare the relative preservation qualtiy of low and high-temperature samples. An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite. We therefore look to these samples as the most reliable records of ocean temperature and \deltao O during the Silurian.

\subsection{Constraining the temperature and $\mathbf{\delta^{18}}$O of Silurian oceans}

Independent assessments of preservation quality confirm that the lowest-temperature brachiopods are very well preserved. To provide additional support for this hypothesis, we measured the clumped isotope composition of the micrite adjecent to four brachiopods (Figure \ref{comparisons}). In all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod. We expect micrite to be less pristine than our brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis. Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with our hypothesis that the lowest temperature samples are the most pristine.

The analyses described above have allowed us to define the temperature consequences of diagenesis, but to confidently interpret the \deltao O of Silurian seawater, we also need to characterize the oxygen isotopic consequences of diagenesis. To constrain the effect of diagenetic alteration on the \deltao O of water calculated from the samples, we plotted the temperature versus the \deltao O of water for all of the samples in this study (Figure \ref{ellipses}). There is a clear positive relationship between the temperature and \deltao O of water, indicating that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased. This relationship is diagnostic of diagenesis occurring in the presence of a low water-to-rock ratio. In this diagenetic scheme, also called rock-buffered diagenesis, as the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. Importantly, rock-buffered diagenesis causes the temperature of more altered samples to increase, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.

The lowest, most-pristine sample in this study (Figure \ref{ellipses}) indicates that tropical Silurian oceans had a temperature of $30\pm 2$\degrees C and a \deltao O of $-1.7\pm 0.5\permil$. One caveat of these values is that they are based on only one measurement (G13 10.5 to 10.8 B1). The first time this brachiopod was picked, it produced enough material for two analyses, but one analysis was rejected because its $\Delta_{48}$ was more than 1$\permil$ above the heated gas standard line. Later, more material was picked from the same bachiopod, but in order to get enough material for a replicate measurement, very small flakes had to be included. We worried that the small flakes would be lost in the rinsing procedure, so this replicate sample was not rinsed. The temperature of the replicate sample (G13 10.5 to 10.8 B1 leftovers) was $41\pm 4$ degrees C, which may reflect contamination that ordinarily would have been removed during the rinse step of sample preparation. Due to the non-standard treatment of this replicate sample, we do not treat it as a true replicate, and we report its temperature separately.

A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates. We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples, and we are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine sample. However, the micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, suggesting that the upper estimate is likely very similar to the true value.

Our clumped isotope measurements indicate that Silurian tropical oceans were similar to modern tropical oceans in their temperature and \deltao O. Modern measurements of sea surface temperature in the Solomon Sea, east of Papua New Guinea, are 30\degrees C \citep{Reynolds1994}, matching our estimate of the temperature at a similar latitude during the Silurian. The average \deltao O of the modern ocean is approximately $-0.3\permil$ \citep{Shackleton1974}, but without ice, the average \deltao O would decrease to $-1.4\permil$ \citep{Lhomme2005}. The volume of continental glaciers is poorly constrained for the Early Silurian, but regardless, the \deltao O of Silurian oceans was likely within a few permil of the modern value.

Our results, along with those of earlier clumped isotope studies \citep{Came2007,Finnegan2011,Dennis2013}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates. Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($\sim 70$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

Given the large potential for change in the \deltao O of seawater through exchange of oxygen isotopes with crustal material, it is surprising that four separate clumped isotope studies, each from a different time in the Phanerozoic, have all implied similar \deltao O values of seawater.  Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}. The controls on the magnitudes of seafloor and continental weathering do not form a clear negative feedback loop, so we would not expect the \deltao O of seawater to be broadly constant through time, contrary to the evidence of several clumped isotope studies.

\section{Conclusions}

We used clumped isotopes to constrain the temperature and \deltao O of Silurian oceans because they provide a temperature that is independent of the \deltao O of the water from which the carbonate precipitated. Samples were collected from relatively well-preserved sections on the island of Gotland, Sweden. By sampling well-preserved, fossil-rich rocks that record a climate phenomenon thought to be associated with changing seawater temperatures, we hoped to test the limit of resolution of clumped isotopes. We were not able to resolve a change in temperature through the section, due to the variable preservation quality of the samples. Trace metal analysis was not able to detect variations in preservation among brachiopods, despite a range in brachiopod temperatures, suggesting that clumped isotopes are a more sensitive indicator of diagenetic alteration than trace metal compositions. Micro-scale analysis of calcite fabrics, comparisons between brachiopods and micrite, and the positive relationship between temperature and \deltao O of water calculated from the samples all support the hypothesis that the lowest-temperature samples are the most pristine. Based on the most pristine sample in our dataset, we estimate that tropical Silurian oceans had a temperature of  $30\pm 2$\degrees C and a \deltao O of $-1.7\pm 0.5\permil$. Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the \deltao O of seawater has remained broadly constant over time, in spite of a large potential for change through exchange of oxygen isotopes with crustal materials.

\vspace{0.7cm} \noindent \large \textbf{Acknowldgements} \normalsize

We acknowledge Nami Kitchen, Kristin Bergmann, Daniel Stolper, and the rest of the Eiler lab for help with the clumped isotope measurements. We thank Chi Ma for his assistance with the SEM and the E-probe, and Joel Hurowitz for allowing us to use his ICP-AES. We also thank Mark Garcia and David Mann for their help with sample processing. This work was supported by NSF grant EAR-1053523 and the Agouron Institute.

\newpage

\bibliographystyle{garynat}
\bibliography{draft_05_08_13}

\begin{sidewaystable}[htbp]
\centering
\small
\begin{tabular}{ | l | c | c | c | c | c | c | c | c | c | c | }
\multirow{2}{*}{Sample} & Sample & Number of & \deltac C & \deltao O & $\Delta_{47}$ & Temperature & \deltao O water & [Fe] & [Mn] & [Sr] \\
 & location & replicates & ($\permil$ PDB) & ($\permil$ PDB) & ($\permil$) & (\degrees C) & ($\permil$ SMOW) & (ppm) & (ppm) & (ppm)\\ \hline
G1 -0.3 to -1.0 M1 & 1 & 3 & 2.575 & -5.277 & 0.589 & 38.7 $\pm$ 1.6 & -0.294 $\pm$ 0.30 & 4905 $\pm$ 29 & 327 $\pm$ 3 & 238 $\pm$ 2 \\
G1 -0.3 to -1.0 R1 & 1 & 3 & 2.319 & -5.320 & 0.596 & 36.8 $\pm$ 1.8 & -0.696 $\pm$ 0.33 & 337 $\pm$ 2 & 159 $\pm$ 1 & 332 $\pm$ 2 \\
G1 -1.2 to -0.8 B1 & 1 & 1 & 2.019 & -5.523 & 0.565 & 45.0 $\pm$ 2.7 & 0.602 $\pm$ 0.49 & 182 $\pm$ 5 & 66 $\pm$ 2 & 1351 $\pm$ 33 \\
G1 -3.2 to -3.0 R1 & 1 & 1 & 2.397 & -5.851 & 0.588 & 38.9 $\pm$ 3.1 & -0.833 $\pm$ 0.57 & 1330 $\pm$ 15 & 437 $\pm$ 3 & 428 $\pm$ 3 \\
G1 -3.4 to -3.2 B2 & 1 & 1 & 2.023 & -5.597 & 0.591 & 38.2 $\pm$ 1.5 & -0.713 $\pm$ 0.27 & 177 $\pm$ 3 & 53 $\pm$ 1 & 1507 $\pm$ 9 \\
G1 2.2 R1 & 1 & 1 & 3.734 & -5.190 & 0.587 & 39.1 $\pm$ 1.4 & -0.126 $\pm$ 0.25 & 172 $\pm$ 4 & 168 $\pm$ 1 & 431 $\pm$ 5 \\
G2 -0.2 to 0.0 R1 & 2 & 1 & 2.315 & -5.233 & 0.584 & 39.8 $\pm$ 2.5 & -0.046 $\pm$ 0.47 & 379 $\pm$ 3 & 153 $\pm$ 1 & 414 $\pm$ 5 \\
G2 -0.5 to -0.25 R1 & 2 & 1 & 2.208 & -5.351 & 0.592 & 37.9 $\pm$ 2.9 & -0.514 $\pm$ 0.53 & 887 $\pm$ 14 & 157 $\pm$ 1 & 600 $\pm$ 2 \\
G2 0.0 to 0.2 B1 & 2 & 1 & 2.399 & -5.692 & 0.584 & 40.0 $\pm$ 3.7 & -0.462 $\pm$ 0.68 & 302 $\pm$ 9 & 70 $\pm$ 1 & 1289 $\pm$ 23 \\
G2 0.5 to 0.6 B1 & 2 & 1 & 3.178 & -5.123 & 0.572 & 43.0 $\pm$ 2.8 & 0.639 $\pm$ 0.51 & 46 $\pm$ 3 & 64 $\pm$ 0.4 & 1447 $\pm$ 9 \\
G2 0.5 to 0.6 B2 & 2 & 1 & 2.925 & -5.347 & 0.602 & 35.4 $\pm$ 1.2 & -0.990 $\pm$ 0.23 & 38 $\pm$ 2 & 27 $\pm$ 1 & 1372 $\pm$ 3 \\
G3 0.0 B1 & 3 & 1 & 2.248 & -5.876 & 0.573 & 42.7 $\pm$ 2.1 & -0.166 $\pm$ 0.39 & 122 $\pm$ 1 & 48 $\pm$ 0.3 & 1127 $\pm$ 8 \\
G3 LVF B1 & 3 & 2 & 1.769 & -6.402 & 0.598 & 36.4 $\pm$ 1.9 & -1.850 $\pm$ 0.38 & 124 $\pm$ 2 & 65 $\pm$ 0.4 & 1493 $\pm$ 3 \\
G3 LVF B1 cement & 3 & 2 & 2.613 & -6.395 & 0.583 & 40.1 $\pm$ 2.9 & -1.155 $\pm$ 0.54 & 953 $\pm$ 9 & 663 $\pm$ 8 & 259 $\pm$ 4 \\
G3 LVF B1 micrite & 3 & 3 & 1.768 & -5.998 & 0.581 & 40.7 $\pm$ 1.5 & -0.647 $\pm$ 0.29 & 5091 $\pm$ 14 & 457 $\pm$ 2 & 226 $\pm$ 3 \\
G3 LVF B5 & 3 & 1 & 1.792 & -5.453 & 0.610 & 33.4 $\pm$ 1.6 & -1.479 $\pm$ 0.30 &  $\pm$  &  $\pm$  &  $\pm$  \\
G3 LVF B5 micrite & 3 & 2 & 1.628 & -5.723 & 0.586 & 39.4 $\pm$ 2.7 & -0.614 $\pm$ 0.49 & 4153 $\pm$ 22 & 449 $\pm$ 3 & 170 $\pm$ 2 \\
G3 LVF R1 & 3 & 2 & 2.075 & -5.440 & 0.577 & 41.8 $\pm$ 2.7 & 0.112 $\pm$ 0.48 & 724 $\pm$ 7 & 161 $\pm$ 2 & 435 $\pm$ 4 \\
G3 LVF R2 & 3 & 2 & 5.942 & -5.368 & 0.595 & 37.2 $\pm$ 2.0 & -0.658 $\pm$ 0.37 & 69 $\pm$ 3 & 211 $\pm$ 1 & 309 $\pm$ 5 \\
G4 2.0 to 3.0 B1 & 4 & 1 & 4.223 & -6.092 & 0.580 & 40.9 $\pm$ 2.8 & -0.712 $\pm$ 0.51 &  nd  &  nd  &  nd  \\
G4 2.0 to 3.0 bryozoan1 & 4 & 1 & 4.582 & -5.604 & 0.589 & 38.6 $\pm$ 3.8 & -0.635 $\pm$ 0.71 &  nd  &  nd  &  nd  \\
G4 2.0 to 3.0 crinoid1 & 4 & 1 & 5.033 & -6.216 & 0.564 & 45.2 $\pm$ 3.3 & -0.050 $\pm$ 0.58 &  nd  &  nd  &  nd  \\
G5 1.6 to 2.2 gastropod1 & 5 & 1 & 5.589 & -6.210 & 0.598 & 36.4 $\pm$ 2.4 & -1.667 $\pm$ 0.45 &  nd  &  nd  &  nd  \\
G5 1.6 to 2.2 ostracod1 & 5 & 1 & 4.425 & -4.839 & 0.585 & 39.6 $\pm$ 3.3 & 0.309 $\pm$ 0.60 &  nd  &  nd  &  nd  \\
G6 B234 & 6 & 1 & 4.692 & -5.152 & 0.591 & 38.1 $\pm$ 1.4 & -0.286 $\pm$ 0.26 &  nd  &  nd  &  nd  \\
G6 crinoid1 & 6 & 1 & 5.457 & -6.380 & 0.535 & 53.3 $\pm$ 4.0 & 1.187 $\pm$ 0.68 &  nd  &  nd  &  nd  \\
G7 9.0 R1 & 7 & 2 & 5.336 & -5.495 & 0.575 & 42.2 $\pm$ 2.5 & 0.135 $\pm$ 0.45 & 121 $\pm$ 3 & 192 $\pm$ 1 & 86 $\pm$ 6 \\
G7 9.0 R2 & 7 & 1 & 5.173 & -5.698 & 0.551 & 48.9 $\pm$ 2.6 & 1.120 $\pm$ 0.45 &  nd  &  nd  &  nd  \\
G8 0.0 to 1.2 R1 & 8 & 1 & -0.507 & -5.208 & 0.593 & 37.7 $\pm$ 2.7 & -0.419 $\pm$ 0.50 & 187 $\pm$ 4 & 78 $\pm$ 1 & 339 $\pm$ 3 \\
G8 1.2 to 2.0 M1 & 8 & 1 & -0.768 & -7.093 & 0.561 & 46.0 $\pm$ 2.2 & -0.801 $\pm$ 0.39 & 3635 $\pm$ 13 & 248 $\pm$ 1 & 24 $\pm$ 6 \\
G8 1.2 to 2.0 R1 & 8 & 2 & -0.082 & -5.022 & 0.553 & 48.2 $\pm$ 2.7 & 1.676 $\pm$ 0.47 & 128 $\pm$ 4 & 78 $\pm$ 1 & 191 $\pm$ 3 \\
\end{tabular} 
\caption{Isotopic composition, clumped isotope temperature, calculated \deltao O of water and bulk trace metal concentrations of all the samples analyzed in this study. Errors are standard error of the mean for all acqusitions of that sample. $\text{nd}=\text{no data.}$}
\label{results}
\end{sidewaystable}

\begin{sidewaystable}[htbp]
\centering
\small
\begin{tabular}{ | l | c | c | c | c | c | c | c | c | c | c | }
\multirow{2}{*}{Sample} & Sample & Number of & \deltac C & \deltao O & $\Delta_{47}$ & Temperature & \deltao O water & [Fe] & [Mn] & [Sr] \\
 & location & replicates & ($\permil$ PDB) & ($\permil$ PDB) & ($\permil$) & (\degrees C) & ($\permil$ SMOW) & (ppm) & (ppm) & (ppm)\\ \hline
G11 0.0 to 0.2 R1 & 11 & 1 & 3.164 & -4.847 & 0.599 & 36.1 $\pm$ 2.4 & -0.352 $\pm$ 0.46 & 116 $\pm$ 4 & 140 $\pm$ 1 & 769 $\pm$ 4 \\
G11 0.2 to 0.4 B2 & 11 & 1 & 2.836 & -5.360 & 0.587 & 39.2 $\pm$ 3.6 & -0.289 $\pm$ 0.67 & 252 $\pm$ 6 & 59 $\pm$ 1 & 1424 $\pm$ 15 \\
G11 0.2 to 0.4 B6 & 11 & 1 & 2.454 & -6.146 & 0.605 & 34.6 $\pm$ 2.3 & -1.939 $\pm$ 0.43 & 89 $\pm$ 2 & 107 $\pm$ 1 & 1450 $\pm$ 8 \\
G11 0.2 to 0.4 R1 & 11 & 1 & 3.169 & -5.495 & 0.569 & 43.9 $\pm$ 2.2 & 0.440 $\pm$ 0.40 & 487 $\pm$ 7 & 346 $\pm$ 2 & 481 $\pm$ 4 \\
G11 0.4 to 0.6 B1 & 11 & 1 & 2.766 & -5.439 & 0.596 & 36.9 $\pm$ 1.6 & -0.788 $\pm$ 0.30 & 58 $\pm$ 1 & 40 $\pm$ 0.2 & 1476 $\pm$ 4 \\
G11 0.6 B1 & 11 & 1 & 2.927 & -5.269 & 0.588 & 38.9 $\pm$ 1.8 & -0.242 $\pm$ 0.33 & 59 $\pm$ 2 & 59 $\pm$ 1 & 1627 $\pm$ 11 \\
G11 0.6 to 0.8 M1 & 11 & 1 & 2.431 & -6.468 & 0.610 & 33.5 $\pm$ 2.0 & -2.472 $\pm$ 0.39 & 20948 $\pm$ 90 & 288 $\pm$ 7 & 79 $\pm$ 3 \\
G11 0.6 to 0.8 R1 & 11 & 1 & 3.447 & -4.886 & 0.596 & 36.8 $\pm$ 3.5 & -0.258 $\pm$ 0.65 & 75 $\pm$ 4 & 68 $\pm$ 1 & 503 $\pm$ 6 \\
G12 B3 & 12 & 1 & 2.803 & -5.757 & 0.558 & 46.9 $\pm$ 1.8 & 0.707 $\pm$ 0.32 &  nd  &  nd  &  nd  \\
G12 B4 & 12 & 1 & 4.397 & -5.179 & 0.581 & 40.8 $\pm$ 3.7 & 0.191 $\pm$ 0.67 & 64 $\pm$ 3 & 41 $\pm$ 1 & 1358 $\pm$ 4 \\
G12 R1 & 12 & 3 & 2.381 & -5.221 & 0.604 & 35.0 $\pm$ 1.6 & -0.935 $\pm$ 0.31 & 601 $\pm$ 6 & 124 $\pm$ 1 & 412 $\pm$ 4 \\
G12 R1 micrite & 12 & 1 & 2.102 & -5.048 & 0.568 & 44.2 $\pm$ 3.8 & 0.939 $\pm$ 0.69 &  nd  &  nd  &  nd  \\
G12 R2 & 12 & 1 & 2.043 & -5.611 & 0.582 & 40.6 $\pm$ 2.8 & -0.287 $\pm$ 0.52 & 1474 $\pm$ 18 & 417 $\pm$ 2 & 342 $\pm$ 2 \\
G13 1.0 to 2.0 R1 & 13 & 1 & -0.568 & -4.974 & 0.582 & 40.4 $\pm$ 4.4 & 0.320 $\pm$ 0.81 & 26 $\pm$ 4 & 107 $\pm$ 1 & 200 $\pm$ 3 \\
G13 10.5 to 10.8 B1 & 13 & 1 & -0.598 & -4.986 & 0.626 & 29.6 $\pm$ 2.4 & -1.741 $\pm$ 0.48 & 32 $\pm$ 5 & 16 $\pm$ 2 & 1372 $\pm$ 109 \\
G13 10.5 to 10.8 B1 leftovers & 13 & 1 & -0.722 & -5.207 & 0.579 & 41.3 $\pm$ 4.4 & 0.251 $\pm$ 0.81 &  nd  &  nd  &  nd  \\
G13 10.5 to 10.8 M1 & 13 & 1 & -1.496 & -6.298 & 0.547 & 50.0 $\pm$ 2.7 & 0.703 $\pm$ 0.46 & 3487 $\pm$ 7 & 277 $\pm$ 1 & 31 $\pm$ 4 \\
G13 10.5 to 10.8 R1 tip & 13 & 2 & -0.564 & -4.810 & 0.594 & 37.2 $\pm$ 2.0 & -0.097 $\pm$ 0.37 &  nd  &  nd  &  nd  \\
G13 10.5 to 10.8 R1 top & 13 & 2 & -1.044 & -5.829 & 0.583 & 40.3 $\pm$ 1.5 & -0.557 $\pm$ 0.28 & 571 $\pm$ 65 & 193 $\pm$ 13 & 232 $\pm$ 15 \\
G13 31.4 to 32.4 B1 & 13 & 3 & -0.316 & -5.082 & 0.592 & 37.9 $\pm$ 1.8 & -0.243 $\pm$ 0.34 & 141 $\pm$ 3 & 27 $\pm$ 0.3 & 1436 $\pm$ 11 \\
G13 31.4 to 32.4 M1 & 13 & 1 & -0.778 & -5.965 & 0.570 & 43.5 $\pm$ 2.6 & -0.109 $\pm$ 0.47 & 3671 $\pm$ 10 & 237 $\pm$ 1 & 42 $\pm$ 3 \\
G13 31.4 to 32.4 R1 & 13 & 1 & -0.809 & -5.484 & 0.535 & 53.4 $\pm$ 4.0 & 2.103 $\pm$ 0.68 & 96 $\pm$ 7 & 143 $\pm$ 1 & 118 $\pm$ 6 \\
G13 33.5 B1 bad & 13 & 1 & -0.556 & -5.253 & 0.576 & 42.1 $\pm$ 3.3 & 0.361 $\pm$ 0.61 & 107 $\pm$ 1 & 46 $\pm$ 0.3 & 1320 $\pm$ 5 \\
G13 33.5 B1 leftovers & 13 & 1 & -0.191 & -5.338 & 0.575 & 42.4 $\pm$ 3.0 & 0.321 $\pm$ 0.54 & 19 $\pm$ 5 & 14 $\pm$ 2 & 983 $\pm$ 109 \\
G13 33.5 B1 perfect & 13 & 1 & -0.159 & -5.407 & 0.596 & 36.9 $\pm$ 3.0 & -0.756 $\pm$ 0.56 &  nd  &  nd  &  nd  \\
G13 35.6 to 36.0 B1 & 13 & 1 & -0.288 & -5.397 & 0.580 & 41.0 $\pm$ 1.9 & 0.011 $\pm$ 0.35 & 39 $\pm$ 1 & 14 $\pm$ 0.1 & 1657 $\pm$ 9 \\
G13 5.0 to 6.0 B3 & 13 & 1 & -0.216 & -5.614 & 0.576 & 42.1 $\pm$ 3.5 & -0.001 $\pm$ 0.64 &  nd  &  nd  &  nd  \\
G13 5.0 to 6.0 B3 cement & 13 & 1 & -0.909 & -7.169 & 0.564 & 45.3 $\pm$ 2.2 & -0.996 $\pm$ 0.39 &  nd  &  nd  &  nd  \\
G13 6.5 to 7.5 M1 & 13 & 1 & -0.118 & -5.206 & 0.559 & 46.7 $\pm$ 3.9 & 1.229 $\pm$ 0.69 & 5708 $\pm$ 34 & 289 $\pm$ 3 & 0 $\pm$ 2 \\
G13 6.5 to 7.5 R1 & 13 & 1 & 0.180 & -4.728 & 0.536 & 53.3 $\pm$ 2.4 & 2.847 $\pm$ 0.41 & 32 $\pm$ 2 & 80 $\pm$ 1 & 204 $\pm$ 6 \\
G13 UF B1 & 13 & 3 & -0.347 & -5.416 & 0.598 & 36.4 $\pm$ 1.4 & -0.857 $\pm$ 0.28 & 27 $\pm$ 1 & 35 $\pm$ 0.5 & 1564 $\pm$ 7 \\
G13 UF B1 cement & 13 & 2 & -0.868 & -7.730 & 0.578 & 41.4 $\pm$ 2.0 & -2.259 $\pm$ 0.37 & 234 $\pm$ 2 & 266 $\pm$ 2 & 100 $\pm$ 1 \\
G13 UF B1 micrite & 13 & 2 & -1.024 & -6.330 & 0.543 & 51.1 $\pm$ 2.2 & 0.852 $\pm$ 0.38 & 1428 $\pm$ 6 & 251 $\pm$ 2 & 126 $\pm$ 2 \\
\end{tabular} 
\end{sidewaystable}

\begin{figure}[p]
\centering
\includegraphics[width=0.5\textwidth]{Fig_map.pdf}
\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}. Approximate sample locations are marked by colored circles. Numbering corresponds to the sample location in Table \ref{results}.}
\label{map}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\textwidth]{Fig_column-01.jpg}
\caption{On the left, a stratigraphic column showing the generalized lithstratigraphic units that were sampled for this study. On the right are the carbon and oxygen isotopic compositions of apparently well-preserved brachiopods, collected from the same section in Gotland. Isotopic compositions of brachiopods are modified from \cite{Bickert1997}.}
\label{column}
\end{figure}

\begin{sidewaysfigure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_section-01.jpg}
\caption{The \deltac C, \deltao O and $\Delta_{47}$ temperature of all the samples analyzed in this study, plotted against stratigraphic height. A height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds. Also included are data from Munnecke et al. (2003).}
\label{section}
\end{sidewaysfigure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_Brand-01.jpg}
\caption{Bulk manganese and strontium concentrations of the brachiopods, micrites and cements in this study. The brachiopods in this study have more strontium and less manganese than most of the brachiopods analyzed by Brand et al. (2012), which is also shown in this figure. Lines represent estimates by different authors of the range of pristine concentrations.}
\label{Brand}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_metals-01.jpg}
\caption{Scatter plots showing the relationship between temperature and bulk trace metal concentration. The only statistically robust correlation ($\text{p}<0.05$ compared to zero correlation) is that between temperature and Sr concentration of rugose corals.}
\label{metals}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{Fig_EBSD_coral-01.jpg}
\caption{Four different images of the same area of a rugose coral thin section, all at the same magnification. A) Reflected light. The line of dark pits are damaged caused by the E-probe electron beam. The dark triangle labelled ``RC'' was identified as a recrystallized domain of the coal septum. B) EBSD analysis. The recrystallized domain is clearly marked by many calcite crystals with the same orientation. The black scale bar represents 500 $\mu$m. C) Iron and D) manganese concentrations measured using the E-probe. The maps are qualitative, with warmer colors representing higher concentrations. The overlaid spots were measured quantitatively, and are labeled in units of ppm. For both metals, the recrystallized domain is indistinguishable from the rest of the septum.}
\label{EBSD_coral}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_EBSD_brach-01.jpg}
\caption{A) Transmitted polarized light image of a thin section of a brachiopod with $\Delta_{47}$ temperature 34\degrees C. The color overlay is an electron backscatter diffraction image of the same sample. The orientation of calcite crystals is indicated by color. B) Euler pole figure generated from the EBSD analysis.}
\label{EBSD_brach}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Fig_flakes-01.jpg}
\caption{Scanning electron microscope images of flakes of brachiopod shell. Images were taken in secondary electron mode. Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis. Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis. Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 34 $\pm$ 2\degrees C; B) 35 $\pm$ 3\degrees C; C) 35 $\pm$ 3\degrees C.}
\label{flakes}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{Fig_comparisons-01.jpg}
\caption{Comparisons between the \deltac C, \deltao O and $\Delta_{47}$ temperature of brachiopod samples and adjacent micrite.}
\label{comparisons}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.7\textwidth]{Fig_ellipses-01.jpg}
\caption{The $\Delta_{47}$ temperature and calculated water \deltao O for all samples analyzed in this study. Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
\label{ellipses}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{Fig_veizer-01.jpg}
\caption{Caption}
\label{veizer}
\end{figure}

\end{document}

### 2013-05-15 draft.tex
switched the order of the methods sections
added paragraph explaining cant do thin sections for all fossils

@@
 Flakes were then sonicated briefly, rinsed, dried in a 50\degrees C oven, and powdered using a mortar and pestle.
-Only the apical ends of rugose corals were sampled, after removal of any micrite from the surface using an abrasive bit.
+Only the apical ends of rugose corals were sampled, after removal of any sediment from the surface using an abrasive bit.
 The apical end was then cut off using a diamond microsaw, and powdered using a mortar and pestle.
@@
 However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite.
-To assess diagenetic alteration, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
+Finally, to assess diagenetic alteration, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
+
+While all fossils were inspected under a microscope prior to sampling, we were not able to inspect all fossils in thin section, because carbonate clumped isotope measurements require large amounts of material, some of which is lost during preparation of thin sections.
+However, two representative brachiopods and rugose corals were examined in thin section.
+30 $\mu$m-thick uncovered, polished thin sections were obtained from cuts through the fossils.
+These thin sections were examined first in transmitted and reflected light to characterize calcite textures, and were then characterized using several different spectroscopic techniques, described below.
+
+\subsection{Electron microscopy}
+
+For electron backscatter diffraction (EBSD),  The thin sections were analyzed using the same SEM as above, under variable pressure vacuum, with an electron high tension of 20 kV.
+The stage was tilted 70$^{\circ}$.
+The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame.
+The step size was 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
+The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite.
+The diffracted beam then interacts with the EBSD detector, depdending on the orientation of the calcite crystal.
+The data were analyzed using software from Oxford instruments.
+
+Rinsed and dried brachiopod flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
+The coated flakes were imaged using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under high vacuum using the secondary electron detector.
+The working distance was $\sim 6$ mm and the electron high tension was 10 kV.
+
+\subsection{Wavelength-dispersive spectroscopy}
+
+The same thin sections as above were first coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer.
+Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite.
+Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
+Quantitative spot measurements were made using the JEOL quantitiative analysis program.
+
+\subsection{Bulk trace metal analysis}
+
+Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
+Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects.
+Solutions were sampled using a Cetac ASX 260 autosampler, aspirated into the Ar plasma using a peristaltic pump, and then analyzed on a Thermo Scientific iCAP 6000 series inductively-coupled plasma optical emission spectrometer (ICP-OES).
+Between each set of ten unknown samples, three standard solutions (10 ppb, 100 ppb, and 1 ppm) were measured.
+Reproducibility was on average 3\% for Fe, 2\% for Mn and 3\% for Sr.
 
@@
 \noindent where T is temperature in degrees Kelvin.
-
-\subsection{Electron microscopy}
-
-Rinsed and dried brachiopod flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
-The coated flakes were imaged using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under high vacuum using the secondary electron detector.
-The working distance was $\sim 6$ mm and the electron high tension was 10 kV.
-
-For electron backscatter diffraction (EBSD), 30 nm-thick uncovered, polished thin sections were obtained from cuts through the fossils.
-The thin sections were analyzed using the same SEM as above, under variable pressure vacuum, with an electron high tension of 20 kV.
-The stage was tilted 70$^{\circ}$.
-The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame.
-The step size was 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
-The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite.
-The diffracted beam then interacts with the EBSD detector, depdending on the orientation of the calcite crystal.
-The data were analyzed using software from Oxford instruments.
-
-\subsection{Wavelength-dispersive spectroscopy}
-
-The same thin sections as above were first coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer.
-Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite.
-Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
-Quantitative spot measurements were made using the JEOL quantitiative analysis program.
-
-\subsection{Bulk trace metal analysis}
-
-Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
-Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects.
-Solutions were sampled using a Cetac ASX 260 autosampler, aspirated into the Ar plasma using a peristaltic pump, and then analyzed on a Thermo Scientific iCAP 6000 series inductively-coupled plasma optical emission spectrometer (ICP-OES).
-Between each set of ten unknown samples, three standard solutions (10 ppb, 100 ppb, and 1 ppm) were measured.
-Reproducibility was on average 3\% for Fe, 2\% for Mn and 3\% for Sr.
 
@@
 \bibliographystyle{garynat}
-\bibliography{draft_05_08_13}
+\bibliography{draft}
 

### 2013-05-16 draft.tex
edited thin section paragraph, electron microscopy section

@@
 While all fossils were inspected under a microscope prior to sampling, we were not able to inspect all fossils in thin section, because carbonate clumped isotope measurements require large amounts of material, some of which is lost during preparation of thin sections.
-However, two representative brachiopods and rugose corals were examined in thin section.
-30 $\mu$m-thick uncovered, polished thin sections were obtained from cuts through the fossils.
-These thin sections were examined first in transmitted and reflected light to characterize calcite textures, and were then characterized using several different spectroscopic techniques, described below.
+However, 30 $\mu$m-thick uncovered, polished thin sections were obtained from cuts through four representative fossils, two brachiopods and two rugose corals.
+These thin sections were inspected first in transmitted and reflected light to characterize calcite textures, and were then examined using several different spectroscopic techniques, described below.
 
@@
 
-For electron backscatter diffraction (EBSD),  The thin sections were analyzed using the same SEM as above, under variable pressure vacuum, with an electron high tension of 20 kV.
+Electron backscatter diffraction (EBSD) provides an independent measure of recrystallization by mapping the orientation of calcite crystals in a thin section.
+This allows more detailed characterization of calcite fabrics than is possible using optical microscopy, and in the case of brachiopods, comparison to pristine modern examples \citep{PerezHuerta2007}.
+The thin sections were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under variable pressure vacuum, with an electron high tension of 20 kV.
 The stage was tilted 70$^{\circ}$.
-The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame.
-The step size was 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
+The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame and a step size of 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
 The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite.
@@
 
-Rinsed and dried brachiopod flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
-The coated flakes were imaged using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under high vacuum using the secondary electron detector.
+Flakes of the secondary fibrous layer from over ten different brachiopods were also imaged to further inspect the preservation of original fibrous calcite texture.
+The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
+The coated flakes were imaged using using the secondary electron detector in the same SEM as above, under high vacuum.
 The working distance was $\sim 6$ mm and the electron high tension was 10 kV.

### 2013-05-16 draft.tex
edited Eprobe and bulk trace metal sections to explain why we did them

@@
 This allows more detailed characterization of calcite fabrics than is possible using optical microscopy, and in the case of brachiopods, comparison to pristine modern examples \citep{PerezHuerta2007}.
-The thin sections were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under variable pressure vacuum, with an electron high tension of 20 kV.
+Two thin sections, one rugose coral and one brachiopod, were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under variable pressure vacuum, with an electron high tension of 20 kV.
 The stage was tilted 70$^{\circ}$.
@@
 
-The same thin sections as above were first coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer.
+The use of wavelength-dispersive spectroscopy in combination with optical microscopy and EBSD allows characterization of the trace metal contents of specific calcite textures.
+By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set.
+The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer.
 Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite.
@@
 
-Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
+Although we attempted to sample only well-preserved skeletal material for the carbonate clumped isotope measurements, the large mass of material required makes it very difficult to avoid including small amounts of other phases.
+The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998,Brand1980,Brand2012,Came2007,Finnegan2011,Shields2003}.
+To measure bulk trace metal concentrations, between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
 Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects.

### 2013-05-17 draft.tex
fixed the clumped isotope analysis section of the methods, still may need to be changed to reflect calculation of ETF

@@
 Approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90\degrees C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate.
-The extraction and subsequent cleaning of the CO$_2$ followed procedures outlined by \cite{Ghosh2006}.
+The subsequent cleaning of the CO$_2$ followed procedures outlined by \cite{Ghosh2006}.
 The cleaned CO$_2$ was analyzed on a Finnigan MAT253 gas source mass spectrometer configured to collect masses 44 through 49.
 The sample gas was measured in eight separate acquisitions, each of which was bracketed by a measurement of reference CO$_2$ with known \deltac C and \deltao O.
-The \deltac C, \deltao O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample using the reference CO$_2$ as the reference material.
+The \deltac C, \deltao O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample relative to the reference CO$_2$.
 Acquisition-to-acquisition standard deviations were on average 0.03$\permil$ for $\delta_{47}$, 0.04$\permil$ for \deltao O, and 0.02$\permil$ for \deltac C.
-Preliminary values for $\Delta_{47}$ and $\Delta_{48}$ of each sample were also calculated using the reference CO$_2$ as the reference material and the following equations:
+Raw measured values for $\Delta_{47}$ and $\Delta_{48}$ of each sample, defined as $\Delta_{47-[\text{SGvsWG}]}$ and $\Delta_{48-[\text{SGvsWG}]}$, were also calculated relative to the reference CO$_2$ following equations 3 and 4 in \cite{Huntington2009}.
 
-\begin{align}
-\Delta_{47\text{ prelim}}&=\left[\left(\frac{\text{R}^{47}_{\text{sa}}}{\text{R}^{47*}_{\text{sa}}}\frac{\text{R}^{47*}_{\text{ref}}}{\text{R}^{47}_{\text{ref}}}-1\right)-\left(\frac{\text{R}^{46}_{\text{sa}}}{\text{R}^{46*}_{\text{sa}}}\frac{\text{R}^{46*}_{\text{ref}}}{\text{R}^{46}_{\text{ref}}}-1\right)-\left(\frac{\text{R}^{45}_{\text{sa}}}{\text{R}^{45*}_{\text{sa}}}\frac{\text{R}^{45*}_{\text{ref}}}{\text{R}^{45}_{\text{ref}}}-1\right)\right]\times 1000\\
-\Delta_{48\text{ prelim}}&=\left[\left(\frac{\text{R}^{48}_{\text{sa}}}{\text{R}^{48*}_{\text{sa}}}\frac{\text{R}^{48*}_{\text{ref}}}{\text{R}^{48}_{\text{ref}}}-1\right)-2\left(\frac{\text{R}^{46}_{\text{sa}}}{\text{R}^{46*}_{\text{sa}}}\frac{\text{R}^{46*}_{\text{ref}}}{\text{R}^{46}_{\text{ref}}}-1\right)\right]\times 1000
-\end{align}
+All $\Delta_{47}$ values are projected into the absolute reference frame \citep{Dennis2011}.
+When possible, raw $\Delta_{47-[\text{SGvsWG}]}$ values are converted to the absolute reference frame using an empirical transfer function (ETF) constructed from the heated and equilibrated  gases measured during the same week as the samples.
+Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000\degrees C for three hours, and then cooled rapidly to room temperature.
+Equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25\degrees C for at least 24 hours.
+Heated gases were made from two separate reservoirs of CO$_2$ with different \deltao O compositions, and equilibrated gases were equilibrated with either of two separate waters, also differing in their \deltao O compositions.
+Heated and equilibrated gases were cleaned following the same procedure as the samples.
+The ETF was constructed and then used to convert raw $\Delta_{47-[\text{SGvsWG}]}$ values following \cite{Dennis2011}.
+Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90\degrees C by adding 0.092$\permil$ \citep{Henkes2013}.
 
-\noindent where R* refers to the stochastic distribution of masses calculated from the bulk isotopic composition of the sample (sa) or reference gas (ref).
+Some samples were measured during weeks when no equilibrated gases were measured, so an ETF could not be constructed for those weeks.
+These samples were projected into the absolute reference frame using a secondary transfer function (STF).
+Accepted values of two California Institute of Technology intralab standards were determined in the absolute reference frame from measurements of these standards during weeks when an ETF could be constructed from heated and equilibrated gas measurements.
+To create the STF, the accepted values for the two standards, as well as the accepted value of 0.0266$\permil$ for heated gases, were plotted against their $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
+$\Delta_{47}$ values of the samples in the \cite{Ghosh2006} reference frame were corrected using this STF following equation C1 in \cite{Dennis2011}.
+A separate STF was constructed for each week of measurements during which no equilibrated gases were measured.
 
-Measurements were standardized by comparison to heated gases, which are CO$_2$ with two different bulk isotopic compositions that was heated at 1000\degrees C for three hours, and then cooled rapidly to room temperature.
-At least one heated gas was measured each day, and each sample was compared only to the heated gases that were run during the seven days before and after the date on which the sample was measured.
-The $\delta_{48}$ and $\Delta_{48\text{ prelim}}$ of each sample was compared to a line of best fit through a plot of $\delta_{48}$ versus $\Delta_{48}$ of the heated gases.
-All samples whose $\Delta_{48}$ fell more than 1$\permil$ off the line were discarded due to likely contamination.
-The $\Delta_{47}$ values of the remaining samples were corrected so that the $\Delta_{47}$ reflects a comparison to the stochastic heated gases, rather than the reference gas, using the following equation:
+Although all samples were cleaned prior to introduction into the mass spectrometer, the possibility for contamination from hydrocarbons and chlorocarbons still exists.
+To screen for contamination, the $\delta_{48}$ and $\Delta_{48}$ of each sample were compared to a line of best fit through a plot of $\delta_{48}$ versus $\Delta_{48}$ of the heated gases, all in the \cite{Ghosh2006} reference frame.
+Samples whose $\Delta_{48}$ fell more than 1$\permil$ off the line were discarded due to likely contamination.
 
-\begin{align}
-\Delta_{47}=\Delta_{47\text{ prelim}}-\left(\delta^{47}\cdot\text{M}_{\text{HG}}+\text{b}_{\text{HG}}\right)\cdot\text{K}_{\text{stretch}}+\text{K}_{\text{acid}}
-\end{align}
-
-\noindent where M$_{\text{HG}}$ is the slope of the best fit line through a plot of $\delta_{47}$ versus $\Delta_{47}$ of heated gases, b$_{\text{HG}}$ is the intercept of the same line, K$_{\text{stretch}}=-0.8453$ and K$_{\text{acid}}=0.08$.
-Finally, the $\Delta_{47}$ of each sample was corrected based on the average offset from the accepted value of the $\Delta_{47}$ of two lab standards measured during the same week as the sample.
-
-$\Delta_{47}$ values were converted to water temperature using the following equation \citep{Ghosh2006}:
-
-\begin{align}
-\Delta_{47}=50.0592\left(10^6\text{T}^{-2}\right)-0.02
-\end{align}
-
-\noindent where T is temperature in degrees Kelvin.
-
-The \deltao O of water was calcuated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite \citep{Kim1997}:
-
-\begin{align}
-1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42
-\end{align}
-
-\noindent where T is temperature in degrees Kelvin.
+$\Delta_{47-\text{RF-AC}}$ values were converted to temperature using the \cite{Ghosh2006} calibration line, which is projected into the absolute reference frame in equation 9 in \cite{Dennis2011}.
+The \deltao O of water was calcuated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.
 

### 2013-05-18 draft.tex
fixed the stratigraphic trends section of the discussion - says more about d18O, less about diagenesis

@@
 In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease.
-In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the water is diagnostically low.
+In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low.
 The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}.
@@
 Heated and equilibrated gases were cleaned following the same procedure as the samples.
-The ETF was constructed and then used to convert raw $\Delta_{47-[\text{SGvsWG}]}$ values following \cite{Dennis2011}.
+The ETF was constructed and then used to convert raw $\Delta_{47-[\text{SGvsWG}]}$ values to the absolute reference frame following \cite{Dennis2011}.
 Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90\degrees C by adding 0.092$\permil$ \citep{Henkes2013}.
@@
 We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters.
-The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C values of approximately -0.5$\permil$.
+The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$.
 The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
 
-The \deltao O values of fossil material is consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
-We were not able to resolve a \deltao O excursion from our sample set, but the range in \deltao O values of fossil material is comparable to the range observed by \cite{Bickert1997} and \cite{Munnecke2003}.
+The \deltao O values of fossil material are consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
+We are not able to resolve an excursion in \deltao O, although this may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}.
+The average \deltao O$_{\text{PDB}}$ of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples consistent with those of \cite{Munnecke2003} but our Upper Visby samples more depleted in $^{18}$O.
+Higher in the section, the \deltao O of the fossils in this study appear consistent with values reported by \cite{Bickert1997}.
 
 The clumped isotope temperature of samples is highly variable within beds, often covering a range of 10\degrees C within a single bed (Figure \ref{section}).
-The lowest temperature sample, a brachiopod, and the highest temperature sample, a rugose coral, are both from the Slite Formation, deposited well after the hypothesized climate phenomenon associated with the isotope excursions in the Lower and Upper Visby Beds.
-Due to the scatter in temperature, we were not able to resolve a change in temperature with time.
+The lowest temperature sample, a brachiopod, and the highest temperature sample, a rugose coral, are both from the Slite Formation, deposited well after the hypothesized climate phenomenon associated with the isotopic excursions in the Lower and Upper Visby Beds.
+Due to the scatter in temperature, we are not able to resolve a change in temperature with time.
 
@@
 Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C.
-Diagenetic alteration can occur at elevated temperatures deep underground, which can increase the clumped isotope temperature of altered samples, but alteration can also occur near the Earth's surface, in the presence of cool, likely meteoric water.
-Consequently, low-temperature samples cannot be assumed to be more pristine than high-temperature samples.
-However, alteration in the presence of cool, meteoric water will deplete the \deltao O of cabonate, because meteoric water usually has a very light \deltao O.
-Most of our fossil samples are very similar in \deltao O, making alteration by meteoric water unlikely.
+All of the fossil samples have fairly high values of \deltao O, making alteration by meteoric water unlikely.
 This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature.

### 2013-05-18 draft.tex
fixed independent assessments section to be more subtle, more clear, 5% estimate

@@
 By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set.
-The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer.
+The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer (E-probe).
 Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite.
@@
 Although we attempted to sample only well-preserved skeletal material for the carbonate clumped isotope measurements, the large mass of material required makes it very difficult to avoid including small amounts of other phases.
-The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998,Brand1980,Brand2012,Came2007,Finnegan2011,Shields2003}.
-To measure bulk trace metal concentrations, between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
+The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998,Brand2012,Came2007,Finnegan2011}.
+Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980,Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples.
+Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
 Standards were prepared from standard solutions of Sr, Mn and Fe with the same concentration of Ca and HNO$_3$ as the sample solutions to minimize matrix effects.
@@
 We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters.
-The sample set does not capture the relaxation of this excursion, but our samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$.
+The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$.
 The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
@@
 
-The large scatter in temperatures also suggests variable preservation quality among the samples.
+The large scatter in temperatures suggests variable preservation quality among the samples.
 Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C.
@@
 
-The concentrations of trace metals in fossil calcite have been widely used as indicators of recrystallization and contamination from cements and sediments.
-Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980,Shields2003}, so we measured the concentrations of these three elements in a large subset of our samples.
-The Mn and Sr concentrations of almost all of the brachiopods analyzed in this study fall within the most strictly defined limits for the Mn and Sr concentrations of pristine brachiopods (Figure \ref{Brand}).
-The rugose corals tend to have more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while the cements and micrites tend to have the most highly altered trace metal signatures.
+Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration, suggest that all the brachiopods in this study are pristine.
+The bulk Mn and Sr concentrations of all brachiopods fall within previously defined limits for the Mn and Sr concentrations of pristine brachiopods (Figure \ref{Brand}).
+However, micro-scale studies of calcite fabrics, as well as clumped isotope measurements, demonstrate that there is a range in preservation quality among the brachiopods that cannot be identified on the basis of trace metal measurements alone.
+We first explore possible reasons for the insensitivity of trace metal concentrations to preservation quality.
+Then we consider EBSD and SEM evidence for the inclusion of diagenetic material in the typical fossil sample.
+Finally, we explore the use of clumped isotopes to distinguish the best-preserved samples from among those that would ordinarily all be identified as pristine based on their trace metal contents.
 
-The implications of the brachiopod trace metal compositions are inconsistent with the clumped isotope results.
-The brachiopods in this study have a range of temperatures of more than 15\degrees C, indicating variable preservation, but they would all be judged pristine or nearly pristine based on their trace metal compositions.
-We hesitate to extend this interpretation to rugose corals, because their trace metal composition is less well-studied than that of brachiopods.
-Regardless, the results in Figure \ref{Brand} imply that the carbonate clumped isotope thermometer is a more sensitive indicator of diagenetic alteration than are bulk trace metal compositions.
-
-We further tested the sensitivity of trace metal composition to diagenetic alteration by investigating possible relationships between trace metal composition and temperature (Figure \ref{metals}).
-We tested brachiopods and rugose corals separately and combined with micrites and cements for the robustness of a linear correlation between trace element composition and temperature.
-We also tested for a linear correlation between the log of trace element composition and temperature.
+There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration.
+Our sample set may not include a sufficient number of highly altered samples to capture the full range in severity of alteration.
+This hypothesis is supported by a conspicuous lack of statistically robust correlations between trace metal composition and temperature (Figure \ref{metals}).
+We tested brachiopods and rugose corals separately and combined with micrites and cements for the robustness of a linear correlation between trace metal concentrations and temperature, as well as the log of trace metal concentrations and temperature.
 We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero.
 Only one pair of variables, the Sr concentration and temperature of rugose corals, has a p value of less than 0.05, meaning we can be 95\% confident that the correlation is real.
+While these tests suggest that the scatter in trace metal concentrations is a large cause of their insensitivity to diagenetic alteration, there is also the possibility that early void-filling cements, which would have cool temperatures but altered trace metal concentrations, interfere with the signal.
 
-There are several possible explanations for the scarcity of strong correlations between trace metal concentrations and temperature.
-Our sample set may not include a sufficient number of highly altered samples to capture the full range in severity of alteration.
-Indeed, the brachiopods in this study have a much smaller range in trace metal concentrations than the brachiopods analyzed by \cite{Brand2012} (Figure \ref{Brand}).
-The variable temperatures of cements, depending on whether the cements formed early during shallow burial or later in the presence of high-temperature fluids, may also contribute to the lack of correlation between trace metal concentrations and temperatures.
-
-An additional explanation for the lack of strong correlations between temperature and trace metal concentrations is provided by electron microprobe (E-probe analysis) of the trace metal signatures of recrystallization.
+E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization may not actually have any effect on Fe and Mn concentrations.
 A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B).
 However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D).
-This is only one example, but it suggests that variations in trace metal concentrations may be dominated by contamination from cements and sediments, rather than recrystallization.
+This is only one example, but it suggests that recrystallization may not affect trace metal concentrations, and instead, variations in trace metal concentrations may be dominated by contamination from cements and sediments.
 If recrystallization is an important contributor to the elevated temperatures of fossil material, trace metals may not be expected to be reliable indicators of diagenetic alteration.
@@
 There are large interlocking cement crystals in the upper left hand corner of the EBSD image, but these formed in the void space between the two valves, and would not have been included in the sample for clumped isotope analysis.
+Based on visual inspection of the EBSD image of this brachiopod's secondary fibrous layer, we suggest that this layer contains a maximum of 5\% recrystallized material.
 
@@
 The micro-scale studies of calcite textures in Figures \ref{EBSD_coral}--\ref{flakes} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite.
-Variable fractions of late-phase calcite likely contributed to the range in temperature among fossil samples.
-We only performed SEM and EBSD analysis on a small subset of samples, so we will not attempt to use this evidence to compare the relative preservation qualtiy of low and high-temperature samples.
-An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite.
+Variable fractions of late-phase calcite likely contribute to the range in temperature among fossil samples.
+We only performed SEM and EBSD analysis on a small subset of samples, so we will not attempt to use this evidence to compare the relative preservation quality of low and high-temperature samples.
+An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite, likely less than 5\% of the sample.
 We therefore look to these samples as the most reliable records of ocean temperature and \deltao O during the Silurian.
@@
 \includegraphics[width=\textwidth]{Fig_Brand-01.jpg}
-\caption{Bulk manganese and strontium concentrations of the brachiopods, micrites and cements in this study.
-The brachiopods in this study have more strontium and less manganese than most of the brachiopods analyzed by Brand et al.
-(2012), which is also shown in this figure.
-Lines represent estimates by different authors of the range of pristine concentrations.}
+\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
+The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by Brand et al.
+(2012), which are also shown in this figure.
+Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods.
+Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.}
 \label{Brand}

### 2013-05-19 draft.tex
fixed 1st 5 paragraphs of constraining temp section: added solid-state, 35 degrees brachs, 5% estimates

@@
 
-Independent assessments of preservation quality confirm that the lowest-temperature brachiopods are very well preserved.
-To provide additional support for this hypothesis, we measured the clumped isotope composition of the micrite adjecent to four brachiopods (Figure \ref{comparisons}).
-In all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod.
-We expect micrite to be less pristine than our brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis.
-Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with our hypothesis that the lowest temperature samples are the most pristine.
+% comparisons support hypothesis that low temp most pristine
+% array shows rock-buffered diagenesis
+% most pristine sample only one replicate
+% only upper estimates because all samples could have a little contamination
+% Silurian warm and similar d18O to today
+% resolve veizer controversy
+% surprising that d18O has not changed
 
-The analyses described above have allowed us to define the temperature consequences of diagenesis, but to confidently interpret the \deltao O of Silurian seawater, we also need to characterize the oxygen isotopic consequences of diagenesis.
-To constrain the effect of diagenetic alteration on the \deltao O of water calculated from the samples, we plotted the temperature versus the \deltao O of water for all of the samples in this study (Figure \ref{ellipses}).
-There is a clear positive relationship between the temperature and \deltao O of water, indicating that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased.
-This relationship is diagnostic of diagenesis occurring in the presence of a low water-to-rock ratio.
-In this diagenetic scheme, also called rock-buffered diagenesis, as the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water.
-Importantly, rock-buffered diagenesis causes the temperature of more altered samples to increase, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.
+% array. probably not solid-state reordering, so we interpret rock-buffered diagenesis
+% comparisons support hypothesis that low temp most pristine, EBSD supports, metals dont negate
+% most pristine samples population at 34, one at 30, resample not replicate, no big deal!
+% only upper estimates, but probably pretty close
+% Silurian warm and similar in d18O to today. matches previous clumped studies
+% resolves veizer controversy
+% implications of warm oceans for Ireviken climate rule out big changes, can't resolve small
+% implications of constant d18O?? relative amounts constant. either both constant or coupled.
 
-The lowest, most-pristine sample in this study (Figure \ref{ellipses}) indicates that tropical Silurian oceans had a temperature of $30\pm 2$\degrees C and a \deltao O of $-1.7\pm 0.5\permil$.
-One caveat of these values is that they are based on only one measurement (G13 10.5 to 10.8 B1).
-The first time this brachiopod was picked, it produced enough material for two analyses, but one analysis was rejected because its $\Delta_{48}$ was more than 1$\permil$ above the heated gas standard line.
-Later, more material was picked from the same bachiopod, but in order to get enough material for a replicate measurement, very small flakes had to be included.
-We worried that the small flakes would be lost in the rinsing procedure, so this replicate sample was not rinsed.
-The temperature of the replicate sample (G13 10.5 to 10.8 B1 leftovers) was $41\pm 4$ degrees C, which may reflect contamination that ordinarily would have been removed during the rinse step of sample preparation.
-Due to the non-standard treatment of this replicate sample, we do not treat it as a true replicate, and we report its temperature separately.
+To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore record the most reliable record of Silurian ocean conditions.
+This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected.
+As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the \deltao O of interstitial water.
+When the clumped isotope temperatures of the samples in this study are plotted versus the \deltao O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}).
+This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is diagnostic of rock-buffered diagenesis.
+Importantly, during rock-buffered diagenesis, more altered samples have higher temperatures, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.
+
+Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis, but we rule it out as a major contributor to our results.
+During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite.
+This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and reprecipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
+However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the variation in temperature among the samples.
+Additionally, solid-state diffusion is thought to occur very slowly at temperatures below 100\degrees C (D.
+Stolper personal communication 2013), which is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983,Wenzel2000}.
+Solid-state diffusion is not consistent with our observations, leaving rock-buffered diagenesis as the most likely characterization of the diagenetic regime.
+
+Comparisons of the clumped isotope temperature of different calcite textures provide additional support for the hypothesis that the samples underwent rock-buffered diagenesis.
+We measured the clumped isotope composition of the micrite adjecent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod (Figure \ref{comparisons}).
+Micrite is expected to be less pristine than the brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis.
+Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
+
+The lowest-temperature, most pristine sample in this study, a brachiopod, indicates that tropical Silurian oceans had a temperature of $30\pm 2$\degrees C and a \deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$ (Figure \ref{ellipses}), where the error estimates represent the standard error of the mean over eight acquisitions.
+The two next-most pristine samples, also brachiopods, have temperatures of $33\pm 2$\degrees C and $35\pm 2$\degrees C, which correspond to seawater \deltao O$_{\text{SMOW}}$ values of $-1.5\pm 0.3\permil$ and $-1.9\pm 0.4\permil$, respectively.
+One caveat of these values is that they are each based on only one measurement due to sample-size limitations.
+The lowest-temperature brachiopod was picked a second time, but to collect enough material for analysis, some less pristine flakes had to be included, which may explain the $41\pm 4$\degrees C temperature of this re-sample (G13 10.5 to 10.8 B1 leftovers).
+However, we argue that the differences between the temperatures and \deltao O values implied by the three most pristine samples are small enough that they essentially represent the same warm, tropical Silurian ocean water.
 
 A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates.
-We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples, and we are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine sample.
-However, the micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, suggesting that the upper estimate is likely very similar to the true value.
+We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples.
+We are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine samples.
+However, micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, comprising a maximum of 5\% of the material.
+Assuming linear mixing, which is not technically correct but close enough for our purposes, if 5\% of the sample material was recrystallized at 100\degrees C and the rest of the sample material retained an original temperature of 30\degrees C, the measured temperature of the sample would be $\sim34$\degrees C.
+Therefore, the constraints provided by the EBSD mapping of the bachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.
 
@@
 Our results, along with those of earlier clumped isotope studies \citep{Came2007,Finnegan2011,Dennis2013}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates.
-Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O of early Paleozoic oceans \citep{Grossman2012}.
+Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ \citep{Jaffres2007,Veizer1999} for the \deltao O of early Paleozoic oceans \citep{Grossman2012}.
 Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{Fig_comparisons-01.jpg}
-\caption{Comparisons between the \deltac C, \deltao O and $\Delta_{47}$ temperature of brachiopod samples and adjacent micrite.}
-\label{comparisons}
+\includegraphics[width=0.7\textwidth]{Fig_ellipses-01.jpg}
+\caption{The $\Delta_{47}$ temperature and calculated water \deltao O for all samples analyzed in this study.
+Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
+\label{ellipses}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.7\textwidth]{Fig_ellipses-01.jpg}
-\caption{The $\Delta_{47}$ temperature and calculated water \deltao O for all samples analyzed in this study.
-Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
-\label{ellipses}
+\includegraphics[width=0.5\textwidth]{Fig_comparisons-01.jpg}
+\caption{Comparisons between the \deltac C, \deltao O and $\Delta_{47}$ temperature of brachiopod samples and adjacent micrite.}
+\label{comparisons}
 \end{figure}

### 2013-05-19 draft.tex
fixed the rest of the constraining temps section, havent read thru yet

@@
 
-% comparisons support hypothesis that low temp most pristine
-% array shows rock-buffered diagenesis
-% most pristine sample only one replicate
-% only upper estimates because all samples could have a little contamination
-% Silurian warm and similar d18O to today
-% resolve veizer controversy
-% surprising that d18O has not changed
-
 % array. probably not solid-state reordering, so we interpret rock-buffered diagenesis
@@
 
-To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore record the most reliable record of Silurian ocean conditions.
+To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore preserve the most reliable record of Silurian ocean conditions.
 This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected.
@@
 This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and reprecipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
-However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the variation in temperature among the samples.
+However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the observed variability in temperature among the samples.
 Additionally, solid-state diffusion is thought to occur very slowly at temperatures below 100\degrees C (D.
@@
 
-Comparisons of the clumped isotope temperature of different calcite textures provide additional support for the hypothesis that the samples underwent rock-buffered diagenesis.
+The hypothesis that the samples underwent rock-buffered diagenesis is further supported by comparisons of the clumped isotope temperature of different calcite textures.
 We measured the clumped isotope composition of the micrite adjecent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod (Figure \ref{comparisons}).
@@
 
-Our results, along with those of earlier clumped isotope studies \citep{Came2007,Finnegan2011,Dennis2013}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates.
-Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ \citep{Jaffres2007,Veizer1999} for the \deltao O of early Paleozoic oceans \citep{Grossman2012}.
+The results of this study are also consistent with those of earlier clumped isotope studies.
+\cite{Came2007} and \cite{Finnegan2011} both measured the clumped isotope compositions of brachiopods from the Early Silurian on Anticosti Island, Canada.
+These two studies report temperatures of $30-35$\degrees C and seawater \deltao O$_{\text{SMOW}}$ values of around $-1\permil$, which are similar to the results of this study.
+Importantly, Anticosti Island recorded deposition on the eastern margin of the Panthalassic Ocean during the Silurian, while Gotland recorded deposition in the Rheic Ocean, which was separated from the Panthalassic Ocean by a mountain range.
+The strong similarity in temperatures and \deltao O between the two studies indicates that both studies likely reflect water that is representative of the global ocean, rather than an isolated inland sea.
+
+Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates.
+Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O of early Paleozoic oceans \citep{Grossman2012}.
 Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans.
@@
 
-Given the large potential for change in the \deltao O of seawater through exchange of oxygen isotopes with crustal material, it is surprising that four separate clumped isotope studies, each from a different time in the Phanerozoic, have all implied similar \deltao O values of seawater.
+Due in large part to diagenetic alteration, our results are unable to resolve a change in temperature or the \deltao O of the oceans through the Ireviken event.
+The consistent scatter in fossil temperatures throughout the section likely rules out any large changes in temperature or \deltao O of the oceans, but this scatter prevents resolution of any smaller changes.
+Temperatures of 30\degrees C in tropical oceans are similar to those of modern tropical oceans, and therefore are not incompatible with the presence of polar ice sheets, as suggested by Early Silurian age tillites in Brazil \citep{Grahn1992}.
+
+The results from this study join those of previous clumped isotope studies in implying that the \deltao O of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$.
 Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}.
-The controls on the magnitudes of seafloor and continental weathering do not form a clear negative feedback loop, so we would not expect the \deltao O of seawater to be broadly constant through time, contrary to the evidence of several clumped isotope studies.
+If this buffer cannot explain the relative constancy of seawater \deltao O values through time, then the two competing controls on seawater \deltao O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years.
+This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.
 

### 2013-05-19 draft.tex
ready to send to Dave and Seth

@@
 Modern measurements of sea surface temperature in the Solomon Sea, east of Papua New Guinea, are 30\degrees C \citep{Reynolds1994}, matching our estimate of the temperature at a similar latitude during the Silurian.
-The average \deltao O of the modern ocean is approximately $-0.3\permil$ \citep{Shackleton1974}, but without ice, the average \deltao O would decrease to $-1.4\permil$ \citep{Lhomme2005}.
+The average \deltao O$_{\text{SMOW}}$ of the modern ocean is approximately $-0.3\permil$ \citep{Shackleton1974}, but without ice, the average \deltao O$_{\text{SMOW}}$ would decrease to $-1.4\permil$ \citep{Lhomme2005}.
 The volume of continental glaciers is poorly constrained for the Early Silurian, but regardless, the \deltao O of Silurian oceans was likely within a few permil of the modern value.
@@
 \cite{Came2007} and \cite{Finnegan2011} both measured the clumped isotope compositions of brachiopods from the Early Silurian on Anticosti Island, Canada.
-These two studies report temperatures of $30-35$\degrees C and seawater \deltao O$_{\text{SMOW}}$ values of around $-1\permil$, which are similar to the results of this study.
+These two studies report temperatures of 30 to 35\degrees C and seawater \deltao O$_{\text{SMOW}}$ values of around $-1\permil$, which are similar to the results of this study.
 Importantly, Anticosti Island recorded deposition on the eastern margin of the Panthalassic Ocean during the Silurian, while Gotland recorded deposition in the Rheic Ocean, which was separated from the Panthalassic Ocean by a mountain range.
@@
 Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates.
-Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O of early Paleozoic oceans \citep{Grossman2012}.
+Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans \citep{Grossman2012}.
 Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans.
 The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature.
-Clumped isotopes additionally demonstrate that extremely hot ($\sim 70$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates.
+Clumped isotopes additionally demonstrate that extremely hot ($>50$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates.
 Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality.
@@
 
-The results from this study join those of previous clumped isotope studies in implying that the \deltao O of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$.
+The results from this study join those of previous clumped isotope studies in implying that the \deltao O$_{\text{SMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$.
 Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{Fig_veizer-01.jpg}
-\caption{Caption}
+\includegraphics[width=0.6\textwidth]{Fig_veizer-01.jpg}
+\caption{In gray, the \deltao O$_{\text{PDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}.
+In black are two different hypotheses for the trend in the \deltao O$_{\text{SMOW}}$ of the oceans through time.}
 \label{veizer}

### 2013-05-19 draft.tex
fixed some spelling mistakes, added reference to veizer figure

@@
 This carbonate \deltao O proxy has two significant sources of complications.
-First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary envrionmental signature.
+First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature.
 Second, the measured \deltao O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the \deltao O of the seawater.
@@
 The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995,Tajika1993,Wallmann2001} and the magnitude of contintental weathering, which would lead to slow shifts in the \deltao O of the ocean.
-The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controvery in geochemistry.
+The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controversy in geochemistry.
 
@@
 In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease.
-In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low.
+In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low.
 The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}.
@@
 The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite.
-The diffracted beam then interacts with the EBSD detector, depdending on the orientation of the calcite crystal.
+The diffracted beam then interacts with the EBSD detector, depending on the orientation of the calcite crystal.
 The data were analyzed using software from Oxford instruments.
@@
 The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
-The coated flakes were imaged using using the secondary electron detector in the same SEM as above, under high vacuum.
+The coated flakes were imaged using the secondary electron detector in the same SEM as above, under high vacuum.
 The working distance was $\sim 6$ mm and the electron high tension was 10 kV.
@@
 Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
-Quantitative spot measurements were made using the JEOL quantitiative analysis program.
+Quantitative spot measurements were made using the JEOL quantitative analysis program.
 
@@
 $\Delta_{47-\text{RF-AC}}$ values were converted to temperature using the \cite{Ghosh2006} calibration line, which is projected into the absolute reference frame in equation 9 in \cite{Dennis2011}.
-The \deltao O of water was calcuated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.
+The \deltao O of water was calculated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.
 
@@
 
-% array. probably not solid-state reordering, so we interpret rock-buffered diagenesis
-% comparisons support hypothesis that low temp most pristine, EBSD supports, metals dont negate
-% most pristine samples population at 34, one at 30, resample not replicate, no big deal!
-% only upper estimates, but probably pretty close
-% Silurian warm and similar in d18O to today. matches previous clumped studies
-% resolves veizer controversy
-% implications of warm oceans for Ireviken climate rule out big changes, can't resolve small
-% implications of constant d18O?? relative amounts constant. either both constant or coupled.
-
 To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore preserve the most reliable record of Silurian ocean conditions.
@@
 Assuming linear mixing, which is not technically correct but close enough for our purposes, if 5\% of the sample material was recrystallized at 100\degrees C and the rest of the sample material retained an original temperature of 30\degrees C, the measured temperature of the sample would be $\sim34$\degrees C.
-Therefore, the constraints provided by the EBSD mapping of the bachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.
+Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.
 
@@
 
-Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates.
+Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates (Figure \ref{Veizer}).
 Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans \citep{Grossman2012}.
@@
 
-\vspace{0.7cm} \noindent \large \textbf{Acknowldgements} \normalsize
+\vspace{0.7cm} \noindent \large \textbf{Acknowledgements} \normalsize
 
@@
 In black are two different hypotheses for the trend in the \deltao O$_{\text{SMOW}}$ of the oceans through time.}
-\label{veizer}
+\label{Veizer}
 \end{figure}

### 2013-05-20 CIDS_processing.m
initial commit before adding absolute ref frame conversions

@@ -1,0 +1,675 @@
+% Script to import CIDS data into Matlab
+%
+% First I opened the CIDS in excel and saved them as a csv. Then I deleted
+% all the columns to the right of the last column of voltages you copy from
+% isodat. Then I selected all the columns containing the voltages you copy
+% from isodat, clicked on Format in the top menu, and then clicked on
+% Cells. I changed the format to 0.000 which is an option in the custom
+% menu. This makes sure the full decimal accuracy is preserved in the csv
+% when you import it into matlab. Then I saved the CIDS again. I saved each
+% CIDS with a different number in the name, and then added it to the if 
+% else-if statement below % import data. 
+% 
+% If the CIDS is from the 01 autoline, sometimes it has the string 
+%'Rx @90C, 15 min GC, -20C' in the cell next to 'Notes'. This messes up the
+% data import because of the commas, so I used find and replace to delete
+% all these strings. Don't use find and replace to delete all the commas,
+% because this script uses the commas in the Background report to separate
+% the background voltages. 
+%
+% Some older CIDS also have a space before the word 'Background'. This 
+% space will mess up the import, so I used find and replace to delete this 
+% space. 
+%
+% I also deleted the rows of samples from the CIDS files that ended up
+% being way off the D48 line, or anything else funky. 
+% 
+% I also made a file called HG_residuals.csv that has one line of headers
+% and then the heated gas slope and intercept and the average standard
+% residual by date and machine (old or new). 
+% 
+% Made by Renata Cummins in 2013
+
+clear all
+
[…]

### 2013-05-20 CIDS_definitions.m
initial commit

@@ -1,0 +1,204 @@
+% definitions used in the script to import CIDS data into matlab
+
+% define the isotopic compositions of VPDB and VSMOW
+VPDB_R13 = 0.0112372;
+VSMOW_R18 = 0.0020052;
+VSMOW_R17 = 0.0003799;
+
+% define the mass law used to relate R17 to R18
+beta = 0.5164;
+
+% define the locations of variables in the flat list array for each sample
+% S{samp_counter}{12}(1,#)
+reps                = 1;
+aqs                 = 2;
+d13C_PDB            = 3;
+sigma_d13C          = 4;
+sterr_d13C          = 5;
+d18O_gas_SMOW       = 6;
+d18O_min_PDB        = 7;
+sigma_d18O_min      = 8;
+sterr_d18O_min      = 9;
+d47_flat            = 10; 
+sigma_d47           = 11;
+D47                 = 12;
+sigma_D47           = 13;
+sterr_D47           = 14;
+d48_flat            = 15;
+D48_flat            = 16;
+temp                = 17;
+sigma_temp          = 18;
+sterr_temp          = 19;
+d18O_min_SMOW       = 20;
+sigma_d18O_min_SMOW = 21;
+d18O_water          = 22;
+sigma_d18O_water    = 23;
+sterr_d18Ow         = 24;
+height              = 25;
+sample_type         = 26;
+bad                 = 27;
+Fe                  = 28;
+Fe_sterr            = 29;
+Mg                  = 30;
+Mg_sterr            = 31;
+Mn                  = 32;
+Mn_sterr            = 33;
+Sr                  = 34;
+Sr_sterr            = 35;
+
+% define the locations of strings in the summary cell array for each sample
+% S{samp_counter}{11}(#)
+analyst = 1;
+type = 2;
+ID = 3;
+mach_id = 4;
+
+% define locations of variables in the acquisition array for each sample
+% S{samp_counter}{10}(:,#)
+month = 1;
[…]

### 2013-05-20 Fig_ellipses.m
initial commit

@@ -1,0 +1,91 @@
+% A script to plot error ellipses showing the correlated errors in two
+% variables. In this case, the two variables are temperature (temp) and
+% d18O_water. This needs to be in the same folder as your data and the
+% matlab script error_ellipse.m
+% 
+% By Renata Cummins 2013
+
+clear all
+
+test_definitions;
+
+% load processed CIDS data 
+CIDS_data = load('fig_data_ellipses','Samples');
+S = CIDS_data.Samples;
+num_of_samps = size(S,1);
+
+% prepare the figure
+clf;
+hold;
+
+% initalize the counters of the different sample types
+cem_counter = 1;
+mic_counter = 1;
+b_counter = 1;
+r_counter = 1;
+bry_counter = 1;
+cri_counter = 1;
+gas_counter = 1;
+ost_counter = 1;
+
+% calculate and plot the error ellipse for each data point
+for i = 1:num_of_samps
+    
+    cov_mat = cov(S{i}{10}(:,temp_aq),S{i}{10}(:,d18O_water_aq));
+    cov_mat(1,1) = cov_mat(1,1)/S{i}{12}(aqs);
+    cov_mat(2,2) = cov_mat(2,2)/S{i}{12}(aqs);
+    cov_mat(1,2) = cov_mat(1,2)/S{i}{12}(aqs);
+    cov_mat(2,1) = cov_mat(2,1)/S{i}{12}(aqs);
+
+    position = [S{i}{12}(temp) S{i}{12}(d18O_water)];
+    
+    if S{i}{12}(sample_type) == 7;
+        cem_d18Ow(cem_counter) = position(2);
+        cem_temp(cem_counter)  = position(1);
+        cem_counter = cem_counter +1;
+    elseif S{i}{12}(sample_type) == 5;
+        mic_d18Ow(mic_counter) = position(2);
+        mic_temp(mic_counter)  = position(1);
+        mic_counter = mic_counter +1;
+    elseif S{i}{12}(sample_type) == 19;
+        b_d18Ow(b_counter) = position(2);
+        b_temp(b_counter)  = position(1);
+        b_counter = b_counter +1;
+    elseif S{i}{12}(sample_type) == 23;
+        r_d18Ow(r_counter) = position(2);
+        r_temp(r_counter)  = position(1);
+        r_counter = r_counter +1;
[…]

### 2013-05-20 error_ellipse.m
initial commit

@@ -1,0 +1,277 @@
+function h=error_ellipse(varargin)
+% ERROR_ELLIPSE - plot an error ellipse, or ellipsoid, defining confidence region
+%    ERROR_ELLIPSE(C22) - Given a 2x2 covariance matrix, plot the
+%    associated error ellipse, at the origin. It returns a graphics handle
+%    of the ellipse that was drawn.
+%
+%    ERROR_ELLIPSE(C33) - Given a 3x3 covariance matrix, plot the
+%    associated error ellipsoid, at the origin, as well as its projections
+%    onto the three axes. Returns a vector of 4 graphics handles, for the
+%    three ellipses (in the X-Y, Y-Z, and Z-X planes, respectively) and for
+%    the ellipsoid.
+%
+%    ERROR_ELLIPSE(C,MU) - Plot the ellipse, or ellipsoid, centered at MU,
+%    a vector whose length should match that of C (which is 2x2 or 3x3).
+%
+%    ERROR_ELLIPSE(...,'Property1',Value1,'Name2',Value2,...) sets the
+%    values of specified properties, including:
+%      'C' - Alternate method of specifying the covariance matrix
+%      'mu' - Alternate method of specifying the ellipse (-oid) center
+%      'conf' - A value betwen 0 and 1 specifying the confidence interval.
+%        the default is 0.5 which is the 50% error ellipse.
+%      'scale' - Allow the plot the be scaled to difference units.
+%      'style' - A plotting style used to format ellipses.
+%      'clip' - specifies a clipping radius. Portions of the ellipse, -oid,
+%        outside the radius will not be shown.
+%
+%    NOTES: C must be positive definite for this function to work properly.
+
+default_properties = struct(...
+  'C', [], ... % The covaraince matrix (required)
+  'mu', [], ... % Center of ellipse (optional)
+  'conf', 0.683, ... % Percent confidence/100
+  'scale', 1, ... % Scale factor, e.g. 1e-3 to plot m as km
+  'style', 'k', ...  % Plot style
+  'clip', inf); % Clipping radius
+
+if length(varargin) >= 1 && isnumeric(varargin{1})
+  default_properties.C = varargin{1};
[…]

### 2013-05-22 CIDS_definitions.m
added STF slope & intercept, D47_RF, D47_RF_AC

@@ -161,6 +161,10 @@ temp_aq = 103;
 d18O_min_PDB_aq = 104;
 d18O_min_SMOW_aq = 105;
 d18O_water_aq = 106;
+STF_slope = 107;
+STF_int = 108;
+D47_RF = 109;
+D47_RF_AC = 110;
 
 % define the locations of voltages copied from isodat results
 % S{samp_counter}{aq_counter}(:,#)

### 2013-05-22 CIDS_processing.m
changed so that D47, temp, d18Owater is absolute reference frame, need to update STF values

@@ -41,7 +41,7 @@ S = cell(total_samples,1);
 
 % load the heated gas line and standard residuals with corresponding dates
 fileID = fopen('HG_residuals.csv');
-HG_residuals = textscan(fileID, '%s %s %s %s %s %s %s %s %s %s','delimiter',',');
+HG_residuals = textscan(fileID, '%s %s %s %s %s %s %s %s %s %s %s %s','delimiter',',');
 fclose(fileID);
 fd = str2double(HG_residuals{1}(2:end));
 fm = str2double(HG_residuals{2}(2:end));
@@ -50,9 +50,11 @@ ld = str2double(HG_residuals{4}(2:end));
 lm = str2double(HG_residuals{5}(2:end));
 ly = str2double(HG_residuals{6}(2:end));
 mach      = HG_residuals{7}(2:end);
-slope     = str2double(HG_residuals{8}(2:end));
-intercept = str2double(HG_residuals{9}(2:end));
+slope_HG     = str2double(HG_residuals{8}(2:end));
+intercept_HG = str2double(HG_residuals{9}(2:end));
 residual  = str2double(HG_residuals{10}(2:end));
+slope_STF = str2double(HG_residuals{11}(2:end));
+intercept_STF = str2double(HG_residuals{12}(2:end));
 
 for CIDS = 1:number_of_CIDS
 
@@ -89,7 +91,7 @@ number_of_samples = 0;
     for j = 1:9
         S{i}{j} = zeros(7,38);
     end
-    S{i}{10} = zeros(9,106);
+    S{i}{10} = zeros(9,110);
     S{i}{11} = cell(1,4);
     S{i}{12} = zeros(1,35);
   end
@@ -308,27 +310,34 @@ for i = 1:total_samples
     S{i}{10}(:,D48_raw) = S{i}{10}(:,D_48) - 2.*S{i}{10}(:,D_46);
     S{i}{10}(:,D49_raw) = S{i}{10}(:,D_49) - 2.*S{i}{10}(:,D_46) - S{i}{10}(:,D_45);
     
-    % look up the heated gas line and std residuals based on the date
+    % look up HG line and secondary transfer function based on the date
     sd = S{i}{10}(1,day);
[…]

### 2013-07-24 draft.tex
changed to elsarticle, added frontmatter, deleted big tables from the end

@@
-\documentclass{article}
-\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
+\documentclass[preprint, authoryear]{elsarticle}
 \usepackage{amsmath}
-\pagestyle{empty}
 \usepackage{graphicx}
@@
 \usepackage{wasysym}
+\usepackage{natbib}
 \usepackage{lineno}
-\usepackage{natbib}
-\bibpunct{(}{)}{;}{a}{}{,}
-
-\newenvironment{my_enumerate}{
-\begin{enumerate}
-  \setlength{\itemsep}{5pt}
-  \setlength{\parskip}{0pt}
-  \setlength{\parsep}{0pt}}{\end{enumerate}
-}
 
@@
 
-\subsection*{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}
+\begin{frontmatter}
 
-\noindent Cummins, Renata C.$^1$; Finnegan, Seth$^2$; Fike, David A.$^3$; Eiler, John M.$^1$; Fischer, Woodward W.$^1$
+\title{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}
 
-\vspace{10pt}
+\author[caltech]{Renata C. Cummins\corref{cor1}}
+\ead{renatacummins@gmail.com}
 
-\footnotesize
+\author[berkley]{Seth Finnegan}
 
-\noindent $^1$California Institute of Technology, Geological and Planetary Sciences, Pasadena, CA 91125
+\author[washU]{David A. Fike}
 
-\noindent $^2$Department of Integrative Biology, University of California, 1005 Valley Life Sciences Bldg \#3140, Berkeley, CA 94720
+\author[caltech]{John M. Eiler}
 
-\noindent $^3$Department of Earth and Planetary Sciences, Washington University, St. Louis, Missouri 63130
+\author[caltech]{Woodward W. Fischer}
 
-\normalsize
+\cortext[cor1]{Corresponding author}
 
-\linenumbers
+\address[caltech]{California Institute of Technology, Geological and Planetary Sciences, MC 100-23, Pasadena, CA 91125}
 
-\vspace{0.7cm} \noindent \large \textbf{Abstract} \normalsize
+\address[berkley]{Department of Integrative Biology, University of California, 1005 Valley Life Sciences Bldg \#3140, Berkeley, CA 94720}
 
+\address[washU]{Department of Earth and Planetary Sciences, Washington University, St. Louis, Missouri 63130}
+
+\begin{abstract}
 Much of what we know about the history of Earth's climate derives from the geologic record of carbonates.
@@
 This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time.
+\end{abstract}
+
+\end{frontmatter}
+
+\linenumbers
 
@@
 
-\vspace{0.7cm} \noindent \large \textbf{Acknowledgements} \normalsize
+\section*{Acknowledgements}
 
@@
 
-\bibliographystyle{garynat}
+\section*{References}
+\bibliographystyle{my-harv}
 \bibliography{draft}
-
-\begin{sidewaystable}[htbp]
-\centering
-\small
-\begin{tabular}{ | l | c | c | c | c | c | c | c | c | c | c | }
-\multirow{2}{*}{Sample} & Sample & Number of & \deltac C & \deltao O & $\Delta_{47}$ & Temperature & \deltao O water & [Fe] & [Mn] & [Sr] \\
- & location & replicates & ($\permil$ PDB) & ($\permil$ PDB) & ($\permil$) & (\degrees C) & ($\permil$ SMOW) & (ppm) & (ppm) & (ppm)\\ \hline
-G1 -0.3 to -1.0 M1 & 1 & 3 & 2.575 & -5.277 & 0.589 & 38.7 $\pm$ 1.6 & -0.294 $\pm$ 0.30 & 4905 $\pm$ 29 & 327 $\pm$ 3 & 238 $\pm$ 2 \\
-G1 -0.3 to -1.0 R1 & 1 & 3 & 2.319 & -5.320 & 0.596 & 36.8 $\pm$ 1.8 & -0.696 $\pm$ 0.33 & 337 $\pm$ 2 & 159 $\pm$ 1 & 332 $\pm$ 2 \\
-G1 -1.2 to -0.8 B1 & 1 & 1 & 2.019 & -5.523 & 0.565 & 45.0 $\pm$ 2.7 & 0.602 $\pm$ 0.49 & 182 $\pm$ 5 & 66 $\pm$ 2 & 1351 $\pm$ 33 \\
-G1 -3.2 to -3.0 R1 & 1 & 1 & 2.397 & -5.851 & 0.588 & 38.9 $\pm$ 3.1 & -0.833 $\pm$ 0.57 & 1330 $\pm$ 15 & 437 $\pm$ 3 & 428 $\pm$ 3 \\
-G1 -3.4 to -3.2 B2 & 1 & 1 & 2.023 & -5.597 & 0.591 & 38.2 $\pm$ 1.5 & -0.713 $\pm$ 0.27 & 177 $\pm$ 3 & 53 $\pm$ 1 & 1507 $\pm$ 9 \\
-G1 2.2 R1 & 1 & 1 & 3.734 & -5.190 & 0.587 & 39.1 $\pm$ 1.4 & -0.126 $\pm$ 0.25 & 172 $\pm$ 4 & 168 $\pm$ 1 & 431 $\pm$ 5 \\
-G2 -0.2 to 0.0 R1 & 2 & 1 & 2.315 & -5.233 & 0.584 & 39.8 $\pm$ 2.5 & -0.046 $\pm$ 0.47 & 379 $\pm$ 3 & 153 $\pm$ 1 & 414 $\pm$ 5 \\
-G2 -0.5 to -0.25 R1 & 2 & 1 & 2.208 & -5.351 & 0.592 & 37.9 $\pm$ 2.9 & -0.514 $\pm$ 0.53 & 887 $\pm$ 14 & 157 $\pm$ 1 & 600 $\pm$ 2 \\
-G2 0.0 to 0.2 B1 & 2 & 1 & 2.399 & -5.692 & 0.584 & 40.0 $\pm$ 3.7 & -0.462 $\pm$ 0.68 & 302 $\pm$ 9 & 70 $\pm$ 1 & 1289 $\pm$ 23 \\
-G2 0.5 to 0.6 B1 & 2 & 1 & 3.178 & -5.123 & 0.572 & 43.0 $\pm$ 2.8 & 0.639 $\pm$ 0.51 & 46 $\pm$ 3 & 64 $\pm$ 0.4 & 1447 $\pm$ 9 \\
-G2 0.5 to 0.6 B2 & 2 & 1 & 2.925 & -5.347 & 0.602 & 35.4 $\pm$ 1.2 & -0.990 $\pm$ 0.23 & 38 $\pm$ 2 & 27 $\pm$ 1 & 1372 $\pm$ 3 \\
-G3 0.0 B1 & 3 & 1 & 2.248 & -5.876 & 0.573 & 42.7 $\pm$ 2.1 & -0.166 $\pm$ 0.39 & 122 $\pm$ 1 & 48 $\pm$ 0.3 & 1127 $\pm$ 8 \\
-G3 LVF B1 & 3 & 2 & 1.769 & -6.402 & 0.598 & 36.4 $\pm$ 1.9 & -1.850 $\pm$ 0.38 & 124 $\pm$ 2 & 65 $\pm$ 0.4 & 1493 $\pm$ 3 \\
-G3 LVF B1 cement & 3 & 2 & 2.613 & -6.395 & 0.583 & 40.1 $\pm$ 2.9 & -1.155 $\pm$ 0.54 & 953 $\pm$ 9 & 663 $\pm$ 8 & 259 $\pm$ 4 \\
-G3 LVF B1 micrite & 3 & 3 & 1.768 & -5.998 & 0.581 & 40.7 $\pm$ 1.5 & -0.647 $\pm$ 0.29 & 5091 $\pm$ 14 & 457 $\pm$ 2 & 226 $\pm$ 3 \\
-G3 LVF B5 & 3 & 1 & 1.792 & -5.453 & 0.610 & 33.4 $\pm$ 1.6 & -1.479 $\pm$ 0.30 &  $\pm$  &  $\pm$  &  $\pm$  \\
-G3 LVF B5 micrite & 3 & 2 & 1.628 & -5.723 & 0.586 & 39.4 $\pm$ 2.7 & -0.614 $\pm$ 0.49 & 4153 $\pm$ 22 & 449 $\pm$ 3 & 170 $\pm$ 2 \\
-G3 LVF R1 & 3 & 2 & 2.075 & -5.440 & 0.577 & 41.8 $\pm$ 2.7 & 0.112 $\pm$ 0.48 & 724 $\pm$ 7 & 161 $\pm$ 2 & 435 $\pm$ 4 \\
-G3 LVF R2 & 3 & 2 & 5.942 & -5.368 & 0.595 & 37.2 $\pm$ 2.0 & -0.658 $\pm$ 0.37 & 69 $\pm$ 3 & 211 $\pm$ 1 & 309 $\pm$ 5 \\
-G4 2.0 to 3.0 B1 & 4 & 1 & 4.223 & -6.092 & 0.580 & 40.9 $\pm$ 2.8 & -0.712 $\pm$ 0.51 &  nd  &  nd  &  nd  \\
-G4 2.0 to 3.0 bryozoan1 & 4 & 1 & 4.582 & -5.604 & 0.589 & 38.6 $\pm$ 3.8 & -0.635 $\pm$ 0.71 &  nd  &  nd  &  nd  \\
-G4 2.0 to 3.0 crinoid1 & 4 & 1 & 5.033 & -6.216 & 0.564 & 45.2 $\pm$ 3.3 & -0.050 $\pm$ 0.58 &  nd  &  nd  &  nd  \\
-G5 1.6 to 2.2 gastropod1 & 5 & 1 & 5.589 & -6.210 & 0.598 & 36.4 $\pm$ 2.4 & -1.667 $\pm$ 0.45 &  nd  &  nd  &  nd  \\
-G5 1.6 to 2.2 ostracod1 & 5 & 1 & 4.425 & -4.839 & 0.585 & 39.6 $\pm$ 3.3 & 0.309 $\pm$ 0.60 &  nd  &  nd  &  nd  \\
-G6 B234 & 6 & 1 & 4.692 & -5.152 & 0.591 & 38.1 $\pm$ 1.4 & -0.286 $\pm$ 0.26 &  nd  &  nd  &  nd  \\
-G6 crinoid1 & 6 & 1 & 5.457 & -6.380 & 0.535 & 53.3 $\pm$ 4.0 & 1.187 $\pm$ 0.68 &  nd  &  nd  &  nd  \\
-G7 9.0 R1 & 7 & 2 & 5.336 & -5.495 & 0.575 & 42.2 $\pm$ 2.5 & 0.135 $\pm$ 0.45 & 121 $\pm$ 3 & 192 $\pm$ 1 & 86 $\pm$ 6 \\
-G7 9.0 R2 & 7 & 1 & 5.173 & -5.698 & 0.551 & 48.9 $\pm$ 2.6 & 1.120 $\pm$ 0.45 &  nd  &  nd  &  nd  \\
-G8 0.0 to 1.2 R1 & 8 & 1 & -0.507 & -5.208 & 0.593 & 37.7 $\pm$ 2.7 & -0.419 $\pm$ 0.50 & 187 $\pm$ 4 & 78 $\pm$ 1 & 339 $\pm$ 3 \\
-G8 1.2 to 2.0 M1 & 8 & 1 & -0.768 & -7.093 & 0.561 & 46.0 $\pm$ 2.2 & -0.801 $\pm$ 0.39 & 3635 $\pm$ 13 & 248 $\pm$ 1 & 24 $\pm$ 6 \\
-G8 1.2 to 2.0 R1 & 8 & 2 & -0.082 & -5.022 & 0.553 & 48.2 $\pm$ 2.7 & 1.676 $\pm$ 0.47 & 128 $\pm$ 4 & 78 $\pm$ 1 & 191 $\pm$ 3 \\
-\end{tabular} 
-\caption{Isotopic composition, clumped isotope temperature, calculated \deltao O of water and bulk trace metal concentrations of all the samples analyzed in this study.
-Errors are standard error of the mean for all acqusitions of that sample.
-$\text{nd}=\text{no data.}$}
-\label{results}
-\end{sidewaystable}
-
-\begin{sidewaystable}[htbp]
-\centering
-\small
-\begin{tabular}{ | l | c | c | c | c | c | c | c | c | c | c | }
-\multirow{2}{*}{Sample} & Sample & Number of & \deltac C & \deltao O & $\Delta_{47}$ & Temperature & \deltao O water & [Fe] & [Mn] & [Sr] \\
- & location & replicates & ($\permil$ PDB) & ($\permil$ PDB) & ($\permil$) & (\degrees C) & ($\permil$ SMOW) & (ppm) & (ppm) & (ppm)\\ \hline
-G11 0.0 to 0.2 R1 & 11 & 1 & 3.164 & -4.847 & 0.599 & 36.1 $\pm$ 2.4 & -0.352 $\pm$ 0.46 & 116 $\pm$ 4 & 140 $\pm$ 1 & 769 $\pm$ 4 \\
-G11 0.2 to 0.4 B2 & 11 & 1 & 2.836 & -5.360 & 0.587 & 39.2 $\pm$ 3.6 & -0.289 $\pm$ 0.67 & 252 $\pm$ 6 & 59 $\pm$ 1 & 1424 $\pm$ 15 \\
-G11 0.2 to 0.4 B6 & 11 & 1 & 2.454 & -6.146 & 0.605 & 34.6 $\pm$ 2.3 & -1.939 $\pm$ 0.43 & 89 $\pm$ 2 & 107 $\pm$ 1 & 1450 $\pm$ 8 \\
-G11 0.2 to 0.4 R1 & 11 & 1 & 3.169 & -5.495 & 0.569 & 43.9 $\pm$ 2.2 & 0.440 $\pm$ 0.40 & 487 $\pm$ 7 & 346 $\pm$ 2 & 481 $\pm$ 4 \\
-G11 0.4 to 0.6 B1 & 11 & 1 & 2.766 & -5.439 & 0.596 & 36.9 $\pm$ 1.6 & -0.788 $\pm$ 0.30 & 58 $\pm$ 1 & 40 $\pm$ 0.2 & 1476 $\pm$ 4 \\
-G11 0.6 B1 & 11 & 1 & 2.927 & -5.269 & 0.588 & 38.9 $\pm$ 1.8 & -0.242 $\pm$ 0.33 & 59 $\pm$ 2 & 59 $\pm$ 1 & 1627 $\pm$ 11 \\
-G11 0.6 to 0.8 M1 & 11 & 1 & 2.431 & -6.468 & 0.610 & 33.5 $\pm$ 2.0 & -2.472 $\pm$ 0.39 & 20948 $\pm$ 90 & 288 $\pm$ 7 & 79 $\pm$ 3 \\
-G11 0.6 to 0.8 R1 & 11 & 1 & 3.447 & -4.886 & 0.596 & 36.8 $\pm$ 3.5 & -0.258 $\pm$ 0.65 & 75 $\pm$ 4 & 68 $\pm$ 1 & 503 $\pm$ 6 \\
-G12 B3 & 12 & 1 & 2.803 & -5.757 & 0.558 & 46.9 $\pm$ 1.8 & 0.707 $\pm$ 0.32 &  nd  &  nd  &  nd  \\
-G12 B4 & 12 & 1 & 4.397 & -5.179 & 0.581 & 40.8 $\pm$ 3.7 & 0.191 $\pm$ 0.67 & 64 $\pm$ 3 & 41 $\pm$ 1 & 1358 $\pm$ 4 \\
-G12 R1 & 12 & 3 & 2.381 & -5.221 & 0.604 & 35.0 $\pm$ 1.6 & -0.935 $\pm$ 0.31 & 601 $\pm$ 6 & 124 $\pm$ 1 & 412 $\pm$ 4 \\
-G12 R1 micrite & 12 & 1 & 2.102 & -5.048 & 0.568 & 44.2 $\pm$ 3.8 & 0.939 $\pm$ 0.69 &  nd  &  nd  &  nd  \\
-G12 R2 & 12 & 1 & 2.043 & -5.611 & 0.582 & 40.6 $\pm$ 2.8 & -0.287 $\pm$ 0.52 & 1474 $\pm$ 18 & 417 $\pm$ 2 & 342 $\pm$ 2 \\
-G13 1.0 to 2.0 R1 & 13 & 1 & -0.568 & -4.974 & 0.582 & 40.4 $\pm$ 4.4 & 0.320 $\pm$ 0.81 & 26 $\pm$ 4 & 107 $\pm$ 1 & 200 $\pm$ 3 \\
-G13 10.5 to 10.8 B1 & 13 & 1 & -0.598 & -4.986 & 0.626 & 29.6 $\pm$ 2.4 & -1.741 $\pm$ 0.48 & 32 $\pm$ 5 & 16 $\pm$ 2 & 1372 $\pm$ 109 \\
-G13 10.5 to 10.8 B1 leftovers & 13 & 1 & -0.722 & -5.207 & 0.579 & 41.3 $\pm$ 4.4 & 0.251 $\pm$ 0.81 &  nd  &  nd  &  nd  \\
-G13 10.5 to 10.8 M1 & 13 & 1 & -1.496 & -6.298 & 0.547 & 50.0 $\pm$ 2.7 & 0.703 $\pm$ 0.46 & 3487 $\pm$ 7 & 277 $\pm$ 1 & 31 $\pm$ 4 \\
-G13 10.5 to 10.8 R1 tip & 13 & 2 & -0.564 & -4.810 & 0.594 & 37.2 $\pm$ 2.0 & -0.097 $\pm$ 0.37 &  nd  &  nd  &  nd  \\
-G13 10.5 to 10.8 R1 top & 13 & 2 & -1.044 & -5.829 & 0.583 & 40.3 $\pm$ 1.5 & -0.557 $\pm$ 0.28 & 571 $\pm$ 65 & 193 $\pm$ 13 & 232 $\pm$ 15 \\
-G13 31.4 to 32.4 B1 & 13 & 3 & -0.316 & -5.082 & 0.592 & 37.9 $\pm$ 1.8 & -0.243 $\pm$ 0.34 & 141 $\pm$ 3 & 27 $\pm$ 0.3 & 1436 $\pm$ 11 \\
-G13 31.4 to 32.4 M1 & 13 & 1 & -0.778 & -5.965 & 0.570 & 43.5 $\pm$ 2.6 & -0.109 $\pm$ 0.47 & 3671 $\pm$ 10 & 237 $\pm$ 1 & 42 $\pm$ 3 \\
-G13 31.4 to 32.4 R1 & 13 & 1 & -0.809 & -5.484 & 0.535 & 53.4 $\pm$ 4.0 & 2.103 $\pm$ 0.68 & 96 $\pm$ 7 & 143 $\pm$ 1 & 118 $\pm$ 6 \\
-G13 33.5 B1 bad & 13 & 1 & -0.556 & -5.253 & 0.576 & 42.1 $\pm$ 3.3 & 0.361 $\pm$ 0.61 & 107 $\pm$ 1 & 46 $\pm$ 0.3 & 1320 $\pm$ 5 \\
-G13 33.5 B1 leftovers & 13 & 1 & -0.191 & -5.338 & 0.575 & 42.4 $\pm$ 3.0 & 0.321 $\pm$ 0.54 & 19 $\pm$ 5 & 14 $\pm$ 2 & 983 $\pm$ 109 \\
-G13 33.5 B1 perfect & 13 & 1 & -0.159 & -5.407 & 0.596 & 36.9 $\pm$ 3.0 & -0.756 $\pm$ 0.56 &  nd  &  nd  &  nd  \\
-G13 35.6 to 36.0 B1 & 13 & 1 & -0.288 & -5.397 & 0.580 & 41.0 $\pm$ 1.9 & 0.011 $\pm$ 0.35 & 39 $\pm$ 1 & 14 $\pm$ 0.1 & 1657 $\pm$ 9 \\
-G13 5.0 to 6.0 B3 & 13 & 1 & -0.216 & -5.614 & 0.576 & 42.1 $\pm$ 3.5 & -0.001 $\pm$ 0.64 &  nd  &  nd  &  nd  \\
-G13 5.0 to 6.0 B3 cement & 13 & 1 & -0.909 & -7.169 & 0.564 & 45.3 $\pm$ 2.2 & -0.996 $\pm$ 0.39 &  nd  &  nd  &  nd  \\
-G13 6.5 to 7.5 M1 & 13 & 1 & -0.118 & -5.206 & 0.559 & 46.7 $\pm$ 3.9 & 1.229 $\pm$ 0.69 & 5708 $\pm$ 34 & 289 $\pm$ 3 & 0 $\pm$ 2 \\
-G13 6.5 to 7.5 R1 & 13 & 1 & 0.180 & -4.728 & 0.536 & 53.3 $\pm$ 2.4 & 2.847 $\pm$ 0.41 & 32 $\pm$ 2 & 80 $\pm$ 1 & 204 $\pm$ 6 \\
-G13 UF B1 & 13 & 3 & -0.347 & -5.416 & 0.598 & 36.4 $\pm$ 1.4 & -0.857 $\pm$ 0.28 & 27 $\pm$ 1 & 35 $\pm$ 0.5 & 1564 $\pm$ 7 \\
-G13 UF B1 cement & 13 & 2 & -0.868 & -7.730 & 0.578 & 41.4 $\pm$ 2.0 & -2.259 $\pm$ 0.37 & 234 $\pm$ 2 & 266 $\pm$ 2 & 100 $\pm$ 1 \\
-G13 UF B1 micrite & 13 & 2 & -1.024 & -6.330 & 0.543 & 51.1 $\pm$ 2.2 & 0.852 $\pm$ 0.38 & 1428 $\pm$ 6 & 251 $\pm$ 2 & 126 $\pm$ 2 \\
-\end{tabular} 
-\end{sidewaystable}
 

### 2013-07-29 draft.tex
in draft.tex, included newest versions of figs, re-wrote fig captions, also made two column

@@
-\documentclass[preprint, authoryear]{elsarticle}
+\documentclass[5p, authoryear]{elsarticle}
 \usepackage{amsmath}
@@
 \usepackage{lineno}
+\usepackage{dblfloatfix}
+
+\renewcommand{\topfraction}{0.99}
+\renewcommand\bottomfraction{.99}
+\renewcommand{\textfraction}{0.01}
+\renewcommand{\floatpagefraction}{0.99}
+\setcounter{totalnumber}{50}
+\setcounter{topnumber}{50}
+\setcounter{bottomnumber}{50}
 
@@
 \end{frontmatter}
-
-\linenumbers
 
@@
 
+\begin{figure*}[t]
+\centering
+\includegraphics[width=140mm]{Fig_01-01.png}
+\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}.
+Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in the supplementary table.}
+\label{map}
+\end{figure*}
+
 The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones and were likely deposited below storm wave base (Figure \ref{column}).
@@
 Fossils are abundant in the Upper Visby Beds \citep{Calner2004a,Samtleben1996}.
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=140mm]{Fig_02-01.png}
+\caption{Lithology and isotope stratigraphy of the sampled sequence.
+On the left, the stratigraphic column shows the generalized lithostratigraphic units that were sampled for this study.
+On the right are the carbon and oxygen isotopic compositions of apparently well-preserved brachiopods, collected from the same section in Gotland by \cite{Bickert1997}.
+Figure modified from \cite{Bickert1997}.}
+\label{column}
+\end{figure*}
 
@@
 
+\begin{figure*}[b]
+\centering
+\includegraphics[width=\textwidth]{Fig_03_color-01.png}
+\caption{The \deltac C, \deltao O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
+For the Lower \& Upper Visby and H\"{o}gklint Formations, a height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds.
+For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop.
+Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included.
+Error bars represent the standard error of the mean.}
+\label{section}
+\end{figure*}
+
 The \deltao O values of fossil material are consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=90mm]{Fig_04_color-01.png}
+\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
+The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by \cite{Brand2012} which are also shown in this figure.
+Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods.
+Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.
+Figure modified from \cite{Brand2012}.}
+\label{Brand}
+\end{figure}
+
 There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration.
@@
 
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=\textwidth]{Fig_05_color-01.png}
+\caption{Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
+The negative correlation between temperature and the log of Sr concentration in rugose corals is statistically robust, and holds up when combined with the rest of the samples, but no other correlations passed our hypothesis tests.
+These plots demonstrate clear differences in trace metal content between the two fossil taxa and the later carbonate phases.
+Brachiopods tend to have the most pristine trace metal signatures, with less Mn and more Sr than the other samples.
+Micrites and cements tend to have the most altered trace metal signatures, with high Fe and Mn and low Sr concentrations.
+}
+\label{metals}
+\end{figure*}
+
 E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization may not actually have any effect on Fe and Mn concentrations.
@@
 If recrystallization is an important contributor to the elevated temperatures of fossil material, trace metals may not be expected to be reliable indicators of diagenetic alteration.
+
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=140mm]{Fig_06_color-01.png}
+\caption{Calcite fabric and trace metal concentrations in a rugose coral.
+The four images show the same area of a rugose coral thin section, all at the same magnification.
+The black bar at the bottom of panel B is 500 $\mu$m long for scale.
+A) Transmitted light.
+The dark triangle labelled ``RC'' was identified as a recrystallized domain of the coral septum.
+Cement in the void space between coral septa appears as blocky calcite crystals.
+B) Electron backscatter diffraction (EBSD) analysis.
+The orientation of calcite crystals is indicated by different shades.
+The recrystallized domain is clearly marked by many calcite crystals with the same orientation.
+C) Iron and D) manganese concentrations measured using the E-probe.
+The maps are qualitative, with darker shades representing higher concentrations.
+For both metals, the recrystallized domain is indistinguishable from the rest of the septum.
+The black lines show the location of an array of 62 spots where trace metal concentrations were measured quantitatively.
+The concentration of Fe and Mn at these spots is plotted to the right of the respective map.}
+\label{EBSD_coral}
+\end{figure*}
 
@@
 
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=\textwidth]{Fig_07_color-01.png}
+\caption{Calcite fabric in a brachiopod.
+A) Transmitted polarized light image of a thin section of brachiopod G3 LVF B1.
+This sample has a clumped isotope temperature of 35\degrees C.
+The boxed overlay is an electron backscatter diffraction (EBSD) image of the same sample.
+The orientation of calcite crystals is indicated by different shades.
+B) Euler pole figure generated from the EBSD analysis.
+The central band represents the preferential orientation of calcite crystals parallel to the outer surface of the brachiopod shell.}
+\label{EBSD_brach}
+\end{figure*}
+
 Scanning electron microscope (SEM) images of low-temperature brachiopod flakes also demonstrate excellent preservation of original fibrous calcite textures (Figure \ref{flakes}A, B \& C).
@@
 Therefore, although the representative flakes chosen for inspection with the SEM appear pristine, this does not guarantee that all recrystallized material was excluded from the samples.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=90mm]{Fig_08-01.png}
+\caption{Scanning electron microscope images of flakes of brachiopod shell.
+Images were taken in secondary electron mode.
+Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
+Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
+Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2\degrees C; B) 33 $\pm$ 2\degrees C; C) 35 $\pm$ 1\degrees C.}
+\label{flakes}
+\end{figure}
 
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=90mm]{Fig_09-01.png}
+\caption{The clumped isotope temperature and calculated water \deltao O$_{\text{VSMOW}}$ for all samples analyzed in this study.
+Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
+\label{ellipses}
+\end{figure}
+
 Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis, but we rule it out as a major contributor to our results.
@@
 Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=90mm]{Fig_10-01.png}
+\caption{Comparisons between the \deltac C, \deltao O and clumped isotope temperature of brachiopod samples and adjacent micrite.}
+\label{comparisons}
+\end{figure}
 
@@
 Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates (Figure \ref{Veizer}).
-Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans \citep{Grossman2012}.
+Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans.
 Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans.
@@
 
-The results from this study join those of previous clumped isotope studies in implying that the \deltao O$_{\text{SMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$.
+The results from this study join those of previous clumped isotope studies in implying that the \deltao O$_{\text{SMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$ (Figure \ref{Veizer}).
 Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}.
@@
 This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=90mm]{Fig_11-01.png}
+\caption{Implications for the constancy of seawater $\delta^{18}$O through time.
+In gray, the \deltao O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}.
+The black solid line represents the estimated \deltao O$_{\text{VSMOW}}$ of the oceans through time based on the calcite values.
+The two black dashed lines represent alternative hypotheses for the value of seawater \deltao O$_{\text{VSMOW}}$ through time.
+All three lines ignore short-term perturbations  to seawater $\delta^{18}$O caused by glaciations.
+Finally, the black shapes are estimates of seawater \deltao O from clumped isotopes, measured at four different times by three separate studies.
+These clumped isotope estimates suggest that seawater \deltao O has not changed by more than 2$\permil$ in the last 430 million years.}
+\label{Veizer}
+\end{figure}
 
@@
 
-\newpage
+\nocite{*}
 
 \section*{References}
-\bibliographystyle{my-harv}
+\bibliographystyle{my-geochim}
 \bibliography{draft}
 
-\begin{figure}[p]
-\centering
-\includegraphics[width=0.5\textwidth]{Fig_map.pdf}
-\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}.
-Approximate sample locations are marked by colored circles.
-Numbering corresponds to the sample location in Table \ref{results}.}
-\label{map}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.9\textwidth]{Fig_column-01.jpg}
-\caption{On the left, a stratigraphic column showing the generalized lithstratigraphic units that were sampled for this study.
-On the right are the carbon and oxygen isotopic compositions of apparently well-preserved brachiopods, collected from the same section in Gotland.
-Isotopic compositions of brachiopods are modified from \cite{Bickert1997}.}
-\label{column}
-\end{figure}
-
-\begin{sidewaysfigure}[htb]
-\centering
-\includegraphics[width=\textwidth]{Fig_section-01.jpg}
-\caption{The \deltac C, \deltao O and $\Delta_{47}$ temperature of all the samples analyzed in this study, plotted against stratigraphic height.
-A height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds.
-Also included are data from Munnecke et al.
-(2003).}
-\label{section}
-\end{sidewaysfigure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\textwidth]{Fig_Brand-01.jpg}
-\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
-The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by Brand et al.
-(2012), which are also shown in this figure.
-Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods.
-Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.}
-\label{Brand}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\textwidth]{Fig_metals-01.jpg}
-\caption{Scatter plots showing the relationship between temperature and bulk trace metal concentration.
-The only statistically robust correlation ($\text{p}<0.05$ compared to zero correlation) is that between temperature and Sr concentration of rugose corals.}
-\label{metals}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.8\textwidth]{Fig_EBSD_coral-01.jpg}
-\caption{Four different images of the same area of a rugose coral thin section, all at the same magnification.
-A) Reflected light.
-The line of dark pits are damaged caused by the E-probe electron beam.
-The dark triangle labelled ``RC'' was identified as a recrystallized domain of the coal septum.
-B) EBSD analysis.
-The recrystallized domain is clearly marked by many calcite crystals with the same orientation.
-The black scale bar represents 500 $\mu$m.
-C) Iron and D) manganese concentrations measured using the E-probe.
-The maps are qualitative, with warmer colors representing higher concentrations.
-The overlaid spots were measured quantitatively, and are labeled in units of ppm.
-For both metals, the recrystallized domain is indistinguishable from the rest of the septum.}
-\label{EBSD_coral}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\textwidth]{Fig_EBSD_brach-01.jpg}
-\caption{A) Transmitted polarized light image of a thin section of a brachiopod with $\Delta_{47}$ temperature 34\degrees C.
-The color overlay is an electron backscatter diffraction image of the same sample.
-The orientation of calcite crystals is indicated by color.
-B) Euler pole figure generated from the EBSD analysis.}
-\label{EBSD_brach}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=\textwidth]{Fig_flakes-01.jpg}
-\caption{Scanning electron microscope images of flakes of brachiopod shell.
-Images were taken in secondary electron mode.
-Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
-Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
-Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 34 $\pm$ 2\degrees C; B) 35 $\pm$ 3\degrees C; C) 35 $\pm$ 3\degrees C.}
-\label{flakes}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.7\textwidth]{Fig_ellipses-01.jpg}
-\caption{The $\Delta_{47}$ temperature and calculated water \deltao O for all samples analyzed in this study.
-Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
-\label{ellipses}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.5\textwidth]{Fig_comparisons-01.jpg}
-\caption{Comparisons between the \deltac C, \deltao O and $\Delta_{47}$ temperature of brachiopod samples and adjacent micrite.}
-\label{comparisons}
-\end{figure}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.6\textwidth]{Fig_veizer-01.jpg}
-\caption{In gray, the \deltao O$_{\text{PDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}.
-In black are two different hypotheses for the trend in the \deltao O$_{\text{SMOW}}$ of the oceans through time.}
-\label{Veizer}
-\end{figure}
-
 \end{document}

### 2013-08-02 draft.tex
draft has been converted from word. All figures in good places. ready to send to Woody

@@
 \usepackage{dblfloatfix}
+\usepackage[skip=2pt,font=footnotesize]{caption}
+\usepackage{breakcites}
 
@@
 
-\newcommand{\deltao}{$\delta^{18}$}
-\newcommand{\deltac}{$\delta^{13}$}
-\newcommand{\degrees}{$^{\circ}$}
-
 \begin{document}
@@
 
-\title{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}
+\title{Carbonate clumped isotope constraints on Silurian ocean temperature and $\delta^{18}$O}
 
@@
 Much of what we know about the history of Earth's climate derives from the geologic record of carbonates.
-The oxygen isotopic composition (\deltao O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the \deltao O of seawater in the past.
-Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the \deltao O of the water as well.
+The oxygen isotopic composition ($\delta^{18}$O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past.
+Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the $\delta^{18}$O of the water as well.
 The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set.
-These highly pristine samples ultimately provide a more reliable record of paleoclimate.
 
-We tested the ability of carbonate clumped isotopes to constrain paleoclimate by measuring the clumped isotope composition of apparently well-preserved calcitic fossils from several Silurian-age (ca.
-427 Ma) stratigraphic sections on the island of Gotland, Sweden.
-Variability in clumped isotope temperatures suggests differential preservation, complicating a resolved temperature history through these sections.
-Micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm ($30\pm 2$\degrees C) and similar in oxygen isotopic composition to a modern ice-free world (\deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$).
+We evaluated preservation and measured the clumped isotope composition of a suite of calcitic fossils from Silurian (ca.
+433 Ma) stratigraphic sections on the island of Gotland, Sweden.
+Substantial variability in clumped isotope temperatures suggests differential preservation, complicating the generation of a stratigraphically resolved temperature history through these sections.
+However, micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to a modern ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$).
 This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time.
@@
 
-Despite the importance of paleoclimate and more than half a century of investigations on the topic, the temperature history of Earth's oceans remains poorly constrained.
-The most widely-used proxy for past ocean temperatures is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate \deltao O thermometer, which was developed by Harold Urey in the 1940s \citep{Urey1947}.
-This carbonate \deltao O proxy has two significant sources of complications.
+Despite more than half a century of investigations on the topic, the long-term temperature history of Earth's oceans remains poorly constrained.
+The most widely used proxy for past ocean temperature is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate $\delta^{18}$O thermometer \citep{Urey1947}.
+This thermometer is based on the temperature dependence of the fractionation between the $\delta^{18}$O of carbonate, reported relative to the VPDB standard, and the $\delta^{18}$O of seawater, reported relative to the VSMOW standard.
+The carbonate $\delta^{18}$O proxy has two significant sources of complications.
 First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature.
-Second, the measured \deltao O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the \deltao O of the seawater.
-The \deltao O of seawater can be affected by several factors, including evaporation, mixing with freshwater, and changes in the size of continental glaciers, all of which are difficult to quantify in the geologic record.
-The
-\deltao O of the ocean has also been hypothesized to change gradually but substantially through Earth's history \citep{Jaffres2007,Veizer1999}.
+Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the $\delta^{18}$O of the seawater.
+The $\delta^{18}$O of seawater can be influenced by several processes, including local evaporation and mixing with meteoric water, as well as global changes in the size of continental glaciers, all of which are difficult to constrain in the geologic record.
 
-Gradual changes in the \deltao O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the \deltao O of seawater.
-The long-term average \deltao O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust.
-High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the \deltao O of ocean water by equilibrating the water with the average mantle \deltao O of 5.7$\permil$ \citep{Gregory1991}.
-Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the \deltao O of seawater by preferentially moving $^{18}$O from water into rocks.
-The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995,Tajika1993,Wallmann2001} and the magnitude of contintental weathering, which would lead to slow shifts in the \deltao O of the ocean.
-The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controversy in geochemistry.
+The $\delta^{18}$O of the ocean has also been hypothesized to have changed gradually but substantially through Earth's history \citep{Jaffres2007, Veizer1999}.
+Gradual changes in the $\delta^{18}$O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the $\delta^{18}$O of seawater.
+The long-term average $\delta^{18}$O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust.
+High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the $\delta^{18}$O of ocean water by equilibrating the water with the average mantle $\delta^{18}$O$_{\text{VSMOW}}$ of 5.7$\permil$ \citep{Gregory1991}.
+Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the $\delta^{18}$O of seawater by preferentially moving $^{18}$O from water into rocks.
+The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995, Tajika1993, Wallmann2001} and the magnitude of continental weathering, which would lead to slow shifts in the $\delta^{18}$O of the ocean.
+The best geologic evidence for gradual change in seawater $\delta^{18}$O, however, is the subject of a long-standing controversy in geochemistry.
 
-This controversy arises from the complications associated with the carbonate \deltao O proxy, which allow conflicting interpretations of the $\sim 8\permil$ increase in the average \deltao O of marine carbonates from the Cambrian to today.
-Several studies have proposed that early Paleozoic carbonates may have low \deltao O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959,Keith1964,Land1995}, which tends to decrease the \deltao O of carbonate \citep{Marshall1992}.
-In response to this concern, \cite{Veizer1999} measured the \deltao O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics.
-These screened fossils also record increasing \deltao O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low \deltao O values in the early Paleozoic.
-An alternative hypothesis holds that oceans could have been much warmer in the early Paleozoic \citep{Karhu1986,Knauth1976}.
-Assuming that the \deltao O of the oceans was the same as today, the early Paleozoic oceans must have been $\sim 56$\degrees C to produce carbonates with a \deltao O$_{\text{PDB}}$ of $-8\permil$.
-These warm temperatures seem incompatible with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading some studies to suggest a third hypothesis, that the \deltao O$_{\text{SMOW}}$ of seawater has increased from $-8\permil$ in the Cambrian to its present value of $\sim 0\permil$ \citep{Jaffres2007,Veizer1999}.
+This controversy arises from the complications associated with the carbonate $\delta^{18}$O proxy, which allow conflicting interpretations of the $\sim$8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}.
+Several studies have proposed that early Paleozoic carbonates may have low $\delta^{18}$O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959, Keith1964, Land1995}, which tends to decrease the $\delta^{18}$O of carbonate \citep{Marshall1992}.
+In response to this concern, \cite{Veizer1999} measured the $\delta^{18}$O of calcite fossils that were screened for diagenetic alteration using trace metal compositions and micro-scale studies of calcite fabrics.
+These screened fossils also record increasing $\delta^{18}$O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low $\delta^{18}$O values in the early Paleozoic.
+An alternative hypothesis holds that oceans could have been much warmer in the early Paleozoic \citep{Karhu1986, Knauth1976}.
+Assuming that the $\delta^{18}$O of the oceans was the same as today, the Cambrian oceans must have been $\sim$56$^{\circ}$C to produce carbonates with a $\delta^{18}$O$_{\text{VPDB}}$ of -8$\permil$.
+These warm temperatures seem incompatible with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading some studies to suggest that the increase in the $\delta^{18}$O of marine carbonates reflects a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.
 
-Carbonate clumped isotope thermometry provides a unique insight into the controversy described above because it resolves the two main complications of the carbonate \deltao O proxy.
-The clumped isotope composition of a carbonate records a temperature that is independent of the \deltao O of water and highly sensitive to diagenetic alteration.
-The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates.
-For a shelly marine organism that lives in shallow water, the degree of clumping in the shell calcite will reflect the sea surface temperature \citep{Ghosh2006}.
-Measurement of the \deltao O of the shell calcite in addition to the temperature allows calculation of the \deltao O of the seawater, using the experimentally-determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}.
-If the calcite shell is then buried and fossilized, it is often altered in the presence of high temperatures.
-This alteration can take the form of recrystallization or thermally-activated re-ordering \citep{Eiler2011}.
-In either case, the altered material will take on a new clumped-isotope temperature that can be much higher than the temperature of the pristine fossil calcite.
-If the altered material is mixed with pristine fossil material during sample processing, the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived.
-We demonstrate here that carbonate clumped isotopes can be a more sensitive indicator of diagenetic alteration than the \deltao O or the trace metal content of a carbonate.
+Carbonate clumped isotope thermometry provides a unique insight into this controversy because it provides data that can help to resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate records a temperature that is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration.
+The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}.
+Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the seawater.
 
-In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set.
+Shells that have been deeply buried in lithified sediments have often experienced high-temperature alteration, which can change the $\delta^{18}$O of the carbonate and is frequently difficult to detect.
+Clumped isotopes are sensitive to this type of alteration because the altered material will take on a new clumped isotope temperature that can be much higher than the temperature of the pristine fossil calcite.
+If the altered material is incorporated within pristine fossil material selected for sample analysis (e.g.
+as cements lining fossil void space), the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived.
+We demonstrate here that carbonate clumped isotopes can be a more sensitive indicator of diagenetic alteration than the $\delta^{18}$O or the trace metal content of a carbonate.
+
+In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope arrays are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine end-member of a sample set.
 Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water.
 As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water.
-During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected.
-In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease.
-In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low.
+During rock-buffered diagenesis, this causes the $\delta^{18}$O of the water to increase, while the $\delta^{18}$O of the altered carbonate is not affected.
+In a water-buffered diagenetic regime, increasing temperature instead causes the $\delta^{18}$O of the altered carbonate to decrease.
+In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the $\delta^{18}$O of both the carbonate and the water are diagnostically low.
 The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}.
-We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.
+We demonstrate here that this distinction has important implications for the estimation of the temperature and $\delta^{18}$O of oceans in the past.
 
@@
 This study also provides a test of established proxies for diagenetic alteration as filters for clumped isotope data.
-Finally, our data add to the clumped isotope record of past ocean temperatures and \deltao O, including samples from a different continent and a different paleocean basin than previous clumped isotope studies.
+Finally, our data add to the clumped isotope record of past ocean temperatures and $\delta^{18}$O, including samples from a different continent and a different paleocean basin than previous clumped isotope studies.
 
@@
 
-Samples were collected from a section exposed on the island of Gotland, Sweden, that spans the latest Llandovery through the Wenlock in the Early Silurian (Figure \ref{map}).
+Samples were collected from several sections exposed on the island of Gotland, Sweden that span the latest Llandovery through Wenlock epochs of the Silurian Period (Figure \ref{map}).
 Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}.
-The section records shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was within 30$^{\circ}$ of the equator during the Silurian \citep{Torsvik1992}.
+The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was 19$^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.
 
@@
 \caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}.
-Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in the supplementary table.}
+Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in Supplementary Table 1.}
 \label{map}
@@
 
-The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones and were likely deposited below storm wave base (Figure \ref{column}).
-Fossils are present but not abundant \citep{Calner2004a,Samtleben1996}.
-The contact between Lower and Upper Visby beds is marked by a bed with a dense population of \textit{Phaulactis}, a species of solitary rugose coral \citep{Jeppsson1997,Jeppsson2006,Munnecke2003,Samtleben1996}.
-This contact is just above the Llandovery-Wenlock boundary, based on conodont stratigraphy \citep{Aldridge1993,Jeppsson1983,Mabillard1985}.
-During sampling, we defined this contact as a stratigraphic height of zero (Figure \ref{column}).
-In addition to the bed of abundant \textit{Phaulactis} corals, the contact between Lower and Upper Visby Beds is also marked by a facies change that represents a maximum flooding surface \citep{Calner2004b}.
-Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of detritic limestones and reef mounds.
+The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones deposited below storm wave base (Figure \ref{column}).
+Fossils are abundant throughout the section \citep{Calner2004a, Samtleben1996}.
+The contact between Lower and Upper Visby beds is marked by a bed with abundant fossils of the solitary rugose coral genus \textit{Phaulactis} \citep{Jeppsson1997, Jeppsson2006, Munnecke2003, Samtleben1996}.
+This contact is just above the Llandovery-Wenlock boundary as defined by conodont biostratigraphy \citep{Aldridge1993, Jeppsson1983, Mabillard1985}.
+We defined the \textit{Phaulactis} bed as a datum and all stratigraphic positions are measured relative to it, except those in the Slite Formation, which are measured relative to an arbitrary bed in the Slite Formation.
+The contact between Lower and Upper Visby Beds is also characterized by a facies change that represents a maximum flooding surface \citep{Calner2004b}.
+Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of bioclastic limestones and reef mounds.
 Ripple marks indicate deposition above storm wave base.
-Fossils are abundant in the Upper Visby Beds \citep{Calner2004a,Samtleben1996}.
+Fossils are especially abundant in the Upper Visby Beds \citep{Calner2004a, Samtleben1996}.
 
@@
 
-Above the Upper Visby Beds, the H\"{o}gklint Formation consists of algal and crinoidal limestones with some reef mounds, deposited above storm wave base \citep{Riding1991,Samtleben1996,Watts2000}.
+Above the Upper Visby Beds, the H\"{o}gklint Formation consists of algal and crinoidal limestones with some reef mounds, deposited above storm wave base \citep{Riding1991, Samtleben1996, Watts2000}.
 The contact between the H\"{o}gklint Formation and the overlying Tofta Formation represents a hypothesized sequence boundary \citep{Calner2004b}.
-The Tofta Formation consists of oncoid-rich bedded limestone, and likely represents a shallow restricted setting, such as a backreef lagoon \citep{Riding1991,Samtleben1996}.
+The Tofta Formation is characterized by oncoid-rich bedded limestone, and likely represents a shallow restricted setting, such as a backreef lagoon \citep{Riding1991, Samtleben1996}.
 The overlying Hangvar Formation consists of interbedded marls and limestones with some reef mounds, indicating deposition in a less restricted, deeper-water setting.
-The youngest strata sampled for this study were from the Slite Formation, which transitions from limestone-rich to marl-rich interbedded limestones and marls, likely representing deposition in deeper water.
-Fossils are present throughout the section \citep{Calner2004a}.
+The youngest strata sampled for this study were from the Slite Formation, which transitions from limestone-rich to marl-rich interbedded limestones and marls, likely representing deposition in deeper water \citep{Calner2004a}.
 
 The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons.
-First, this section has long been recognized for its excellent preservation.
-The strata on Gotland are very nearly flat-lying, with a dip of less than 1\degrees, and display only minimal evidence of tectonic disturbance \citep{Calner2004a,Jeppsson1983}.
-Conodont fossils from this section are pale yellow in color, giving them a conodont alteration index (CAI) of one, which corresponds to a maximum temperature of $\sim$100\degrees C \citep{Epstein1976,Jeppsson1983,Wenzel2000}.
+First, this section has long been recognized for its generally excellent preservation.
+The strata on Gotland are very nearly flat-lying, with a dip of less than 1$^{\circ}$, and display only minimal evidence of tectonic disturbance \citep{Calner2004a, Jeppsson1983}.
+Conodont fossils from this section yield a conodont alteration index (CAI) of one, which corresponds to a maximum temperature of $\sim$100$^{\circ}$C \citep{Epstein1976, Jeppsson1983, Wenzel2000}.
 
-The Silurian section on Gotland is also attractive due to the relative abundance of fossils, especially brachiopods, throughout the section.
-By sampling fossil material for clumped isotope measurements, we target calcite that was first precipitated directly from seawater.
-All brachiopod samples were taken from the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998,Samtleben2001}.
-The fibrous calcite fabric characteristic of the secondary layer has been well-characterized in modern and fossil brachiopods \citep{PerezHuerta2007,Samtleben2001,Schmahl2004}, providing an expectation for pristine samples in our dataset.
-The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998,Brand2003,Brand2012,Mii1994,Grossman1996,Shields2003}.
-The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been used by other clumped isotope studies of paleotemperature \citep{Came2007,Dennis2013,Finnegan2011}.
-Measurements of several fossil taxa allow comparison of the preservation quality of different calcite materials.
+The Silurian section on Gotland is also attractive because of the abundance of fossils, especially brachiopods, throughout the section.
+As with previous paleoclimate studies we sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements).
+Many studies of paleoenvironment have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}.
+The fibrous calcite fabric characteristic of the secondary layer has been well characterized in modern and fossil brachiopods \citep{PerezHuerta2007, Samtleben2001, Schmahl2004}, providing a standard of comparison for the samples in our dataset.
+The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998, Brand2003, Brand2012, Mii1994, Grossman1996, Shields2003}.
+The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been used by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}.
+Measurements of several fossil taxa allow comparison of the relative preservation of different calcitic materials and can help to detect taxon-specific ``vital effects'' if they are present.
 
-A third attractive feature of the Silurian section on Gotland is the coupled positive excursion in \deltac C and \deltao O across the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
-This excursion is called the Ireviken event, and is recognized globally \citep{Munnecke2003}.
-In Gotland, the 3$\permil$ excursion in \deltac C and 0.6$\permil$ excursion in \deltao O reach a maximum rate of change across the contact between the Lower and Upper Visby Beds \citep{Munnecke2003}.
-The excursions coincide with a shift to deposition in shallower water, but it is difficult to distinguish the relative contributions of progradation and eustatic sea level fall to this facies change \citep{Calner2004a}.
+A third attractive feature of the Silurian section on Gotland is the coupled positive excursion in $\delta^{13}$C and $\delta^{18}$O that initiates at the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
+This excursion, termed the Ireviken event, is now recognized globally \citep{Munnecke2003}.
+In Gotland, a 3$\permil$ excursion in $\delta^{13}$C and 0.6$\permil$ excursion in $\delta^{18}$O attain a maximum rate of change across the contact between the Lower and Upper Visby Beds \citep{Munnecke2003}.
+The excursions coincide with a maximum flooding surface, but it is difficult to distinguish the relative contributions of progradation and eustatic sea level fall to this facies change \citep{Calner2004a}.
 The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.
@@
 The coincident isotopic excursions and stratigraphic change suggest that the Ireviken represents a climate phenomenon, the nature of which is still debated.
-The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998,Kaljo2003,Brand2006,Calner2008}.
-This hypothesis is supported by the widespread presence of glacial tillites in Brazil, dated to the latest Llandovery or earliest Wenlock \citep{Grahn1992}.
-Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics \citep{Bickert1997,Munnecke2003,Samtleben1996}.
-According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would lower the \deltao O of surface waters and also stimulate the upwelling of deep water with low \deltac C onto the shelves.
-Increased precipitation would also cause intense weathering, which explains the dominance of marls over carbonate reefs during this time.
-Then during the arid period beginning in the Upper Visby Beds, evaporation would increase the \deltao O of surface waters, and downwelling of saline water would bring only surface water with high \deltac C onto the shelves.
-Low terrigenous influxes would allow the growth of carbonate reefs, observed in the H\"{o}gklint Formation.
+The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998, Kaljo2003, Brand2006, Calner2008}.
+This hypothesis is supported by the widespread presence of glacial tillites in Brazil that have been dated to the latest Llandovery or earliest Wenlock \citep{Grahn1992}.
+Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Munnecke2003, Samtleben1996}.
+According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves.
+Increased precipitation would also have caused intense weathering, which could explain the dominance of marls over carbonate reefs during this time.
+During the succeeding arid period beginning in the Upper Visby Beds, evaporation would have increased the $\delta^{18}$O of surface waters, and downwelling of saline water would have brought surface water with high $\delta^{13}$C onto the shelves.
+Low terrigenous influxes would have allowed the growth of carbonate reefs as observed in the H\"{o}gklint Formation.
 
-In the present study, we test the ability of clumped isotopes to resolve the changes in ocean temperature and seawater \deltao O that contributed to the 0.6$\permil$ shift in the \deltao O of carbonates through the Ireviken event.
-From first principles, it will be difficult for the carbonate clumped isotope thermometer to resolve these changes, because a shift of 0.6$\permil$ in \deltao O corresponds to a change in temperature of approximately 2\degrees C, which is near the limit of resolution of carbonate clumped isotopes.
+In the present study, we test the ability of clumped isotopes to resolve the changes in ocean temperature and seawater $\delta^{18}$O that contributed to the 0.6$\permil$ shift in the $\delta^{18}$O of carbonates through the Ireviken event.
+From first principles, it will be challenging but not impossible for the carbonate clumped isotope thermometer to resolve these changes, because a shift of 0.6$\permil$ in $\delta^{18}$O corresponds to a change in temperature of approximately 2$^{\circ}$C.
+Single measurements of carbonate clumped isotope composition can attain a resolution of $\pm$1$^{\circ}$C, while running multiple replicates of a sample can improve the resolution to $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.
 
@@
 For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
-Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}.
+Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}, as well as new bulk carbonate $\delta^{13}$C measurements (Fike et al.
+in prep).
+
+Of the 22 brachiopods chosen for sampling, 20 belonged to the species group \textit{Atrypa reticularis}.
+We targeted this species because it is abundant in the sampled strata, the shell has a thick secondary fibrous layer, and it was used in several previous paleoenvironmental studies in Gotland \citep{Bickert1997, Munnecke2003, Samtleben2001}.
 
@@
 For fossils in more strongly lithified beds, fossil material was accessed by cutting away the rock with a saw or microrotary drill.
-After fossils were separated from the rock, they were sonicated for two hours to remove residual sand, silt and mud from the surface of the fossils.
+After fossils were separated from the rock, they were sonicated for two hours in a Cole Parmer 8891 ultrasonic cleaner to remove residual sand, silt and mud from the surface of the fossils.
 Brachiopods, bryozoans, crinoids, gastropods and ostracods were sampled by flaking the surface of the skeleton or shell using a dental chisel.
@@
 For brachiopods, only the flakes preserving the characteristic fabric of the fibrous secondary layer were chosen for analysis (see Figure \ref{flakes}).
-Flakes were then sonicated briefly, rinsed, dried in a 50\degrees C oven, and powdered using a mortar and pestle.
+Flakes were then sonicated briefly, rinsed, dried in a 50$^{\circ}$C oven, and powdered using a mortar and pestle.
 Only the apical ends of rugose corals were sampled, after removal of any sediment from the surface using an abrasive bit.
@@
 However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite.
-Finally, to assess diagenetic alteration, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
+Finally, to assess possible diagenetic end-members, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
 
@@
 The stage was tilted 70$^{\circ}$.
-The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame and a step size of 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
+The AZtec program was used in 4x4 binning mode, with frame averaging over one frame and a step size of 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
 The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite.
@@
 
-Flakes of the secondary fibrous layer from over ten different brachiopods were also imaged to further inspect the preservation of original fibrous calcite texture.
+Flakes of the secondary fibrous layer from twelve different brachiopods were also imaged to further inspect the preservation of original fibrous calcite texture.
 The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
 The coated flakes were imaged using the secondary electron detector in the same SEM as above, under high vacuum.
-The working distance was $\sim 6$ mm and the electron high tension was 10 kV.
+The working distance was $\sim$6 mm and the electron high tension was 10 kV.
 
@@
 Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite.
-Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
+Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nA, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
 Quantitative spot measurements were made using the JEOL quantitative analysis program.
@@
 Although we attempted to sample only well-preserved skeletal material for the carbonate clumped isotope measurements, the large mass of material required makes it very difficult to avoid including small amounts of other phases.
-The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998,Brand2012,Came2007,Finnegan2011}.
-Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980,Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples.
+The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998, Brand2012, Came2007, Finnegan2011}.
+Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980, Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples.
 Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
@@
 Each sample was measured between one and three times.
-Approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90\degrees C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate.
-The subsequent cleaning of the CO$_2$ followed procedures outlined by \cite{Ghosh2006}.
-The cleaned CO$_2$ was analyzed on a Finnigan MAT253 gas source mass spectrometer configured to collect masses 44 through 49.
-The sample gas was measured in eight separate acquisitions, each of which was bracketed by a measurement of reference CO$_2$ with known \deltac C and \deltao O.
-The \deltac C, \deltao O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample relative to the reference CO$_2$.
-Acquisition-to-acquisition standard deviations were on average 0.03$\permil$ for $\delta_{47}$, 0.04$\permil$ for \deltao O, and 0.02$\permil$ for \deltac C.
+For each measurement approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90$^{\circ}$C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate.
+The subsequent purification of evolved CO$_2$ followed procedures outlined by \cite{Ghosh2006}.
+Cleaned CO$_2$ was analyzed on a Finnigan MAT253 gas source mass spectrometer configured to collect masses 44 through 49.
+The sample gas was measured in eight separate acquisitions, each of which was bracketed by a measurement of reference CO$_2$ with known $\delta^{13}$C and $\delta^{18}$O.
+The $\delta^{13}$C, $\delta^{18}$O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample relative to the reference CO$_2$.
+Acquisition-to-acquisition standard deviations were on average 0.03$\permil$ for $\delta_{47}$, 0.04$\permil$ for $\delta^{18}$O, and 0.02$\permil$ for $\delta^{13}$C.
 Raw measured values for $\Delta_{47}$ and $\Delta_{48}$ of each sample, defined as $\Delta_{47-[\text{SGvsWG}]}$ and $\Delta_{48-[\text{SGvsWG}]}$, were also calculated relative to the reference CO$_2$ following equations 3 and 4 in \cite{Huntington2009}.
@@
 All $\Delta_{47}$ values are projected into the absolute reference frame \citep{Dennis2011}.
-When possible, raw $\Delta_{47-[\text{SGvsWG}]}$ values are converted to the absolute reference frame using an empirical transfer function (ETF) constructed from the heated and equilibrated gases measured during the same week as the samples.
-Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000\degrees C for three hours, and then cooled rapidly to room temperature.
-Equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25\degrees C for at least 24 hours.
-Heated gases were made from two separate reservoirs of CO$_2$ with different \deltao O compositions, and equilibrated gases were equilibrated with either of two separate waters, also differing in their \deltao O compositions.
+$\Delta_{47}$ values of two California Institute of Technology intralab standards were converted to the absolute reference frame using a secondary transfer function (STF) constructed from the heated and equilibrated gases measured during the same week as the standards.
+Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000$^{\circ}$C for three hours, and then cooled rapidly to room temperature.
+Equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25$^{\circ}$C for at least 24 hours.
+Heated gases were made from two separate reservoirs of CO$_2$ with different $\delta^{18}$O compositions, and equilibrated gases were equilibrated with either of two separate waters, also differing in their $\delta^{18}$O compositions.
 Heated and equilibrated gases were cleaned following the same procedure as the samples.
-The ETF was constructed and then used to convert raw $\Delta_{47-[\text{SGvsWG}]}$ values to the absolute reference frame following \cite{Dennis2011}.
-Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90\degrees C by adding 0.092$\permil$ \citep{Henkes2013}.
+The STF was constructed by plotting the accepted $\Delta_{47}$ values of 0.0266$\permil$ for heated gases and 0.9252$\permil$ for equilibrated gases against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame \citep{Dennis2011}.
+The STF was used to convert the $\Delta_{47}$ values of the two intralab standards from the \cite{Ghosh2006} reference frame to the absolute reference frame following equation C1 in \cite{Dennis2011}.
 
-Some samples were measured during weeks when no equilibrated gases were measured, so an ETF could not be constructed for those weeks.
-These samples were projected into the absolute reference frame using a secondary transfer function (STF).
-Accepted values of two California Institute of Technology intralab standards were determined in the absolute reference frame from measurements of these standards during weeks when an ETF could be constructed from heated and equilibrated gas measurements.
-To create the STF, the accepted values for the two standards, as well as the accepted value of 0.0266$\permil$ for heated gases, were plotted against their $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
+All of the samples in this study were projected into the absolute reference frame using a separate STF constructed for each week of measurements.
+To create the STF, the accepted $\Delta_{47}$ values for the two intralab standards and heated gases were plotted against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
+Equilibrated gases were measured during some but not all weeks of sample measurements, so the accepted and observed $\Delta_{47}$ values of equilibrated gases were included in the weekly STF during the weeks when these measurements were available.
 $\Delta_{47}$ values of the samples in the \cite{Ghosh2006} reference frame were corrected using this STF following equation C1 in \cite{Dennis2011}.
-A separate STF was constructed for each week of measurements during which no equilibrated gases were measured.
+Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90$^{\circ}$C by adding 0.092$\permil$ \citep{Henkes2013}.
 
 Although all samples were cleaned prior to introduction into the mass spectrometer, the possibility for contamination from hydrocarbons and chlorocarbons still exists.
-To screen for contamination, the $\delta_{48}$ and $\Delta_{48}$ of each sample were compared to a line of best fit through a plot of $\delta_{48}$ versus $\Delta_{48}$ of the heated gases, all in the \cite{Ghosh2006} reference frame.
+To screen for contamination, the $\delta_{48}$ and $\Delta_{48}$ of each sample were compared to a line of best fit through a plot of $\delta_{48}$ versus $\Delta_{48}$ of the heated gases run during the same week, all in the \cite{Ghosh2006} reference frame.
 Samples whose $\Delta_{48}$ fell more than 1$\permil$ off the line were discarded due to likely contamination.
@@
 $\Delta_{47-\text{RF-AC}}$ values were converted to temperature using the \cite{Ghosh2006} calibration line, which is projected into the absolute reference frame in equation 9 in \cite{Dennis2011}.
-The \deltao O of water was calculated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.
+The $\delta^{18}$O of water was calculated from the temperature and the measured $\delta^{18}$O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03(10^3\text{T}^{-1})-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.
 
@@
 
-The \deltac C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}).
-We resolve a 3$\permil$ positive excursion in \deltac C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters.
-The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation, beginning at 76 meters, record a return to lower \deltac C$_{\text{PDB}}$ values of approximately -0.5$\permil$.
-The \deltac C trend through the section is comparable to the brachiopod \deltac C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
+The $\delta^{13}$C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}).
+We resolve a 3$\permil$ positive excursion in $\delta^{13}$C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters.
+The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation record a return to lower $\delta^{13}$C$_{\text{VPDB}}$ values of approximately -0.5$\permil$.
+The $\delta^{13}$C trend through the section is comparable to the brachiopod $\delta^{13}$C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
 
-\begin{figure*}[b]
+\begin{figure*}[t]
 \centering
-\includegraphics[width=\textwidth]{Fig_03_color-01.png}
-\caption{The \deltac C, \deltao O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
+\includegraphics[width=\textwidth]{Fig_03_bw-01.png}
+\caption{The $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
 For the Lower \& Upper Visby and H\"{o}gklint Formations, a height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds.
@@
 
-The \deltao O values of fossil material are consistent within beds, but micrite and cements tend to have lower \deltao O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
-We are not able to resolve an excursion in \deltao O, although this may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}.
-The average \deltao O$_{\text{PDB}}$ of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples consistent with those of \cite{Munnecke2003} but our Upper Visby samples more depleted in $^{18}$O.
-Higher in the section, the \deltao O of the fossils in this study appear consistent with values reported by \cite{Bickert1997}.
+The $\delta^{18}$O values of fossil material are consistent within beds, but micrite and cements tend to have lower $\delta^{18}$O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
+We are not able to resolve an excursion in $\delta^{18}$O.
+This may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}, or because our samples include more contamination from secondary calcite phases.
+The average $\delta^{18}$O$_{\text{VPDB}}$ of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples consistent with those of \cite{Munnecke2003}, but our Upper Visby samples more depleted in $^{18}$O.
+The low $\delta^{18}$O values of Upper Visby fossils may be caused by increased contamination from secondary cements, which could be expected to accompany the transition from marl to limestone.
+Higher in the section, the $\delta^{18}$O of the fossils in this study appear consistent with values reported by \cite{Bickert1997}.
 
-The clumped isotope temperature of samples is highly variable within beds, often covering a range of 10\degrees C within a single bed (Figure \ref{section}).
-The lowest temperature sample, a brachiopod, and the highest temperature sample, a rugose coral, are both from the Slite Formation, deposited well after the hypothesized climate phenomenon associated with the isotopic excursions in the Lower and Upper Visby Beds.
-Due to the scatter in temperature, we are not able to resolve a change in temperature with time.
+Clumped isotope temperatures of samples are highly variable within beds, often covering a range of 10$^{\circ}$C within a single bed (Figure \ref{section}).
+The lowest temperature sample, a brachiopod, is from the Slite Formation, deposited well after the hypothesized climate changes associated with the isotopic excursions in the Lower and Upper Visby Beds.
+The large scatter in clumped isotope temperatures suggests variable preservation quality among these samples.
+Even brachiopods, which are commonly thought to be relatively resistant to diagenetic alteration, have a range in temperature of more than 15$^{\circ}$C.
+The variable preservation quality of these samples means that beds with fewer samples are less likely to contain a pristine fossil with a reliable temperature.
+Because many of the beds in this study are only represented by a few samples, we are not able to stratigraphically resolve a change in temperature with time.
 
-The large scatter in temperatures suggests variable preservation quality among the samples.
-Even brachiopods, which are commonly thought to be resistant to diagenetic alteration, have a range in temperature of more than 15\degrees C.
-All of the fossil samples have fairly high values of \deltao O, making alteration by meteoric water unlikely.
-This leads to a hypothesis that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature.
-We tested this hypothesis using independent methods to assess the preservation quality of representative samples.
+Although we are not able to resolve a change in temperature, we can determine a representative temperature for these sections by using constraints on the mechanism of diagenetic alteration to identify the most pristine samples.
+All of the fossil samples have fairly high values of $\delta^{18}$O, making alteration by meteoric water unlikely.
+This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature.
+These higher temperatures could be caused by incorporation of late-stage cements or wholesale recrystallization.
+We tested this hypothesis using independent methods, described above, to assess the preservation quality of representative samples.
 
@@
 
-Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration, suggest that all the brachiopods in this study are pristine.
-The bulk Mn and Sr concentrations of all brachiopods fall within previously defined limits for the Mn and Sr concentrations of pristine brachiopods (Figure \ref{Brand}).
+\begin{figure*}[b]
+\centering
+\includegraphics[width=\textwidth]{Fig_05_bw-01.png}
+\captionsetup{labelformat=empty} 
+\caption{Figure 5: Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
+The negative correlation between temperature and the log of Sr concentration in rugose corals is weakly significant (79\% confidence), and strengthens when combined with the rest of the samples.
+There are also weakly significant positive correlations between temperature and the log of Fe concentration (63\% confidence) and between temperature and the log of Mn concentration (83\% confidence) among all the samples combined.
+These plots demonstrate clear differences in trace metal content between the two fossil taxa and the later carbonate phases.
+Brachiopods have the lowest average temperature of all the sample subsets, and also tend to have the most pristine trace metal signatures, with less Fe and Mn and more Sr than the other samples.
+Micrites and cements have the highest average temperature and tend to have the most altered trace metal signatures, with high Fe and Mn and low Sr concentrations.}
+\label{Brand}
+\end{figure*}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=90mm]{Fig_04_bw-01.png}
+\captionsetup{labelformat=empty} 
+\caption{Figure 4: Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
+The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by \cite{Brand2012} which are also shown in this figure.
+Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods.
+Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.
+Figure modified from \cite{Brand2012}.}
+\label{metals}
+\end{figure}
+
+Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration, suggest that all the brachiopods in this study are comparatively pristine.
 However, micro-scale studies of calcite fabrics, as well as clumped isotope measurements, demonstrate that there is a range in preservation quality among the brachiopods that cannot be identified on the basis of trace metal measurements alone.
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=90mm]{Fig_04_color-01.png}
-\caption{Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
-The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by \cite{Brand2012} which are also shown in this figure.
-Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods.
-Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.
-Figure modified from \cite{Brand2012}.}
-\label{Brand}
-\end{figure}
-
-There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration.
-Our sample set may not include a sufficient number of highly altered samples to capture the full range in severity of alteration.
-This hypothesis is supported by a conspicuous lack of statistically robust correlations between trace metal composition and temperature (Figure \ref{metals}).
-We tested brachiopods and rugose corals separately and combined with micrites and cements for the robustness of a linear correlation between trace metal concentrations and temperature, as well as the log of trace metal concentrations and temperature.
-We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero.
-Only one pair of variables, the Sr concentration and temperature of rugose corals, has a p value of less than 0.05, meaning we can be 95\% confident that the correlation is real.
-While these tests suggest that the scatter in trace metal concentrations is a large cause of their insensitivity to diagenetic alteration, there is also the possibility that early void-filling cements, which would have cool temperatures but altered trace metal concentrations, interfere with the signal.
+Clumped isotope thermometry reveals variable preservation among the brachiopods in this study, so we expected to observe equivalent variations in trace metal content.
+Diagenetic recrystallization of brachiopod skeletal calcite is thought to enrich Fe and Mn because these metals are insoluble in oxic seawater but often found at elevated concentrations in anoxic basinal fluids.
+Diagenetic recrystallization is also thought to deplete Sr because brachiopod skeletons concentrate Sr relative to its abundance in seawater and meteoric waters \citep{Brand1980, Shields2003}.
+Many previous studies have measured the concentrations of these three elements in fossil brachiopods, and some have defined limits for the Mn and Sr concentrations of brachiopods that can be considered pristine.
+The bulk Mn and Sr concentrations of all brachiopods in this study fall within these limits (Figure \ref{Brand}); the small variations in trace metal concentrations among the brachiopods would not ordinarily be considered as evidence for variable preservation.
 
@@
 \centering
-\includegraphics[width=\textwidth]{Fig_05_color-01.png}
-\caption{Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
-The negative correlation between temperature and the log of Sr concentration in rugose corals is statistically robust, and holds up when combined with the rest of the samples, but no other correlations passed our hypothesis tests.
-These plots demonstrate clear differences in trace metal content between the two fossil taxa and the later carbonate phases.
-Brachiopods tend to have the most pristine trace metal signatures, with less Mn and more Sr than the other samples.
-Micrites and cements tend to have the most altered trace metal signatures, with high Fe and Mn and low Sr concentrations.
-}
-\label{metals}
-\end{figure*}
-
-E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization may not actually have any effect on Fe and Mn concentrations.
-A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B).
-However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D).
-This is only one example, but it suggests that recrystallization may not affect trace metal concentrations, and instead, variations in trace metal concentrations may be dominated by contamination from cements and sediments.
-If recrystallization is an important contributor to the elevated temperatures of fossil material, trace metals may not be expected to be reliable indicators of diagenetic alteration.
-
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=140mm]{Fig_06_color-01.png}
+\includegraphics[width=140mm]{Fig_06_bw-01.png}
 \caption{Calcite fabric and trace metal concentrations in a rugose coral.
@@
 A) Transmitted light.
-The dark triangle labelled ``RC'' was identified as a recrystallized domain of the coral septum.
+The dark triangle labeled ``RC'' was identified as a recrystallized domain of the coral septum.
 Cement in the void space between coral septa appears as blocky calcite crystals.
@@
 
+Given our hypothesis that the lowest temperature samples are the most pristine, we also expected to observe a correlation between clumped isotope temperatures and trace metal evidence for diagenetic alteration.
+We tested brachiopods, rugose corals, and micrites/cements separately and combined for the robustness of a linear correlation between temperature and the log of trace metal concentrations (Figure \ref{metals}).
+We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero.
+We used a Bonferroni correction to adjust the significance level ($\alpha$) to account for the multiple hypothesis tests.
+There are weakly significant positive correlations between temperature and Fe concentration (63\% confidence) and also between temperature and Mn concentration (83\% confidence) when all samples are combined.
+These correlations are likely caused by differences between fossil taxa and later phases; brachiopods have the lowest average temperature, Fe, and Mn concentrations, while micrites and cements combined have the highest average temperature, Fe and Mn concentrations.
+There is also a weakly significant (79\% confidence) negative correlation between temperature and Sr concentration among rugose corals, which strengthens (to 99\% confidence) when corals are combined with brachiopods, and with micrites and cements.
+However, none of the other variable pairs are significantly correlated within individual taxa (Supplementary table 2).
+These tests may be complicated by contamination from early void-filling cements, which we expect to have low temperatures but altered trace metal signatures.
+Variability in Mn concentrations of low-temperature samples may obscure a positive correlation between temperature and Mn concentration, but low-temperature contamination does not appear to explain the lack of correlation between temperature and Fe or Sr.
+We suggest that the most likely explanation for these observations is that trace metal concentrations, with the exception of Sr in rugose corals, are relatively insensitive indicators of diagenetic alteration within fossil taxa, and are only sensitive enough to reflect broad differences between taxa.
+
+\begin{figure*}[tb]
+\centering
+\includegraphics[width=\textwidth]{Fig_07_bw-01.png}
+\caption{Calcite fabric in a brachiopod.
+A) Transmitted polarized light image of a thin section of brachiopod G3 LVF B1.
+This sample has a clumped isotope temperature of 35$^{\circ}$C.
+The boxed overlay is an electron backscatter diffraction (EBSD) image of the same sample.
+The orientation of calcite crystals is indicated by different shades.
+B) Euler pole figure generated from the EBSD analysis.
+The central band represents the preferential orientation of calcite crystals parallel to the outer surface of the brachiopod shell.}
+\label{EBSD_brach}
+\end{figure*}
+
+There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration.
+Our sample set likely does not include a sufficient number of highly altered samples to capture the full range in severity of alteration, because visibly altered fossils were rejected during visual screening of samples and not included in the sample set.
+Variability in primary trace metal concentrations, which is large among modern brachiopods \citep{Brand2003}, may also mask any trends superimposed by diagenetic alteration.
+However, E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization need not necessarily have any effect on Fe and Mn concentrations.
+A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B).
+However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D).
+This is only one example, but it suggests that recrystallization may not affect trace metal concentrations, and instead, variations in trace metal concentrations may be dominated by contamination from cements and sediments.
+The trace metal signature associated with recrystallization likely depends on conditions during recrystallization, suggesting that this coral underwent closed-system diagenesis.
+If closed-system recrystallization is an important contributor to the elevated temperatures of fossil material in our sample set, trace metals may not be expected to be reliable indicators of diagenetic alteration.
+
 We therefore turn to micro-scale studies of calcite textures as indicators of diagenetic alteration.
-Reflected-light and EBSD images of rugose corals reveal substantial amounts of late-phase diagenetic calcite, including cements and recrystallized regions (Figure \ref{EBSD_coral}A \& B).
+Reflected-light and EBSD images of rugose corals reveal substantial amounts of late-phase diagenetic calcite, including recrystallized regions and cements (Figure \ref{EBSD_coral}A \& B).
 As discussed above, recrystallized regions are visible as dark colors in reflected light and aligned calcite crystals in EBSD images.
@@
 The fraction of late-phase calcite in the rugose coral samples may have been lower than suggested by the thin section in Figure \ref{EBSD_coral}, because samples were taken from the apical end of the corals where the void space between the septa was smaller, but sample preparation probably included some late-phase calcite along with pristine coral skeletal material.
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=90mm]{Fig_08-01.png}
+\caption{Scanning electron microscope images of flakes of brachiopod shell.
+Images were taken in secondary electron mode.
+Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
+Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
+Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2$^{\circ}$C; B) 33 $\pm$ 2$^{\circ}$C; C) 35 $\pm$ 1$^{\circ}$C.}
+\label{flakes}
+\end{figure}
 
@@
 
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=\textwidth]{Fig_07_color-01.png}
-\caption{Calcite fabric in a brachiopod.
-A) Transmitted polarized light image of a thin section of brachiopod G3 LVF B1.
-This sample has a clumped isotope temperature of 35\degrees C.
-The boxed overlay is an electron backscatter diffraction (EBSD) image of the same sample.
-The orientation of calcite crystals is indicated by different shades.
-B) Euler pole figure generated from the EBSD analysis.
-The central band represents the preferential orientation of calcite crystals parallel to the outer surface of the brachiopod shell.}
-\label{EBSD_brach}
-\end{figure*}
-
 Scanning electron microscope (SEM) images of low-temperature brachiopod flakes also demonstrate excellent preservation of original fibrous calcite textures (Figure \ref{flakes}A, B \& C).
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=90mm]{Fig_08-01.png}
-\caption{Scanning electron microscope images of flakes of brachiopod shell.
-Images were taken in secondary electron mode.
-Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
-Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
-Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2\degrees C; B) 33 $\pm$ 2\degrees C; C) 35 $\pm$ 1\degrees C.}
-\label{flakes}
-\end{figure}
-
 The micro-scale studies of calcite textures in Figures \ref{EBSD_coral}--\ref{flakes} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite.
@@
 An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite, likely less than 5\% of the sample.
-We therefore look to these samples as the most reliable records of ocean temperature and \deltao O during the Silurian.
+We therefore look to these samples as the most reliable records of ocean temperature and $\delta^{18}$O during the Silurian.
 
@@
 
-To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore preserve the most reliable record of Silurian ocean conditions.
+\begin{figure}[b]
+\centering
+\includegraphics[width=90mm]{Fig_09-01.png}
+\caption{The clumped isotope temperature and calculated water $\delta^{18}$O$_{\text{VSMOW}}$ for all samples analyzed in this study.
+Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.
+The positive correlation between temperature and water $\delta^{18}$O$_{\text{VSMOW}}$ is diagnostic of rock-buffered diagenesis.
+Under this regime, the lowest-temperature samples are the most pristine.}
+\label{ellipses}
+\end{figure}
+
+To estimate the temperature and $\delta^{18}$O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore preserve the most reliable record of Silurian ocean conditions.
 This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected.
-As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the \deltao O of interstitial water.
-When the clumped isotope temperatures of the samples in this study are plotted versus the \deltao O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}).
+As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the $\delta^{18}$O of interstitial water.
+When the clumped isotope temperatures of the samples in this study are plotted versus the $\delta^{18}$O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}).
 This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is diagnostic of rock-buffered diagenesis.
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=90mm]{Fig_09-01.png}
-\caption{The clumped isotope temperature and calculated water \deltao O$_{\text{VSMOW}}$ for all samples analyzed in this study.
-Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.}
-\label{ellipses}
-\end{figure}
-
-Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis, but we rule it out as a major contributor to our results.
-During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite.
-This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and reprecipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
-However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the observed variability in temperature among the samples.
-Additionally, solid-state diffusion is thought to occur very slowly at temperatures below 100\degrees C (D.
-Stolper personal communication 2013), which is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983,Wenzel2000}.
-Solid-state diffusion is not consistent with our observations, leaving rock-buffered diagenesis as the most likely characterization of the diagenetic regime.
-
-The hypothesis that the samples underwent rock-buffered diagenesis is further supported by comparisons of the clumped isotope temperature of different calcite textures.
-We measured the clumped isotope composition of the micrite adjecent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod (Figure \ref{comparisons}).
-Micrite is expected to be less pristine than the brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis.
-Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
-
-\begin{figure}[h]
+\begin{figure}[b]
 \centering
 \includegraphics[width=90mm]{Fig_10-01.png}
-\caption{Comparisons between the \deltac C, \deltao O and clumped isotope temperature of brachiopod samples and adjacent micrite.}
+\caption{Comparisons between the $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of brachiopod samples and adjacent micrite.
+Micrite is expected to be more easily altered than brachiopod shell material, so the consistently higher temperature of micrite supports our hypothesis that the lowest temperature samples are the most pristine.}
 \label{comparisons}
@@
 
-The lowest-temperature, most pristine sample in this study, a brachiopod, indicates that tropical Silurian oceans had a temperature of $30\pm 2$\degrees C and a \deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$ (Figure \ref{ellipses}), where the error estimates represent the standard error of the mean over eight acquisitions.
-The two next-most pristine samples, also brachiopods, have temperatures of $33\pm 2$\degrees C and $35\pm 2$\degrees C, which correspond to seawater \deltao O$_{\text{SMOW}}$ values of $-1.5\pm 0.3\permil$ and $-1.9\pm 0.4\permil$, respectively.
+Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis \citep{Dennis2013, Huntington2011, Quade2013, VanDeVelde2013}, and we cannot rule it out as a contributor to our results.
+During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite.
+This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and re-precipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
+Calcites with different distributions and concentrations of trace elements have been demonstrated to have different rates of solid-state diffusion \citep{Passey2012}, which means that this process could potentially explain the observed variability in temperature among the samples.
+Temperatures as low as 100$^{\circ}$C may allow solid-state diffusion to increase the clumped isotope temperature of a calcite, provided that the calcite experiences these temperatures for millions of years \citep{Dennis2010}.
+100$^{\circ}$C is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983, Wenzel2000}, and it is difficult to estimate the length of time for which Gotland strata were exposed to elevated temperature.
+However, the effects of solid-state diffusion are consistent with our hypothesis that the lowest temperature samples are the most pristine, because solid-state diffusion increases the clumped isotope temperature of calcite.
+
+Additional evidence for a diagenetic regime that acted to increase the temperature of more altered samples is provided by comparisons of the clumped isotope temperature of different calcite textures (Figure \ref{comparisons}).
+We measured the clumped isotope composition of micrites adjacent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod.
+Micrite is expected to be less pristine than the brachiopod samples because micrite was cemented post-burial, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis.
+Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
+
+The lowest-temperature sample of the apparently pristine best preserved samples we studied, a brachiopod (G13 10.5 to 10.8 B1, Slite Fm.), suggests that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$ (Figure \ref{ellipses}), where the error estimates represent the one-sigma standard error of the mean over all acquisitions.
+The two next-most pristine samples, also brachiopods (G11 0.2 to 0.4 B6, stratigraphic height 1.3 m; G2 0.5 to 0.6 B2, stratigraphic height 0.5 m), have temperatures of 33$\pm$2$^{\circ}$C and 34$\pm$1$^{\circ}$C.
+Together with the $\delta^{18}$O$_{\text{VPDB}}$ of each brachiopod, these temperatures can be used to calculate seawater $\delta^{18}$O$_{\text{VSMOW}}$ values of -2.2$\pm$0.4$\permil$ and -1.2$\pm$0.2$\permil$, respectively.
 One caveat of these values is that they are each based on only one measurement due to sample-size limitations.
-The lowest-temperature brachiopod was picked a second time, but to collect enough material for analysis, some less pristine flakes had to be included, which may explain the $41\pm 4$\degrees C temperature of this re-sample (G13 10.5 to 10.8 B1 leftovers).
-However, we argue that the differences between the temperatures and \deltao O values implied by the three most pristine samples are small enough that they essentially represent the same warm, tropical Silurian ocean water.
+The lowest-temperature brachiopod was picked a second time, but to collect enough material for analysis, some less pristine flakes had to be included, which may explain the 41$\pm$4$^{\circ}$C temperature of this re-sample (G13 10.5 to 10.8 B1 leftovers).
+However, we argue that the differences between the temperatures and $\delta^{18}$O values implied by the three most pristine samples are small enough that they essentially represent the same warm, tropical Silurian ocean water.
 
-A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates.
-We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples.
+A second caveat of our estimates of the temperature and $\delta^{18}$O of Silurian tropical oceans is that they are technically upper estimates.
+We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature recorded by the samples.
 We are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine samples.
 However, micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, comprising a maximum of 5\% of the material.
-Assuming linear mixing, which is not technically correct but close enough for our purposes, if 5\% of the sample material was recrystallized at 100\degrees C and the rest of the sample material retained an original temperature of 30\degrees C, the measured temperature of the sample would be $\sim34$\degrees C.
-Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.
+Assuming closed-system diagenesis so that the late-phase calcite is similar in bulk $\delta^{13}$C and $\delta^{18}$O to the pristine fossil calcite, $\Delta_{47}$ values will mix linearly \citep{Affek2006}.
+In this case, if the measured temperature of the sample is 31$^{\circ}$C and 5\% of the sample material was recrystallized at 100$^{\circ}$C, the primary temperature of the pristine fossil material would be 28.3$^{\circ}$C.
+Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 3$^{\circ}$C, provided that solid-state diffusion has not significantly altered the temperature of the most pristine samples in our dataset.
 
-Our clumped isotope measurements indicate that Silurian tropical oceans were similar to modern tropical oceans in their temperature and \deltao O.
-Modern measurements of sea surface temperature in the Solomon Sea, east of Papua New Guinea, are 30\degrees C \citep{Reynolds1994}, matching our estimate of the temperature at a similar latitude during the Silurian.
-The average \deltao O$_{\text{SMOW}}$ of the modern ocean is approximately $-0.3\permil$ \citep{Shackleton1974}, but without ice, the average \deltao O$_{\text{SMOW}}$ would decrease to $-1.4\permil$ \citep{Lhomme2005}.
-The volume of continental glaciers is poorly constrained for the Early Silurian, but regardless, the \deltao O of Silurian oceans was likely within a few permil of the modern value.
-
-The results of this study are also consistent with those of earlier clumped isotope studies.
-\cite{Came2007} and \cite{Finnegan2011} both measured the clumped isotope compositions of brachiopods from the Early Silurian on Anticosti Island, Canada.
-These two studies report temperatures of 30 to 35\degrees C and seawater \deltao O$_{\text{SMOW}}$ values of around $-1\permil$, which are similar to the results of this study.
-Importantly, Anticosti Island recorded deposition on the eastern margin of the Panthalassic Ocean during the Silurian, while Gotland recorded deposition in the Rheic Ocean, which was separated from the Panthalassic Ocean by a mountain range.
-The strong similarity in temperatures and \deltao O between the two studies indicates that both studies likely reflect water that is representative of the global ocean, rather than an isolated inland sea.
-
-Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates (Figure \ref{Veizer}).
-Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans.
-Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans.
-The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature.
-Clumped isotopes additionally demonstrate that extremely hot ($>50$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates.
-Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality.
-This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.
-
-Due in large part to diagenetic alteration, our results are unable to resolve a change in temperature or the \deltao O of the oceans through the Ireviken event.
-The consistent scatter in fossil temperatures throughout the section likely rules out any large changes in temperature or \deltao O of the oceans, but this scatter prevents resolution of any smaller changes.
-Temperatures of 30\degrees C in tropical oceans are similar to those of modern tropical oceans, and therefore are not incompatible with the presence of polar ice sheets, as suggested by Early Silurian age tillites in Brazil \citep{Grahn1992}.
-
-The results from this study join those of previous clumped isotope studies in implying that the \deltao O$_{\text{SMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around $-1\permil$ (Figure \ref{Veizer}).
-Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the \deltao O of seawater \citep{Muehlenbachs1976,Gregory1981,Gregory1991,Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the \deltao O of seawater \citep{Jaffres2007}.
-If this buffer cannot explain the relative constancy of seawater \deltao O values through time, then the two competing controls on seawater \deltao O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years.
-This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.
-
-\begin{figure}[h]
+\begin{figure}[tb]
 \centering
@@
 \caption{Implications for the constancy of seawater $\delta^{18}$O through time.
-In gray, the \deltao O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}.
-The black solid line represents the estimated \deltao O$_{\text{VSMOW}}$ of the oceans through time based on the calcite values.
-The two black dashed lines represent alternative hypotheses for the value of seawater \deltao O$_{\text{VSMOW}}$ through time.
+In gray, the $\delta^{18}$O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}.
+The black solid line represents the estimated $\delta^{18}$O$_{\text{VSMOW}}$ of the oceans through time based on the calcite values.
+The two black dashed lines represent alternative hypotheses for the value of seawater $\delta^{18}$O$_{\text{VSMOW}}$ through time.
 All three lines ignore short-term perturbations to seawater $\delta^{18}$O caused by glaciations.
-Finally, the black shapes are estimates of seawater \deltao O from clumped isotopes, measured at four different times by three separate studies.
-These clumped isotope estimates suggest that seawater \deltao O has not changed by more than 2$\permil$ in the last 430 million years.}
+Finally, the black shapes are estimates of seawater $\delta^{18}$O from clumped isotopes, measured at four different times by three separate studies.
+These clumped isotope estimates suggest that seawater $\delta^{18}$O has not changed by more than 2$\permil$ in the last 430 million years.}
 \label{Veizer}
@@
 
+Our clumped isotope measurements indicate that Silurian tropical oceans were broadly similar to modern tropical oceans in their temperature and $\delta^{18}$O.
+Our estimate of the temperature at a Silurian paleolatitude of 19$^{\circ}$ south \citep{Torsvik1992} falls within the range of observed temperatures for the modern ocean at a similar latitude \citep{Reynolds1994}.
+The average $\delta^{18}$O$_{\text{VSMOW}}$ of the modern ocean is approximately -0.3$\permil$ \citep{Shackleton1974}, but without ice, the average $\delta^{18}$O$_{\text{VSMOW}}$ would decrease to -1.4$\permil$ \citep{Lhomme2005}.
+The volume of continental glaciers is poorly constrained for the early Silurian, but assuming Silurian ice volume was less than or equal to modern ice volume, our findings suggest that the $\delta^{18}$O of Silurian oceans was within 2$\permil$ of the modern value.
+
+The results of this study are also consistent with those of an earlier clumped isotope study.
+\cite{Came2007} measured the clumped isotope compositions of brachiopods from the early Silurian on Anticosti Island, Canada, and report temperatures of 30 to 35$^{\circ}$C and seawater $\delta^{18}$O$_{\text{VSMOW}}$ values of around -1$\permil$, which are similar to the results of this study.
+Importantly, Anticosti Island recorded deposition on the eastern margin of the Panthalassic Ocean during the Silurian, while Gotland recorded deposition in the Rheic Ocean, which was separated from the Panthalassic Ocean by a mountain range.
+The strong similarity in temperatures and $\delta^{18}$O between the two studies indicates that both studies likely reflect water that is representative of the global ocean, rather than an isolated inland sea.
+
+Our results rule out large ($>2\permil$) changes in the $\delta^{18}$O of seawater and extremely hot ($>50^{\circ}$C) ocean temperatures as causes of the depleted $\delta^{18}$O values of Silurian carbonates.
+Instead, clumped isotopes demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality.
+This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration must be a key contributor to the low $\delta^{18}$O values of Silurian carbonates.
+
+The results from this study join those of previous clumped isotope studies \citep{Came2007, Dennis2013, Finnegan2011} in implying that the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around -1$\pm$1$\permil$ (Figure \ref{Veizer}).
+Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the $\delta^{18}$O of seawater \citep{Muehlenbachs1976, Gregory1981, Gregory1991, Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the $\delta^{18}$O of seawater \citep{Jaffres2007}.
+If this buffer cannot explain the relative constancy of seawater $\delta^{18}$O values through time, then the two competing controls on seawater $\delta^{18}$O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years.
+This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.
+
 \section{Conclusions}
 
-We used clumped isotopes to constrain the temperature and \deltao O of Silurian oceans because they provide a temperature that is independent of the \deltao O of the water from which the carbonate precipitated.
+We used clumped isotopes to constrain both the temperature and $\delta^{18}$O of Silurian oceans because they provide a temperature that is independent of the $\delta^{18}$O of the water from which the carbonate precipitated.
 Samples were collected from relatively well-preserved sections on the island of Gotland, Sweden.
-By sampling well-preserved, fossil-rich rocks that record a climate phenomenon thought to be associated with changing seawater temperatures, we hoped to test the limit of resolution of clumped isotopes.
+By sampling well-preserved, fossil-rich rocks that potentially record a climate phenomenon associated with changing seawater temperatures, we hoped to test the limit of resolution of clumped isotopes.
 We were not able to resolve a change in temperature through the section, due to the variable preservation quality of the samples.
 Trace metal analysis was not able to detect variations in preservation among brachiopods, despite a range in brachiopod temperatures, suggesting that clumped isotopes are a more sensitive indicator of diagenetic alteration than trace metal compositions.
-Micro-scale analysis of calcite fabrics, comparisons between brachiopods and micrite, and the positive relationship between temperature and \deltao O of water calculated from the samples all support the hypothesis that the lowest-temperature samples are the most pristine.
-Based on the most pristine sample in our dataset, we estimate that tropical Silurian oceans had a temperature of  $30\pm 2$\degrees C and a \deltao O of $-1.7\pm 0.5\permil$.
-Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the \deltao O of seawater has remained broadly constant over time, in spite of a large potential for change through exchange of oxygen isotopes with crustal materials.
+Micro-scale analysis of calcite fabrics, comparisons between brachiopods and micrite, and the positive relationship between temperature and $\delta^{18}$O of water calculated from the samples all support the hypothesis that the lowest-temperature samples are the most pristine.
+Based on the most pristine sample in our dataset, we estimate that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$.
+Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the $\delta^{18}$O of seawater has remained broadly constant over time, in spite of a large potential for change through exchange of oxygen isotopes with crustal materials.
 
@@
 
-We acknowledge Nami Kitchen, Kristin Bergmann, Daniel Stolper, and the rest of the Eiler lab for help with the clumped isotope measurements.
+We acknowledge Nami Kitchen, Kristin Bergmann, Daniel Stolper, and the rest of the Eiler lab for help with the clumped isotope measurements and J.
+Garrecht Metzger for assistance in field collection.
 We thank Chi Ma for his assistance with the SEM and the E-probe, and Joel Hurowitz for allowing us to use his ICP-AES.

### 2013-08-19 draft.tex
moved flakes figure up to methods section and changed to color versions of figures
changed figure numbers to reflect new order, figure positions not perfect

@@
 
+\begin{figure}[tb]
+\centering
+\includegraphics[width=90mm]{Fig_03-01.png}
+\caption{Scanning electron microscope images of flakes of brachiopod shell.
+Images were taken in secondary electron mode.
+Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
+Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
+Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2$^{\circ}$C; B) 33 $\pm$ 2$^{\circ}$C; C) 35 $\pm$ 1$^{\circ}$C.}
+\label{flakes}
+\end{figure}
+
 While all fossils were inspected under a microscope prior to sampling, we were not able to inspect all fossils in thin section, because carbonate clumped isotope measurements require large amounts of material, some of which is lost during preparation of thin sections.
@@
 
+\begin{figure*}[t]
+\centering
+\includegraphics[width=\textwidth]{Fig_04_color-01.png}
+\caption{The $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
+For the Lower \& Upper Visby and H\"{o}gklint Formations, a height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds.
+For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop.
+Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included.
+Error bars represent the standard error of the mean.}
+\label{section}
+\end{figure*}
+
 \section{Results and discussion}
@@
 The $\delta^{13}$C trend through the section is comparable to the brachiopod $\delta^{13}$C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
-
-\begin{figure*}[t]
-\centering
-\includegraphics[width=\textwidth]{Fig_03_bw-01.png}
-\caption{The $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
-For the Lower \& Upper Visby and H\"{o}gklint Formations, a height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds.
-For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop.
-Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included.
-Error bars represent the standard error of the mean.}
-\label{section}
-\end{figure*}
 
@@
 
-Although we are not able to resolve a change in temperature, we can determine a representative temperature for these sections by using constraints on the mechanism of diagenetic alteration to identify the most pristine samples.
-All of the fossil samples have fairly high values of $\delta^{18}$O, making alteration by meteoric water unlikely.
-This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature.
-These higher temperatures could be caused by incorporation of late-stage cements or wholesale recrystallization.
-We tested this hypothesis using independent methods, described above, to assess the preservation quality of representative samples.
-
-\subsection{Independent assessments of preservation quality}
-
 \begin{figure*}[b]
 \centering
-\includegraphics[width=\textwidth]{Fig_05_bw-01.png}
+\includegraphics[width=\textwidth]{Fig_06_color-01.png}
 \captionsetup{labelformat=empty} 
-\caption{Figure 5: Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
+\caption{Figure 6: Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
 The negative correlation between temperature and the log of Sr concentration in rugose corals is weakly significant (79\% confidence), and strengthens when combined with the rest of the samples.
@@
 \centering
-\includegraphics[width=90mm]{Fig_04_bw-01.png}
+\includegraphics[width=90mm]{Fig_05_color-01.png}
 \captionsetup{labelformat=empty} 
-\caption{Figure 4: Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
+\caption{Figure 5: Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
 The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by \cite{Brand2012} which are also shown in this figure.
@@
 \end{figure}
+
+Although we are not able to resolve a change in temperature, we can determine a representative temperature for these sections by using constraints on the mechanism of diagenetic alteration to identify the most pristine samples.
+All of the fossil samples have fairly high values of $\delta^{18}$O, making alteration by meteoric water unlikely.
+This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature.
+These higher temperatures could be caused by incorporation of late-stage cements or wholesale recrystallization.
+We tested this hypothesis using independent methods, described above, to assess the preservation quality of representative samples.
+
+\subsection{Independent assessments of preservation quality}
 
@@
 \centering
-\includegraphics[width=140mm]{Fig_06_bw-01.png}
+\includegraphics[width=140mm]{Fig_07_color-01.png}
 \caption{Calcite fabric and trace metal concentrations in a rugose coral.
@@
 \centering
-\includegraphics[width=\textwidth]{Fig_07_bw-01.png}
+\includegraphics[width=\textwidth]{Fig_08_color-01.png}
 \caption{Calcite fabric in a brachiopod.
@@
 The fraction of late-phase calcite in the rugose coral samples may have been lower than suggested by the thin section in Figure \ref{EBSD_coral}, because samples were taken from the apical end of the corals where the void space between the septa was smaller, but sample preparation probably included some late-phase calcite along with pristine coral skeletal material.
-
-\begin{figure}[tb]
-\centering
-\includegraphics[width=90mm]{Fig_08-01.png}
-\caption{Scanning electron microscope images of flakes of brachiopod shell.
-Images were taken in secondary electron mode.
-Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
-Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
-Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2$^{\circ}$C; B) 33 $\pm$ 2$^{\circ}$C; C) 35 $\pm$ 1$^{\circ}$C.}
-\label{flakes}
-\end{figure}
 

### 2013-08-28 draft.tex
updated draft.tex to aug 28 version of word doc

@@
 
-\title{Carbonate clumped isotope constraints on Silurian ocean temperature and $\delta^{18}$O}
+\title{Carbonate clumped isotope constraints on Silurian ocean temperature \\and seawater $\delta^{18}$O}
 
@@
 \begin{abstract}
-Much of what we know about the history of Earth's climate derives from the geologic record of carbonates.
-The oxygen isotopic composition ($\delta^{18}$O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past.
-Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the $\delta^{18}$O of the water as well.
-The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set.
 
-We evaluated preservation and measured the clumped isotope composition of a suite of calcitic fossils from Silurian (ca.
+Much of what we know about the history of Earth's climate derives from the chemistry of carbonate minerals in the sedimentary record.
+The oxygen isotopic composition ($\delta^{18}$O) of calcitic marine fossils and cements has been widely used as a proxy for past seawater temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past.
+Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a calcite phase precipitated, allowing direct calculation of the $\delta^{18}$O of the water as well.
+The clumped isotope composition of calcites is also highly sensitive to and can help diagnose different modes of diagenetic alteration, enabling evaluation of preservation states and identification of the most pristine materials from within a sample set---critical information for evaluating the quality of paleoproxy data generated from carbonates.
+
+We evaluated preservation and measured the clumped isotope composition of a suite of calcitic fossils (primarily brachiopods and corals), sedimentary grains, and cements from Silurian (ca.
 433 Ma) stratigraphic sections on the island of Gotland, Sweden.
-Substantial variability in clumped isotope temperatures suggests differential preservation, complicating the generation of a stratigraphically resolved temperature history through these sections.
-However, micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to a modern ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$).
-This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time.
+Substantial variability in clumped isotope temperatures suggests differential preservation with alteration largely tied to rock-buffered diagenesis, complicating the generation of a stratigraphically resolved climate history through these sections.
+Despite the generally high preservation quality of samples from these sections, micro-scale observations of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to only a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to that estimated for a ``modern'' ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$).
+This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's ocean waters has remained broadly constant through time.
+
 \end{abstract}
@@
 The most widely used proxy for past ocean temperature is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate $\delta^{18}$O thermometer \citep{Urey1947}.
-This thermometer is based on the temperature dependence of the fractionation between the $\delta^{18}$O of carbonate, reported relative to the VPDB standard, and the $\delta^{18}$O of seawater, reported relative to the VSMOW standard.
-The carbonate $\delta^{18}$O proxy has two significant sources of complications.
+This thermometer is based on the temperature dependence of the fractionation between the $\delta^{18}$O of carbonate, here reported relative to the VPDB standard, and the $\delta^{18}$O of seawater, here reported relative to the VSMOW standard.
+The carbonate $\delta^{18}$O proxy has two significant complications.
 First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature.
-Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the $\delta^{18}$O of the seawater.
-The $\delta^{18}$O of seawater can be influenced by several processes, including local evaporation and mixing with meteoric water, as well as global changes in the size of continental glaciers, all of which are difficult to constrain in the geologic record.
+This is a fundamental challenge associated with carbonate minerals, which commonly undergo recrystallization during diagenesis and burial, altering primary geochemical signals and impacting the quality of proxy data.
+Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the water from which the carbonates precipitated, but also on the $\delta^{18}$O of the fluid (seawater and/or pore fluids).
+The $\delta^{18}$O of seawater can be influenced by several processes, including local evaporation and mixing with meteoric water, as well as global changes in the volume and isotopic composition of continental glaciers, all of which are difficult to constrain independently in the geologic record.
 
-The $\delta^{18}$O of the ocean has also been hypothesized to have changed gradually but substantially through Earth's history \citep{Jaffres2007, Veizer1999}.
-Gradual changes in the $\delta^{18}$O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the $\delta^{18}$O of seawater.
-The long-term average $\delta^{18}$O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust.
-High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the $\delta^{18}$O of ocean water by equilibrating the water with the average mantle $\delta^{18}$O$_{\text{VSMOW}}$ of 5.7$\permil$ \citep{Gregory1991}.
+Based on observations of carbonates, it was hypothesized that the $\delta^{18}$O of the ocean increased gradually but substantially through Earth's history---by $\sim$8$\permil$ over Phanerozoic time \citep{Jaffres2007, Veizer1999}.
+On 100-million-year timescales, gradual changes in $\delta^{18}$O are, in principle, plausible given current knowledge of the fluxes impacting the seawater $\delta^{18}$O budget.
+The long-term average $\delta^{18}$O of all the water on Earth's surface is largely controlled by the balance between two classes of processes within the silicate Earth: exchange during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange during low-temperature weathering of oceanic and continental crust \citep{Muehlenbachs1998}.
+High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the $\delta^{18}$O of ocean water by equilibrating the water with the average mantle $\delta^{18}$O$_{\text{VSMOW}}$ of 5.8$\permil$ \citep{Muehlenbachs1998}.
 Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the $\delta^{18}$O of seawater by preferentially moving $^{18}$O from water into rocks.
-The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995, Tajika1993, Wallmann2001} and the magnitude of continental weathering, which would lead to slow shifts in the $\delta^{18}$O of the ocean.
-The best geologic evidence for gradual change in seawater $\delta^{18}$O, however, is the subject of a long-standing controversy in geochemistry.
+Furthermore, we might reasonably anticipate that the relative magnitudes of high- and low-temperature weathering have changed over time with continental assembly and rifting \citep{Collins2003}, changes in seafloor spreading and production of oceanic crust \citep{Muller2008, Fornari1995}, climate \citep{Frakes2005}, and sedimentation \citep{Molnar2004}, unless a set of feedbacks, albeit enigmatic, operate to buffer the $\delta^{18}$O of seawater \citep{Gregory1981, Gregory1991, Muehlenbachs1976, Muehlenbachs1998, Jaffres2007}.
 
-This controversy arises from the complications associated with the carbonate $\delta^{18}$O proxy, which allow conflicting interpretations of the $\sim$8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}.
+The most widely cited evidence for gradual change in seawater $\delta^{18}$O, however, remains a longstanding controversy in geochemistry.
+This arises from the complications associated with the carbonate $\delta^{18}$O proxy, which allow conflicting interpretations of the observed $\sim$8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}.
 Several studies have proposed that early Paleozoic carbonates may have low $\delta^{18}$O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959, Keith1964, Land1995}, which tends to decrease the $\delta^{18}$O of carbonate \citep{Marshall1992}.
@@
 These screened fossils also record increasing $\delta^{18}$O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low $\delta^{18}$O values in the early Paleozoic.
-An alternative hypothesis holds that oceans could have been much warmer in the early Paleozoic \citep{Karhu1986, Knauth1976}.
-Assuming that the $\delta^{18}$O of the oceans was the same as today, the Cambrian oceans must have been $\sim$56$^{\circ}$C to produce carbonates with a $\delta^{18}$O$_{\text{VPDB}}$ of -8$\permil$.
-These warm temperatures seem incompatible with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading some studies to suggest that the increase in the $\delta^{18}$O of marine carbonates reflects a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.
+An alternative hypothesis holds that oceans could have been much warmer in early Paleozoic time \citep{Karhu1986, Knauth1976}.
+Assuming that the $\delta^{18}$O of the oceans was the same as today, Ordovician tropical oceans must have been $\sim$45$^{\circ}$C to produce carbonates with the observed average $\delta^{18}$O$_{\text{VPDB}}$ of -5.9$\permil$ \citep{Veizer1999}.
+Such warm temperatures present an apparent paradox with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading several authors to suggest that the increase in the $\delta^{18}$O of marine carbonates must therefore reflect a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.
 
-Carbonate clumped isotope thermometry provides a unique insight into this controversy because it provides data that can help to resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate records a temperature that is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration.
-The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}.
-Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the seawater.
+Carbonate clumped isotope thermometry supplies a unique insight into this problem because it provides data that can help resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration.
+The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in a wide diversity of shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}.
+Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the water from which the calcite was precipitated.
 
@@
 as cements lining fossil void space), the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived.
-We demonstrate here that carbonate clumped isotopes can be a more sensitive indicator of diagenetic alteration than the $\delta^{18}$O or the trace metal content of a carbonate.
+We demonstrate here that carbonate clumped isotopes can provide a more sensitive indicator of diagenetic alteration than the $\delta^{18}$O or the trace metal content of a carbonate.
 
-In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope arrays are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine end-member of a sample set.
-Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water.
+In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope arrays are also diagnostic of diagenetic regimes, allowing identification of the most pristine end-member of a sample set.
+Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and fluid-buffered diagenesis, respectively) and can also identify fluid-buffered diagenesis in the presence of meteoric water.
 As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water.
@@
 The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}.
-We demonstrate here that this distinction has important implications for the estimation of the temperature and $\delta^{18}$O of oceans in the past.
+We demonstrate here that this distinction has important implications for the estimation of the sea surface temperature and $\delta^{18}$O of seawater in the past.
 
-The present study investigates the implications of diagenetic alteration for the resolution of small temperature changes recorded in relatively well-preserved Paleozoic rocks.
-This study also provides a test of established proxies for diagenetic alteration as filters for clumped isotope data.
-Finally, our data add to the clumped isotope record of past ocean temperatures and $\delta^{18}$O, including samples from a different continent and a different paleocean basin than previous clumped isotope studies.
+\begin{figure*}[t]
+\centering
+\includegraphics[width=140mm]{Fig_01-01.png}
+\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}.
+Locations of stratigraphic sections are marked and labeled with location number, which corresponds to the sample locations in Supplementary Table 1.}
+\label{map}
+\end{figure*}
 
@@
 Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}.
-The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was 19$^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.
+The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was constrained to be 19$^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=140mm]{Fig_01-01.png}
-\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}.
-Approximate sample locations are marked and labeled with location number, which corresponds to the sample locations in Supplementary Table 1.}
-\label{map}
-\end{figure*}
-
-The oldest exposed strata on Gotland are the Lower Visby Beds, which consist of marls interbedded with fine-grained limestones deposited below storm wave base (Figure \ref{column}).
-Fossils are abundant throughout the section \citep{Calner2004a, Samtleben1996}.
-The contact between Lower and Upper Visby beds is marked by a bed with abundant fossils of the solitary rugose coral genus \textit{Phaulactis} \citep{Jeppsson1997, Jeppsson2006, Munnecke2003, Samtleben1996}.
-This contact is just above the Llandovery-Wenlock boundary as defined by conodont biostratigraphy \citep{Aldridge1993, Jeppsson1983, Mabillard1985}.
-We defined the \textit{Phaulactis} bed as a datum and all stratigraphic positions are measured relative to it, except those in the Slite Formation, which are measured relative to an arbitrary bed in the Slite Formation.
-The contact between Lower and Upper Visby Beds is also characterized by a facies change that represents a maximum flooding surface \citep{Calner2004b}.
+The oldest exposed strata on Gotland comprise the Lower Visby Formation, which consists of marls interbedded with fine-grained limestones deposited below storm wave base (Figure \ref{column}).
+Fossils are abundant throughout this unit \citep{Calner2004a, Samtleben1996}.
+The contact between the Lower and Upper Visby formations is marked by a condensed interval with abundant fossils of the large solitary rugose coral genus \textit{Phaulactis} \citep{Jeppsson1997, Jeppsson2006, Munnecke2003, Samtleben1996}.
+This contact occurs just above the Llandovery-Wenlock boundary as defined by conodont biostratigraphy \citep{Aldridge1993, Jeppsson1983, Mabillard1985}.
+We defined the \textit{Phaulactis} bed as a stratigraphic datum to align the sections, except those in the Slite Formation, which are measured relative to an arbitrary bed in the Slite Formation.
+The contact between the Lower and Upper Visby formations is also characterized by a facies change that represents a maximum flooding interval \citep{Calner2004b}.
 Carbonate content increases in the Upper Visby Beds, with a decreasing frequency of marl beds and the appearance of bioclastic limestones and reef mounds.
-Ripple marks indicate deposition above storm wave base.
-Fossils are especially abundant in the Upper Visby Beds \citep{Calner2004a, Samtleben1996}.
+Ripple bedforms and cross-stratification indicate deposition above storm wave base.
+The Upper Visby Formation is particularly fossiliferous \citep{Calner2004a, Samtleben1996}.
 
@@
 \caption{Lithology and isotope stratigraphy of the sampled sequence.
-On the left, the stratigraphic column shows the generalized lithostratigraphic units that were sampled for this study.
-On the right are the carbon and oxygen isotopic compositions of apparently well-preserved brachiopods, collected from the same section in Gotland by \cite{Bickert1997}.
-Figure modified from \cite{Bickert1997}.}
+On the left, the stratigraphic column shows the generalized lithostratigraphic units sampled in this study.
+On the right are the carbon and oxygen isotopic compositions of brachiopods, collected from the same sections by \cite{Bickert1997}.
+Isotope data is from \cite{Bickert1997}.}
 \label{column}
@@
 The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons.
-First, this section has long been recognized for its generally excellent preservation.
-The strata on Gotland are very nearly flat-lying, with a dip of less than 1$^{\circ}$, and display only minimal evidence of tectonic disturbance \citep{Calner2004a, Jeppsson1983}.
-Conodont fossils from this section yield a conodont alteration index (CAI) of one, which corresponds to a maximum temperature of $\sim$100$^{\circ}$C \citep{Epstein1976, Jeppsson1983, Wenzel2000}.
+
+First, this section has long been recognized for its generally excellent fossil preservation.
+The strata on Gotland are very nearly flat lying, with a dip of less than 1$^{\circ}$, display only minimal evidence of tectonic disturbance and have not been deeply buried \citep{Calner2004a, Jeppsson1983}.
+Conodont fossils from this section yield a conodont alteration index (CAI) of one, which suggests a maximum temperature of $\sim$100$^{\circ}$C \citep{Epstein1976, Jeppsson1983, Wenzel2000}.
 
@@
 As with previous paleoclimate studies we sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements).
-Many studies of paleoenvironment have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}.
+Many studies have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}.
 The fibrous calcite fabric characteristic of the secondary layer has been well characterized in modern and fossil brachiopods \citep{PerezHuerta2007, Samtleben2001, Schmahl2004}, providing a standard of comparison for the samples in our dataset.
 The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998, Brand2003, Brand2012, Mii1994, Grossman1996, Shields2003}.
-The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been used by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}.
+The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been examined by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}.
 Measurements of several fossil taxa allow comparison of the relative preservation of different calcitic materials and can help to detect taxon-specific ``vital effects'' if they are present.
 
-A third attractive feature of the Silurian section on Gotland is the coupled positive excursion in $\delta^{13}$C and $\delta^{18}$O that initiates at the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
+A third attractive feature of the Silurian section on Gotland is the previously observed positive excursion in $\delta^{13}$C that initiates at the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
 This excursion, termed the Ireviken event, is now recognized globally \citep{Munnecke2003}.
-In Gotland, a 3$\permil$ excursion in $\delta^{13}$C and 0.6$\permil$ excursion in $\delta^{18}$O attain a maximum rate of change across the contact between the Lower and Upper Visby Beds \citep{Munnecke2003}.
-The excursions coincide with a maximum flooding surface, but it is difficult to distinguish the relative contributions of progradation and eustatic sea level fall to this facies change \citep{Calner2004a}.
+In Gotland, a 3$\permil$ excursion in $\delta^{13}$C is coupled to a poorly resolved 0.6$\permil$ positive excursion in $\delta^{18}$O \citep{Munnecke2003}.
+It is somewhat surprising that the excursions coincide with a maximum flooding interval, but it is difficult to distinguish the relative contributions of sediment supply and eustatic sea level rise to this facies change \citep{Calner2004a}.
 The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.
 
-The coincident isotopic excursions and stratigraphic change suggest that the Ireviken represents a climate phenomenon, the nature of which is still debated.
+The coincident isotopic excursions and stratigraphic change suggest that the Ireviken Event captures a climate change phenomenon, the nature of which is still debated.
 The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998, Kaljo2003, Brand2006, Calner2008}.
-This hypothesis is supported by the widespread presence of glacial tillites in Brazil that have been dated to the latest Llandovery or earliest Wenlock \citep{Grahn1992}.
-Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Munnecke2003, Samtleben1996}.
+This hypothesis was thought to be supported by the presence of glacial tillites in Brazil that were at first interpreted to be early Wenlock in age \citep{Grahn1992}, but these diamictites were later re-interpreted as re-workings of earlier glacial deposits.
+Early Silurian glaciation in South America is now thought to have ended prior to the late Llandovery, before the Ireviken Event \citep{DiazMartinez2007a, DiazMartinez2007b}.
+Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Jeppsson1990, Munnecke2003, Samtleben1996}.
 According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves.
@@
 Low terrigenous influxes would have allowed the growth of carbonate reefs as observed in the H\"{o}gklint Formation.
-
-In the present study, we test the ability of clumped isotopes to resolve the changes in ocean temperature and seawater $\delta^{18}$O that contributed to the 0.6$\permil$ shift in the $\delta^{18}$O of carbonates through the Ireviken event.
-From first principles, it will be challenging but not impossible for the carbonate clumped isotope thermometer to resolve these changes, because a shift of 0.6$\permil$ in $\delta^{18}$O corresponds to a change in temperature of approximately 2$^{\circ}$C.
-Single measurements of carbonate clumped isotope composition can attain a resolution of $\pm$1$^{\circ}$C, while running multiple replicates of a sample can improve the resolution to $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.
+These hypotheses make different predictions for temperature and the oxygen isotopic composition of seawater.
+Here we test our ability to resolve possible changes in ocean temperature and/or seawater $\delta^{18}$O through the Ireviken event using stratigraphic samples from Gotland.
+Resolving a change this small with the uncertainties associated with carbonate clumped isotope thermometer is challenging but not impossible, because a shift of 0.6$\permil$ in $\delta^{18}$O corresponds to a change in temperature of approximately 2$^{\circ}$C.
+Though the uncertainties are commonly higher, single measurements of carbonate clumped isotope composition can attain a resolution of $\pm$1$^{\circ}$C, while running multiple replicates of a sample can improve the resolution to $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.
 
@@
 
-Samples were collected from the island of Gotland, Sweden during the summer of 2011.
+Stratigraphic sections were measured and samples were collected from the island of Gotland, Sweden during the summer of 2011.
 Each sample was labeled with a location, as shown in Figure \ref{map}, and a stratigraphic height.
-For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
+For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby formations (Figure \ref{column}).
 Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}, as well as new bulk carbonate $\delta^{13}$C measurements (Fike et al.
@@
 
-Of the 22 brachiopods chosen for sampling, 20 belonged to the species group \textit{Atrypa reticularis}.
+Of the 22 brachiopods chosen for sampling, 20 belonged to the species \textit{Atrypa reticularis}.
 We targeted this species because it is abundant in the sampled strata, the shell has a thick secondary fibrous layer, and it was used in several previous paleoenvironmental studies in Gotland \citep{Bickert1997, Munnecke2003, Samtleben2001}.
@@
 Brachiopods, bryozoans, crinoids, gastropods and ostracods were sampled by flaking the surface of the skeleton or shell using a dental chisel.
-Every flake was visually inspected under a microscope at circa $25\times$ power.
+Every flake was visually inspected under a stereo microscope at circa $25\times$ power.
 Flakes with visible pyrite grains or rough-textured recrystallized regions were rejected and not included in the material for analysis.
@@
 The apical end was then cut off using a diamond microsaw, and powdered using a mortar and pestle.
-The apical ends of the corals were targeted because the coral septa are closely spaced together there, leaving less room for cement to precipitate between the septa.
+The apical ends of the corals were targeted because the coral septa are closely spaced together there, leaving less room for sediment and cement between the septa.
 However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite.
-Finally, to assess possible diagenetic end-members, micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
 
@@
 Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
-Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2$^{\circ}$C; B) 33 $\pm$ 2$^{\circ}$C; C) 35 $\pm$ 1$^{\circ}$C.}
+Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2$^{\circ}$C, B) 33 $\pm$ 2$^{\circ}$C, C) 35 $\pm$ 1$^{\circ}$C.}
 \label{flakes}
@@
 
+All fossil samples shown in figures and used to estimate Silurian seawater temperature and $\delta^{18}$O  were prepared as described above, with the aim of minimizing the inclusion of cement, sediment and altered material.
+However, in a few separate test cases we purposefully sampled cements, micrites and other textures that are independently known to be more susceptible to diagenetic alteration, as a way to assess possible diagenetic end-members.
+Micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
+During the preparation of one brachiopod sample, flakes with large rough-textured recrystallized regions were picked and analyzed separately from the pristine flakes of fibrous secondary layer.
+
 While all fossils were inspected under a microscope prior to sampling, we were not able to inspect all fossils in thin section, because carbonate clumped isotope measurements require large amounts of material, some of which is lost during preparation of thin sections.
 However, 30 $\mu$m-thick uncovered, polished thin sections were obtained from cuts through four representative fossils, two brachiopods and two rugose corals.
-These thin sections were inspected first in transmitted and reflected light to characterize calcite textures, and were then examined using several different spectroscopic techniques, described below.
+These thin sections were inspected first in transmitted and reflected light to characterize calcite textures, and were then examined using several different micro-scale spectroscopic techniques, described below.
 
@@
 
-Electron backscatter diffraction (EBSD) provides an independent measure of recrystallization by mapping the orientation of calcite crystals in a thin section.
+Electron backscatter diffraction (EBSD) provides an independent measure of the degree and style of recrystallization by mapping the orientation of calcite crystals in a thin section \citep[e.g.][]{Bergmann2013}.
 This allows more detailed characterization of calcite fabrics than is possible using optical microscopy, and in the case of brachiopods, comparison to pristine modern examples \citep{PerezHuerta2007}.
@@
 
-Flakes of the secondary fibrous layer from twelve different brachiopods were also imaged to further inspect the preservation of original fibrous calcite texture.
+Flakes of the secondary fibrous layer from twelve different brachiopods were also imaged via electron microscopy to further inspect the preservation of original fibrous calcite texture at a finer scale than available from the reflected light microscope.
 The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
@@
 The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer (E-probe).
-Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite.
+Using the JEOL map analysis program, the beam was calibrated on crystal standards of dolomite, siderite, strontianite and rhodochrosite.
 Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nA, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
@@
 The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998, Brand2012, Came2007, Finnegan2011}.
-Diagenetic recrystallization of calcite is thought to enrich iron (Fe) and manganese (Mn) and deplete strontium (Sr) \citep{Brand1980, Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples.
+Diagenetic recrystallization is thought to enrich calcite in iron (Fe) and manganese (Mn) and deplete strontium (Sr) from calcite \citep{Brand1980, Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples.
 Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
@@
 
-Each sample was measured between one and three times.
+Each sample was measured between one and three times depending on the amount of available material.
 For each measurement approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90$^{\circ}$C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate.
@@
 Cleaned CO$_2$ was analyzed on a Finnigan MAT253 gas source mass spectrometer configured to collect masses 44 through 49.
-The sample gas was measured in eight separate acquisitions, each of which was bracketed by a measurement of reference CO$_2$ with known $\delta^{13}$C and $\delta^{18}$O.
+The sample gas was measured for eight acquisitions of seven cycles each, where each cycle was bracketed by a measurement of reference CO$_2$ with known $\delta^{13}$C and $\delta^{18}$O.
 The $\delta^{13}$C, $\delta^{18}$O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample relative to the reference CO$_2$.
@@
 
+\begin{figure*}[t]
+\centering
+\includegraphics[width=\textwidth]{Fig_04_color-01.png}
+\caption{The $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
+For the Lower \& Upper Visby and H\"{o}gklint formations, a height of zero meters is defined by the contact between the Lower Visby Formation and the Upper Visby Formation.
+For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop.
+Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included for comparison.
+Error bars in clumped isotope temperature represent the one-sigma standard error of the mean.}
+\label{section}
+\end{figure*}
+
 All $\Delta_{47}$ values are projected into the absolute reference frame \citep{Dennis2011}.
-$\Delta_{47}$ values of two California Institute of Technology intralab standards were converted to the absolute reference frame using a secondary transfer function (STF) constructed from the heated and equilibrated gases measured during the same week as the standards.
+$\Delta_{47}$ values of two California Institute of Technology intralaboratory standards were converted to the absolute reference frame using a secondary transfer function (STF) constructed from the heated and equilibrated gases measured during the same week as the carbonate standards.
 Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000$^{\circ}$C for three hours, and then cooled rapidly to room temperature.
@@
 Heated and equilibrated gases were cleaned following the same procedure as the samples.
-The STF was constructed by plotting the accepted $\Delta_{47}$ values of 0.0266$\permil$ for heated gases and 0.9252$\permil$ for equilibrated gases against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame \citep{Dennis2011}.
-The STF was used to convert the $\Delta_{47}$ values of the two intralab standards from the \cite{Ghosh2006} reference frame to the absolute reference frame following equation C1 in \cite{Dennis2011}.
+The STF was constructed by plotting the accepted absolute reference frame $\Delta_{47}$ values of 0.0266$\permil$ for heated gases and 0.9252$\permil$ for equilibrated gases against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame \citep{Dennis2011}.
+The STF was used to convert the $\Delta_{47}$ values of the two intralaboratory standards from the \cite{Ghosh2006} reference frame to the absolute reference frame following equation C1 in \cite{Dennis2011}.
 
 All of the samples in this study were projected into the absolute reference frame using a separate STF constructed for each week of measurements.
-To create the STF, the accepted $\Delta_{47}$ values for the two intralab standards and heated gases were plotted against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
+To create the STF, the accepted $\Delta_{47}$ values for the two intralaboratory standards and heated gases were plotted against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
 Equilibrated gases were measured during some but not all weeks of sample measurements, so the accepted and observed $\Delta_{47}$ values of equilibrated gases were included in the weekly STF during the weeks when these measurements were available.
 $\Delta_{47}$ values of the samples in the \cite{Ghosh2006} reference frame were corrected using this STF following equation C1 in \cite{Dennis2011}.
-Finally, the resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90$^{\circ}$C by adding 0.092$\permil$ \citep{Henkes2013}.
+The resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90$^{\circ}$C by adding 0.092$\permil$ \citep{Henkes2013} to produce the final value ($\Delta_{47-\text{RF-AC}}$).
 
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=\textwidth]{Fig_04_color-01.png}
-\caption{The $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
-For the Lower \& Upper Visby and H\"{o}gklint Formations, a height of zero meters is defined by the contact between the Lower Visby Beds and the Upper Visby Beds.
-For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop.
-Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included.
-Error bars represent the standard error of the mean.}
-\label{section}
-\end{figure*}
-
 \section{Results and discussion}
@@
 The $\delta^{13}$C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}).
-We resolve a 3$\permil$ positive excursion in $\delta^{13}$C, beginning just below the boundary between the Lower and Upper Visby Beds, and continuing for at least 50 meters.
-The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation record a return to lower $\delta^{13}$C$_{\text{VPDB}}$ values of approximately -0.5$\permil$.
-The $\delta^{13}$C trend through the section is comparable to the brachiopod $\delta^{13}$C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
+We observe a 3$\permil$ positive excursion in $\delta^{13}$C, beginning just below the boundary between the Lower and Upper Visby formations, and continuing for $>$50 meters.
+The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation all have consistently lower $\delta^{13}$C$_{\text{VPDB}}$ values of approximately -0.5$\permil$.
+The $\delta^{13}$C trend through the section is similar to the brachiopod $\delta^{13}$C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
 
 The $\delta^{18}$O values of fossil material are consistent within beds, but micrite and cements tend to have lower $\delta^{18}$O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
-We are not able to resolve an excursion in $\delta^{18}$O.
-This may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}, or because our samples include more contamination from secondary calcite phases.
-The average $\delta^{18}$O$_{\text{VPDB}}$ of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples consistent with those of \cite{Munnecke2003}, but our Upper Visby samples more depleted in $^{18}$O.
+It is interesting that we were not able to resolve an excursion in carbonate $\delta^{18}$O.
+This may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}, although even with the additional samples, the previously reported excursion is poorly resolved.
+No $\delta^{18}$O excursion was observed in the isotopic data generated from bulk sediment (Fike et al.
+in prep.).
+The average $\delta^{18}$O$_{\text{VPDB}}$ values of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples extremely similar to those of \cite{Munnecke2003}, but our Upper Visby samples slightly more depleted in $^{18}$O.
 The low $\delta^{18}$O values of Upper Visby fossils may be caused by increased contamination from secondary cements, which could be expected to accompany the transition from marl to limestone.
-Higher in the section, the $\delta^{18}$O of the fossils in this study appear consistent with values reported by \cite{Bickert1997}.
+Higher in the section, the $\delta^{18}$O values of the fossils in this study are consistent with values reported by \cite{Bickert1997}.
 
 Clumped isotope temperatures of samples are highly variable within beds, often covering a range of 10$^{\circ}$C within a single bed (Figure \ref{section}).
-The lowest temperature sample, a brachiopod, is from the Slite Formation, deposited well after the hypothesized climate changes associated with the isotopic excursions in the Lower and Upper Visby Beds.
-The large scatter in clumped isotope temperatures suggests variable preservation quality among these samples.
-Even brachiopods, which are commonly thought to be relatively resistant to diagenetic alteration, have a range in temperature of more than 15$^{\circ}$C.
-The variable preservation quality of these samples means that beds with fewer samples are less likely to contain a pristine fossil with a reliable temperature.
-Because many of the beds in this study are only represented by a few samples, we are not able to stratigraphically resolve a change in temperature with time.
+The lowest temperature sample, a brachiopod, is from the Slite Formation, deposited well after the hypothesized climate changes associated with the isotopic excursions in the Lower and Upper Visby formations.
+The large, within-bed scatter in clumped isotope temperatures is unlikely to be primary and suggests variable preservation quality among these samples.
+Even the secondary fibrous shell layer of brachiopods, which is commonly thought to be relatively resistant to diagenetic alteration, displays a range in temperature of more than 15$^{\circ}$C.
+The variable preservation quality of these samples means that beds with fewer samples are less likely to contain a pristine fossil with a reliable temperature---introducing a sample bias.
+Because many of the beds in this study are only represented by a few samples, we are not able to stratigraphically resolve what we would interpret as a change in temperature with time.
 
@@
 \captionsetup{labelformat=empty} 
-\caption{Figure 6: Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
-The negative correlation between temperature and the log of Sr concentration in rugose corals is weakly significant (79\% confidence), and strengthens when combined with the rest of the samples.
-There are also weakly significant positive correlations between temperature and the log of Fe concentration (63\% confidence) and between temperature and the log of Mn concentration (83\% confidence) among all the samples combined.
+\caption{Figure \ref{metals}: Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
+The negative correlation between temperature and the log of Sr concentration in rugose corals is weakly significant and strengthens when combined with the rest of the samples.
+There are also weakly significant positive correlations between temperature and the log of Fe concentration and between temperature and the log of Mn concentration among all the samples combined.
 These plots demonstrate clear differences in trace metal content between the two fossil taxa and the later carbonate phases.
-Brachiopods have the lowest average temperature of all the sample subsets, and also tend to have the most pristine trace metal signatures, with less Fe and Mn and more Sr than the other samples.
-Micrites and cements have the highest average temperature and tend to have the most altered trace metal signatures, with high Fe and Mn and low Sr concentrations.}
+Brachiopods have the lowest average temperature of all the sample subsets, and also tend to have the most pristine trace metal signatures, with less Fe and Mn than the other samples.
+Micrites and cements have the highest average temperature and tend to have the most altered trace metal signatures, with high Fe and Mn concentrations.}
 \label{Brand}
@@
 \captionsetup{labelformat=empty} 
-\caption{Figure 5: Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
-The brachiopods in this study have less Mn and more Sr than most of the brachiopods analyzed by \cite{Brand2012} which are also shown in this figure.
-Lines represent estimates by different authors of the range of pristine concentrations defined for brachiopods.
-Rugose corals generally have more Fe (not shown), more Mn and less Sr than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.
+\caption{Figure \ref{Brand}: Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
+The brachiopods in this study have more Sr on average than the brachiopods analyzed by \cite{Brand2012}, which are also shown in this figure.
+Lines represent limits suggested by different authors for the range of pristine concentrations defined for brachiopods.
+Rugose corals generally have more Fe (not shown) and more Mn than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.
 Figure modified from \cite{Brand2012}.}
@@
 
-Although we are not able to resolve a change in temperature, we can determine a representative temperature for these sections by using constraints on the mechanism of diagenetic alteration to identify the most pristine samples.
-All of the fossil samples have fairly high values of $\delta^{18}$O, making alteration by meteoric water unlikely.
-This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at high temperature.
-These higher temperatures could be caused by incorporation of late-stage cements or wholesale recrystallization.
+Although we are not able to measure a change in sea surface temperature through the Ireviken Event, we can use our data to constrain more broadly temperature and $\delta^{18}$O of seawater by uncovering the mechanism of diagenetic alteration to identify the most pristine samples.
+All of the fossil samples have fairly high $\delta^{18}$O values, making low-temperature alteration by meteoric water unlikely.
+This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at elevated temperatures.
+These higher clumped isotope temperatures could be caused by incorporation of late-stage cements or wholesale recrystallization.
 We tested this hypothesis using independent methods, described above, to assess the preservation quality of representative samples.
@@
 
-Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration, suggest that all the brachiopods in this study are comparatively pristine.
+Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration \citep{Azmy1998, Brand2012, Came2007, Finnegan2011}, suggest that all the brachiopods in this study are comparatively pristine.
 However, micro-scale studies of calcite fabrics, as well as clumped isotope measurements, demonstrate that there is a range in preservation quality among the brachiopods that cannot be identified on the basis of trace metal measurements alone.
@@
 Then we consider EBSD and SEM evidence for the inclusion of diagenetic material in the typical fossil sample.
-Finally, we explore the use of clumped isotopes to distinguish the best-preserved samples from among those that would ordinarily all be identified as pristine based on their trace metal contents.
+Finally, we turn to clumped isotopes to distinguish the best-preserved samples from among those that would ordinarily all be identified as pristine based on their trace metal contents.
 
-Clumped isotope thermometry reveals variable preservation among the brachiopods in this study, so we expected to observe equivalent variations in trace metal content.
-Diagenetic recrystallization of brachiopod skeletal calcite is thought to enrich Fe and Mn because these metals are insoluble in oxic seawater but often found at elevated concentrations in anoxic basinal fluids.
-Diagenetic recrystallization is also thought to deplete Sr because brachiopod skeletons concentrate Sr relative to its abundance in seawater and meteoric waters \citep{Brand1980, Shields2003}.
+Clumped isotope thermometry implies variable preservation among the brachiopods in this study, so we expected to observe equivalent variations in trace metal content.
+Diagenetic recrystallization of brachiopod skeletal calcite is thought to enrich Fe and Mn because these metals are insoluble in oxic seawater but often found at elevated concentrations in anoxic pore fluids \citep{Brand1980, Shields2003}.
+Diagenetic recrystallization is also thought to deplete Sr because brachiopod skeletons concentrate Sr relative to its abundance in seawater and meteoric waters \citep{Brand1980, DePaolo2011, Schrag1995}.
 Many previous studies have measured the concentrations of these three elements in fossil brachiopods, and some have defined limits for the Mn and Sr concentrations of brachiopods that can be considered pristine.
-The bulk Mn and Sr concentrations of all brachiopods in this study fall within these limits (Figure \ref{Brand}); the small variations in trace metal concentrations among the brachiopods would not ordinarily be considered as evidence for variable preservation.
+The bulk Mn and Sr concentrations of all brachiopods in this study fall within these proposed limits (Figure \ref{Brand}), the small variations in trace metal concentrations among the brachiopods would not ordinarily be considered as evidence for variable preservation.
 
@@
 
-Given our hypothesis that the lowest temperature samples are the most pristine, we also expected to observe a correlation between clumped isotope temperatures and trace metal evidence for diagenetic alteration.
+Under the hypothesis that the lowest-temperature samples are the most pristine, we also expected to find a correlation between clumped isotope temperatures and trace metal evidence for diagenetic alteration, as observed for similar-aged fossils by \cite{Finnegan2011}.
 We tested brachiopods, rugose corals, and micrites/cements separately and combined for the robustness of a linear correlation between temperature and the log of trace metal concentrations (Figure \ref{metals}).
-We performed a hypothesis test on each pair of variables to test the probability (p) that the observed correlation is due to random variation when the true correlation is zero.
+We performed a hypothesis test on each pair of variables to test the probability (\textit{p}) that the observed correlation is due to random variation when the true correlation is zero.
 We used a Bonferroni correction to adjust the significance level ($\alpha$) to account for the multiple hypothesis tests.
-There are weakly significant positive correlations between temperature and Fe concentration (63\% confidence) and also between temperature and Mn concentration (83\% confidence) when all samples are combined.
-These correlations are likely caused by differences between fossil taxa and later phases; brachiopods have the lowest average temperature, Fe, and Mn concentrations, while micrites and cements combined have the highest average temperature, Fe and Mn concentrations.
-There is also a weakly significant (79\% confidence) negative correlation between temperature and Sr concentration among rugose corals, which strengthens (to 99\% confidence) when corals are combined with brachiopods, and with micrites and cements.
-However, none of the other variable pairs are significantly correlated within individual taxa (Supplementary table 2).
-These tests may be complicated by contamination from early void-filling cements, which we expect to have low temperatures but altered trace metal signatures.
+There are weakly significant positive correlations between temperature and Fe concentration and also between temperature and Mn concentration when all samples are combined.
+These correlations are largely due to differences between fossil taxa and later phases, brachiopods have the lowest average temperature, Fe, and Mn concentrations, while micrites and cements combined have the highest average temperature, Fe and Mn concentrations.
+There is also a weakly significant negative correlation between temperature and Sr concentration among rugose corals, which strengthens (to 99\% confidence) when corals are combined with brachiopods, and with micrites and cements.
+Trace metal concentrations reflect broad differences between taxa and textures, however none of the other variable pairs are significantly correlated within individual taxa (Supplementary Table 2).
+These tests may be complicated by contamination from early void-filling cements, which have low temperatures but ``altered'' trace metal signatures because they precipitate from pore fluids.
 Variability in Mn concentrations of low-temperature samples may obscure a positive correlation between temperature and Mn concentration, but low-temperature contamination does not appear to explain the lack of correlation between temperature and Fe or Sr.
-We suggest that the most likely explanation for these observations is that trace metal concentrations, with the exception of Sr in rugose corals, are relatively insensitive indicators of diagenetic alteration within fossil taxa, and are only sensitive enough to reflect broad differences between taxa.
+From these observations we suggest that trace metal concentrations, perhaps with the exception of Sr in rugose corals, are relatively insensitive indicators of alteration within fossil taxa under mild burial diagenesis, and in this case are only sensitive enough to reflect broad differences between taxa.
 
@@
 Variability in primary trace metal concentrations, which is large among modern brachiopods \citep{Brand2003}, may also mask any trends superimposed by diagenetic alteration.
-However, E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization need not necessarily have any effect on Fe and Mn concentrations.
-A recrystallized region of a rugose coral septum is clearly visible as a dark area in reflected light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B).
+However, E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization may not strongly affect Fe and Mn concentrations.
+A recrystallized region of a rugose coral septum is clearly visible as a dark area in transmitted light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B).
 However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D).
-This is only one example, but it suggests that recrystallization may not affect trace metal concentrations, and instead, variations in trace metal concentrations may be dominated by contamination from cements and sediments.
+This is only one example, but it highlights relatively large calcite domains within samples that were recrystallized but this is not reflected in the trace metal concentrations.
 The trace metal signature associated with recrystallization likely depends on conditions during recrystallization, suggesting that this coral underwent closed-system diagenesis.
@@
 We therefore turn to micro-scale studies of calcite textures as indicators of diagenetic alteration.
-Reflected-light and EBSD images of rugose corals reveal substantial amounts of late-phase diagenetic calcite, including recrystallized regions and cements (Figure \ref{EBSD_coral}A \& B).
+Reflected-light and EBSD images of rugose coral fossil skeletons reveal substantial domains of late-phase diagenetic calcite, including recrystallized regions and cements (Figure \ref{EBSD_coral}A \& B).
 As discussed above, recrystallized regions are visible as dark colors in reflected light and aligned calcite crystals in EBSD images.
 Cements in Figure \ref{EBSD_coral}A \& B are characterized by large, subequant, interlocking calcite crystals.
-The fraction of late-phase calcite in the rugose coral samples may have been lower than suggested by the thin section in Figure \ref{EBSD_coral}, because samples were taken from the apical end of the corals where the void space between the septa was smaller, but sample preparation probably included some late-phase calcite along with pristine coral skeletal material.
 
-In contrast to rugose corals, the EBSD image of a low-temperature brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Figure \ref{EBSD_brach}A).
+In contrast to the rugose corals, the EBSD image of a low-temperature (35$^{\circ}$C) brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Figure \ref{EBSD_brach}A).
 Within the brachiopod shell, calcite crystals are elongated approximately parallel to the surface of the shell, as expected for the fibrous secondary layer.
 This preferential orientation is represented by the vertical band in the stereographic projection of crystallographic planes (Figure \ref{EBSD_brach}B).
-No regions of recrystallization are visible within the shell.
+No large regions of recrystallization are visible within the shell.
 There are large interlocking cement crystals in the upper left hand corner of the EBSD image, but these formed in the void space between the two valves, and would not have been included in the sample for clumped isotope analysis.
-Based on visual inspection of the EBSD image of this brachiopod's secondary fibrous layer, we suggest that this layer contains a maximum of 5\% recrystallized material.
+Based on visual inspection of the EBSD image of this brachiopod's secondary fibrous layer, we suggest that this sample contains a maximum of 5\% recrystallized material.
 
@@
 
-The micro-scale studies of calcite textures in Figures \ref{EBSD_coral}--\ref{flakes} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite.
+The micro-scale studies of calcite textures in Figures  \ref{flakes}, \ref{EBSD_coral} and \ref{EBSD_brach} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite.
 Variable fractions of late-phase calcite likely contribute to the range in temperature among fossil samples.
-We only performed SEM and EBSD analysis on a small subset of samples, so we will not attempt to use this evidence to compare the relative preservation quality of low and high-temperature samples.
-An important conclusion that can be drawn from our micro-scale studies is that the low temperature brachiopods imaged in Figures \ref{EBSD_brach} and \ref{flakes} probably contain only trace amounts of late-phase calcite, likely less than 5\% of the sample.
-We therefore look to these samples as the most reliable records of ocean temperature and $\delta^{18}$O during the Silurian.
+Due to constraints on the amount of available material for analysis, we only performed SEM and EBSD analysis on a small subset of samples, but an important conclusion that can be drawn from the electron microscopy and spectroscopy results is that the low temperature brachiopods imaged in Figures \ref{flakes} and \ref{EBSD_brach} probably contain at most only trace amounts of late-phase calcite, variable but likely less than 5\% of the sample.
+We therefore look to where these samples fall on diagenetic trajectories to estimate Silurian ocean temperature and the value of seawater $\delta^{18}$O.
 
@@
 \caption{The clumped isotope temperature and calculated water $\delta^{18}$O$_{\text{VSMOW}}$ for all samples analyzed in this study.
-Error ellipses represent one standard deviation of the mean of all measurement acquisitions for each sample.
+The contours represent lines of constant carbonate $\delta^{18}$O$_{\text{VPDB}}$.
+We approximate the correlated error as symmetric using ellipses that represent one standard deviation of the mean of all measurement acquisitions for each sample.
 The positive correlation between temperature and water $\delta^{18}$O$_{\text{VSMOW}}$ is diagnostic of rock-buffered diagenesis.
-Under this regime, the lowest-temperature samples are the most pristine.}
+Under this regime, the lowest-temperature samples are the most pristine.
+Note that some cement and micrite samples fall off the trend, suggesting the effects of a more open system diagenetic process.}
 \label{ellipses}
@@
 
-To estimate the temperature and $\delta^{18}$O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore preserve the most reliable record of Silurian ocean conditions.
-This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected.
-As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the $\delta^{18}$O of interstitial water.
+To estimate the temperature and $\delta^{18}$O of Silurian oceans, we first need identify the diagenetic regime responsible for the observed differential alteration of the samples.
+As outlined above, different diagenetic processes have characteristic effects on the relationship between temperature and the $\delta^{18}$O of interstitial water.
 When the clumped isotope temperatures of the samples in this study are plotted versus the $\delta^{18}$O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}).
-This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is diagnostic of rock-buffered diagenesis.
-Importantly, during rock-buffered diagenesis, more altered samples have higher temperatures, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.
+This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is characteristic of rock-buffered diagenesis.
+Importantly, during rock-buffered diagenesis, more altered samples have higher temperatures, which is consistent with the hypothesis that the lowest temperature samples are the most pristine.
 
-\begin{figure}[b]
+\begin{figure}[t]
 \centering
@@
 \caption{Comparisons between the $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of brachiopod samples and adjacent micrite.
-Micrite is expected to be more easily altered than brachiopod shell material, so the consistently higher temperature of micrite supports our hypothesis that the lowest temperature samples are the most pristine.}
+Micrite is more easily altered than brachiopod shell material, so the consistently higher temperature of micrite supports the hypothesis that the lowest temperature samples comprise the more pristine end of the diagenetic trajectory shown in Figure \ref{ellipses}.
+The generally depleted $\delta^{18}$O values of micrites compared to brachiopods suggest some influence from fluid-buffered diagenesis.}
 \label{comparisons}
@@
 
-Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis \citep{Dennis2013, Huntington2011, Quade2013, VanDeVelde2013}, and we cannot rule it out as a contributor to our results.
+Additional evidence for a diagenetic regime that acted to increase the apparent temperature of more altered samples is provided by comparisons of the clumped isotope temperature of different calcite textures within samples.
+We measured the clumped isotope composition of micrites adjacent to four brachiopods, and in all four cases, the temperature of the micrite was higher than the temperature of the neighboring brachiopod (Figure \ref{comparisons}).
+Micrite is expected to be less pristine than the brachiopod samples because it far more commonly recrystallizes during burial and can include different generations of cements.
+Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
+Interestingly, while the $\delta^{13}$C values of micrites are similar to those of adjacent brachiopods, three of the four micrites tested are depleted in $\delta^{18}$O relative to the adjacent brachiopods, suggesting that the micrites may have undergone diagenesis in the presence of a higher water-to-rock ratio.
+
+In a similar experiment, recrystallized skeletal material was also shown to be associated with higher clumped isotope temperatures.
+One brachiopod sample was split into two fractions during picking.
+The first fraction consisted of pristine flakes of fibrous secondary layer and had a temperature of 38$\pm$3$^{\circ}$C, while the other fraction was comprised of flakes with large rough-textured recrystallized regions and had a temperature of 43$\pm$3$^{\circ}$C, where the error estimates represent the one-sigma standard error of the mean over all acquisitions.
+The results of this experiment provide further support for the hypothesis that the lowest-temperature samples are the most pristine, and also demonstrate the importance of careful picking when sampling brachiopods.
+
+Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis \citep{Dennis2013, Huntington2011, Quade2013, VanDeVelde2013}, but we suggest that solid-state diffusion is not a major contributor to our results.
 During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite.
 This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and re-precipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
-Calcites with different distributions and concentrations of trace elements have been demonstrated to have different rates of solid-state diffusion \citep{Passey2012}, which means that this process could potentially explain the observed variability in temperature among the samples.
-Temperatures as low as 100$^{\circ}$C may allow solid-state diffusion to increase the clumped isotope temperature of a calcite, provided that the calcite experiences these temperatures for millions of years \citep{Dennis2010}.
-100$^{\circ}$C is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983, Wenzel2000}, and it is difficult to estimate the length of time for which Gotland strata were exposed to elevated temperature.
-However, the effects of solid-state diffusion are consistent with our hypothesis that the lowest temperature samples are the most pristine, because solid-state diffusion increases the clumped isotope temperature of calcite.
+We argue that solid-state diffusion did not strongly affect any of our samples because Gotland strata were exposed to a maximum burial temperature of only 100$^{\circ}$C, as constrained by the conodont alteration index \citep{Jeppsson1983, Wenzel2000}, Temperatures above 100$^{\circ}$C are thought to be required for solid-state diffusion to increase the clumped isotope temperature of a calcite \citep{Dennis2010}.
+We cannot rule out a minor influence from solid-state diffusion, but due to the shallow burial history and low conodont alteration index of Gotland strata, we are confident that the lowest-temperature, most pristine samples in our dataset record a temperature that is very close to the primary seawater temperature.
 
-Additional evidence for a diagenetic regime that acted to increase the temperature of more altered samples is provided by comparisons of the clumped isotope temperature of different calcite textures (Figure \ref{comparisons}).
-We measured the clumped isotope composition of micrites adjacent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod.
-Micrite is expected to be less pristine than the brachiopod samples because micrite was cemented post-burial, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis.
-Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
-
-The lowest-temperature sample of the apparently pristine best preserved samples we studied, a brachiopod (G13 10.5 to 10.8 B1, Slite Fm.), suggests that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$ (Figure \ref{ellipses}), where the error estimates represent the one-sigma standard error of the mean over all acquisitions.
-The two next-most pristine samples, also brachiopods (G11 0.2 to 0.4 B6, stratigraphic height 1.3 m; G2 0.5 to 0.6 B2, stratigraphic height 0.5 m), have temperatures of 33$\pm$2$^{\circ}$C and 34$\pm$1$^{\circ}$C.
+The lowest-temperature sample within our sample set, a brachiopod (G13 10.5 to 10.8 B1, Slite Fm.), suggests that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$ (Figure \ref{ellipses}).
+The two next-most pristine samples, also brachiopods (G11 0.2 to 0.4 B6, stratigraphic height 1.3 m, G2 0.5 to 0.6 B2, stratigraphic height 0.5 m), have temperatures of 33$\pm$2$^{\circ}$C and 34$\pm$1$^{\circ}$C.
 Together with the $\delta^{18}$O$_{\text{VPDB}}$ of each brachiopod, these temperatures can be used to calculate seawater $\delta^{18}$O$_{\text{VSMOW}}$ values of -2.2$\pm$0.4$\permil$ and -1.2$\pm$0.2$\permil$, respectively.
 One caveat of these values is that they are each based on only one measurement due to sample-size limitations.
-The lowest-temperature brachiopod was picked a second time, but to collect enough material for analysis, some less pristine flakes had to be included, which may explain the 41$\pm$4$^{\circ}$C temperature of this re-sample (G13 10.5 to 10.8 B1 leftovers).
-However, we argue that the differences between the temperatures and $\delta^{18}$O values implied by the three most pristine samples are small enough that they essentially represent the same warm, tropical Silurian ocean water.
+The lowest-temperature brachiopod was picked a second time, but to collect enough material for analysis, some lower quality flakes had to be included, which yielded a temperature of 41$\pm$4$^{\circ}$C (G13 10.5 to 10.8 B1 leftovers).
 
-A second caveat of our estimates of the temperature and $\delta^{18}$O of Silurian tropical oceans is that they are technically upper estimates.
-We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature recorded by the samples.
-We are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine samples.
+Due to the increase of observed sample temperatures during diagenesis, our estimates of the temperature and $\delta^{18}$O of Silurian tropical oceans are technically upper limits.
+We are not able to exclude the possibility that small amounts of diagenesis affected all the samples in our study, even the most pristine samples.
 However, micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, comprising a maximum of 5\% of the material.
@@
 In this case, if the measured temperature of the sample is 31$^{\circ}$C and 5\% of the sample material was recrystallized at 100$^{\circ}$C, the primary temperature of the pristine fossil material would be 28.3$^{\circ}$C.
-Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 3$^{\circ}$C, provided that solid-state diffusion has not significantly altered the temperature of the most pristine samples in our dataset.
+Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 3$^{\circ}$C.
+
+Our clumped isotope measurements indicate that Silurian tropical oceans were broadly similar to modern tropical oceans in their temperature and $\delta^{18}$O.
+Our estimate of the temperature at a Silurian paleolatitude of approximately 19$^{\circ}$ south \citep{Torsvik1992} falls within the range of observed temperatures for the modern ocean at a similar latitude \citep{Reynolds1994}.
+The average $\delta^{18}$O$_{\text{VSMOW}}$ of the modern ocean is approximately -0.3$\permil$ \citep{Shackleton1974}, but without ice, the average $\delta^{18}$O$_{\text{VSMOW}}$ would decrease to approximately -1.4$\permil$ \citep{Lhomme2005}.
+The volume of continental glaciers is poorly constrained for the early Silurian, but assuming Silurian ice volume was less than or equal to modern ice volume, our findings suggest that the $\delta^{18}$O of Silurian oceans was within 2$\permil$ of the modern value.
+
+The results of this study are also consistent with those of an earlier clumped isotope study.
+\cite{Came2007} measured the clumped isotope compositions of brachiopods from the early Silurian on Anticosti Island, Canada, and report temperatures of 30 to 35$^{\circ}$C and seawater $\delta^{18}$O$_{\text{VSMOW}}$ values of around -1$\permil$, which are similar to the results of this study.
+Both of these sedimentary basins were bathed in tropical seawater, but each reflects a different inland sea (Anticosti Island on the craton margin of a foreland basin, and Gotland on the southern margin of the Baltic Shield).
+The strong similarity in temperatures and $\delta^{18}$O between these localities lends some confidence to paleoclimate reconstruction during Silurian time.
+
+Our results rule out large ($>$2$\permil$) changes in the $\delta^{18}$O of seawater and extremely hot ($>$40$^{\circ}$C) ocean temperatures as causes of the depleted $\delta^{18}$O values of Silurian carbonates.
+Instead, clumped isotopes demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality.
+This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition.
+We note that it is common to take the average $\delta^{18}$O value of carbonate samples when producing time series records of paleoclimate \citep{Bickert1997, Jaffres2007, Munnecke2003, Shields2003, Veizer2000}.
+If a carbonate sample set has undergone high-temperature fluid-buffered diagenesis, causing more altered samples to have lower $\delta^{18}$O values, the average $\delta^{18}$O of all the samples would not accurately reflect primary ocean conditions.
+The best-preserved sample in our dataset has a $\delta^{18}$O$_{\text{VPDB}}$ of -5.0$\permil$, which is close to the average $\delta^{18}$O$_{\text{VPDB}}$ value (-4.6$\permil$) of compiled low-magnesium calcites from the Silurian Period \citep{Veizer1999}.
+We expect that the same would not be true if clumped isotopes were used to identify the most pristine sample within a sample set from the Ordovician or Cambrian.
+Indeed, the most pristine sample from the upper Mississippi Valley (pre-glacial Ordovician) measured by \cite{Finnegan2011} has a $\delta^{18}$O$_{\text{VPDB}}$ of -4.9$\permil$, while the average of compiled low-magnesium calcites from the Ordovician Period is -5.9$\permil$ \citep{Veizer1999}.
 
@@
 \includegraphics[width=90mm]{Fig_11-01.png}
-\caption{Implications for the constancy of seawater $\delta^{18}$O through time.
-In gray, the $\delta^{18}$O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}.
-The black solid line represents the estimated $\delta^{18}$O$_{\text{VSMOW}}$ of the oceans through time based on the calcite values.
-The two black dashed lines represent alternative hypotheses for the value of seawater $\delta^{18}$O$_{\text{VSMOW}}$ through time.
-All three lines ignore short-term perturbations to seawater $\delta^{18}$O caused by glaciations.
-Finally, the black shapes are estimates of seawater $\delta^{18}$O from clumped isotopes, measured at four different times by three separate studies.
-These clumped isotope estimates suggest that seawater $\delta^{18}$O has not changed by more than 2$\permil$ in the last 430 million years.}
+\caption{Measurements and estimates of seawater $\delta^{18}$O through time.
+In gray, the $\delta^{18}$O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999} and updated in 2004.
+The data are available online: $\mathtt{http://mysite.science.uottawa.ca/jveizer/isotope_data/index.html}$.
+Values from benthic foraminifera are not included because they tend to record much colder water than the other fossils.
+The black solid line represents one hypothesized time series for the $\delta^{18}$O$_{\text{VSMOW}}$ of the oceans through time based on a linear least-squares fit \citep{Veizer2000} through the skeletal calcite data \citep{Veizer1999}.
+The straight dashed black line represents the hypothesis of \cite{Gregory1981} that seawater $\delta^{18}$O$_{\text{VSMOW}}$ has remained constant through time.
+The curved dashed black line represents the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater through time as simulated by the box model in \cite{Wallmann2001}.
+All three black lines ignore short-term perturbations to seawater $\delta^{18}$O caused by glaciations.
+Finally, the black shapes are estimates of seawater $\delta^{18}$O from clumped isotope studies, measured at five different times by four separate studies.
+These clumped isotope estimates suggest that seawater $\delta^{18}$O has not changed by more than 2$\permil$ in the last 430 million years.
+These estimates also highlight that the average $\delta^{18}$O$_{\text{VPDB}}$ in a given interval offers a biased estimate of the seawater composition.}
 \label{Veizer}
@@
 
-Our clumped isotope measurements indicate that Silurian tropical oceans were broadly similar to modern tropical oceans in their temperature and $\delta^{18}$O.
-Our estimate of the temperature at a Silurian paleolatitude of 19$^{\circ}$ south \citep{Torsvik1992} falls within the range of observed temperatures for the modern ocean at a similar latitude \citep{Reynolds1994}.
-The average $\delta^{18}$O$_{\text{VSMOW}}$ of the modern ocean is approximately -0.3$\permil$ \citep{Shackleton1974}, but without ice, the average $\delta^{18}$O$_{\text{VSMOW}}$ would decrease to -1.4$\permil$ \citep{Lhomme2005}.
-The volume of continental glaciers is poorly constrained for the early Silurian, but assuming Silurian ice volume was less than or equal to modern ice volume, our findings suggest that the $\delta^{18}$O of Silurian oceans was within 2$\permil$ of the modern value.
-
-The results of this study are also consistent with those of an earlier clumped isotope study.
-\cite{Came2007} measured the clumped isotope compositions of brachiopods from the early Silurian on Anticosti Island, Canada, and report temperatures of 30 to 35$^{\circ}$C and seawater $\delta^{18}$O$_{\text{VSMOW}}$ values of around -1$\permil$, which are similar to the results of this study.
-Importantly, Anticosti Island recorded deposition on the eastern margin of the Panthalassic Ocean during the Silurian, while Gotland recorded deposition in the Rheic Ocean, which was separated from the Panthalassic Ocean by a mountain range.
-The strong similarity in temperatures and $\delta^{18}$O between the two studies indicates that both studies likely reflect water that is representative of the global ocean, rather than an isolated inland sea.
-
-Our results rule out large ($>2\permil$) changes in the $\delta^{18}$O of seawater and extremely hot ($>50^{\circ}$C) ocean temperatures as causes of the depleted $\delta^{18}$O values of Silurian carbonates.
-Instead, clumped isotopes demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality.
-This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration must be a key contributor to the low $\delta^{18}$O values of Silurian carbonates.
-
-The results from this study join those of previous clumped isotope studies \citep{Came2007, Dennis2013, Finnegan2011} in implying that the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater has been fairly constant throughout the Phanerozoic, at around -1$\pm$1$\permil$ (Figure \ref{Veizer}).
-Several studies have suggested that high-temperature alteration of the seafloor acts as a buffer for the $\delta^{18}$O of seawater \citep{Muehlenbachs1976, Gregory1981, Gregory1991, Muehlenbachs1998}, but a more recent model implied that this buffer may not be strong enough to overcome slow changes in the $\delta^{18}$O of seawater \citep{Jaffres2007}.
+The results from this study join a growing set of clumped isotope studies \citep{Came2007, Dennis2013, Finnegan2011} which suggest that the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater has been fairly constant throughout much of Phanerozoic time, at around -1$\pm$1$\permil$ (Figure \ref{Veizer}).
+Several studies have suggested that high- and low-temperature alteration of the seafloor acts as a buffer for the $\delta^{18}$O of seawater \citep{Gregory1981, Gregory1991, Muehlenbachs1976, Muehlenbachs1998}.
+Observations from modern ocean drill cores suggest that the average temperature of exchange between the seafloor and seawater is 250 to 350$^{\circ}$C \citep{Muehlenbachs1998}.
+At these temperatures, an ocean with a $\delta^{18}$O$_{\text{VSMOW}}$ of $\sim$0$\permil$ is in isotopic equilibrium with the 5.8$\permil$ mantle.
+However, this buffering capacity rests on the assumption that the flux of $^{18}$O exchange due to continental weathering has remained insignificant in the overall $^{18}$O budget throughout Earth’s history, and also that the average temperature of exchange between the seafloor and seawater has not changed through time.
+More recent theoretical efforts challenged these assumptions and suggested that this buffer may not be strong enough to overcome slow changes in the $\delta^{18}$O of seawater \citep{Jaffres2007}.
 If this buffer cannot explain the relative constancy of seawater $\delta^{18}$O values through time, then the two competing controls on seawater $\delta^{18}$O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years.
-This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through an as-yet unidentified feedback loop to balancing changes in the other.
+This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through as-yet unidentified negative feedbacks to balance changes in the other.
 
@@
 We used clumped isotopes to constrain both the temperature and $\delta^{18}$O of Silurian oceans because they provide a temperature that is independent of the $\delta^{18}$O of the water from which the carbonate precipitated.
-Samples were collected from relatively well-preserved sections on the island of Gotland, Sweden.
-By sampling well-preserved, fossil-rich rocks that potentially record a climate phenomenon associated with changing seawater temperatures, we hoped to test the limit of resolution of clumped isotopes.
-We were not able to resolve a change in temperature through the section, due to the variable preservation quality of the samples.
-Trace metal analysis was not able to detect variations in preservation among brachiopods, despite a range in brachiopod temperatures, suggesting that clumped isotopes are a more sensitive indicator of diagenetic alteration than trace metal compositions.
-Micro-scale analysis of calcite fabrics, comparisons between brachiopods and micrite, and the positive relationship between temperature and $\delta^{18}$O of water calculated from the samples all support the hypothesis that the lowest-temperature samples are the most pristine.
-Based on the most pristine sample in our dataset, we estimate that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$.
-Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the $\delta^{18}$O of seawater has remained broadly constant over time, in spite of a large potential for change through exchange of oxygen isotopes with crustal materials.
+Samples were collected from well-preserved flat-lying sections on the island of Gotland, Sweden.
+By sampling well-preserved, fossil-rich rocks that potentially record a climate phenomenon associated with changing seawater temperatures, we tested the ability of clumped isotopes to resolve potentially small climate change phenomena associated with the Ireviken Event.
+Despite the overall high quality of fossiliferous materials in our Gotland sections, we were not able to resolve a secular change in temperature through the section, due to the variable preservation quality of the samples.
+We observed that trace metal concentrations have only a limited ability to detect variations in preservation among brachiopods, despite a range in brachiopod clumped isotope temperatures, suggesting that clumped isotopes are a more sensitive indicator of diagenetic alteration than trace metal compositions.
+This conclusion is supported by EBSD observations.
+Micro-scale analysis of calcite fabrics, comparisons between brachiopods and micrite, and the positive relationship between temperature and $\delta^{18}$O of water calculated from the samples all support the hypothesis that the lowest-temperature samples are the most pristine with alteration of fossil material largely tied to rock-buffered diagenesis.
+Based on the most pristine materials in our dataset, we estimate that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$.
+Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the $\delta^{18}$O of seawater has remained broadly constant over time.
 
@@
 Garrecht Metzger for assistance in field collection.
-We thank Chi Ma for his assistance with the SEM and the E-probe, and Joel Hurowitz for allowing us to use his ICP-AES.
+We thank Chi Ma for his assistance with the SEM and the E-probe, and Joel Hurowitz for help with trace element geochemistry.
 We also thank Mark Garcia and David Mann for their help with sample processing.

### 2019-03-27 Silurian_ocean_temperature/draft.tex
Move files into Silurian_ocean_temperature directory

--- 2013-08-28 draft.tex
@@
-\documentclass[5p, authoryear]{elsarticle}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{rotating}
-\usepackage{multirow}
-\usepackage{enumerate}
-\usepackage{wasysym}
+\documentclass[preprint, authoryear]{elsarticle}
+\usepackage{amsmath} % required for basic math equations
+\usepackage{wasysym} % required for permil symbol
 \usepackage{natbib}
-\usepackage{lineno}
-\usepackage{dblfloatfix}
-\usepackage[skip=2pt,font=footnotesize]{caption}
-\usepackage{breakcites}
-
-\renewcommand{\topfraction}{0.99}
-\renewcommand\bottomfraction{.99}
-\renewcommand{\textfraction}{0.01}
-\renewcommand{\floatpagefraction}{0.99}
-\setcounter{totalnumber}{50}
-\setcounter{topnumber}{50}
-\setcounter{bottomnumber}{50}
 
@@
 
-\title{Carbonate clumped isotope constraints on Silurian ocean temperature \\and seawater $\delta^{18}$O}
+\title{Carbonate clumped isotope constraints on Silurian ocean temperature and seawater $\delta^{18}$O}
 
 \author[caltech]{Renata C. Cummins\corref{cor1}}
-\ead{renatacummins@gmail.com}
+\ead{rcummins@post.harvard.edu}
 
@@
 Much of what we know about the history of Earth's climate derives from the chemistry of carbonate minerals in the sedimentary record.
-The oxygen isotopic composition ($\delta^{18}$O) of calcitic marine fossils and cements has been widely used as a proxy for past seawater temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past.
-Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a calcite phase precipitated, allowing direct calculation of the $\delta^{18}$O of the water as well.
-The clumped isotope composition of calcites is also highly sensitive to and can help diagnose different modes of diagenetic alteration, enabling evaluation of preservation states and identification of the most pristine materials from within a sample set---critical information for evaluating the quality of paleoproxy data generated from carbonates.
+The oxygen isotopic compositions ($\delta^{18}$O) of calcitic marine fossils and cements have been widely used as a proxy for past seawater temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the $\delta^{18}$O of seawater in the past.
+Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a calcite phase precipitated, and allows direct calculation of the $\delta^{18}$O of the water.
+The clumped isotope composition of calcites is also highly sensitive to recrystallization and can help diagnose different modes of diagenetic alteration, enabling evaluation of preservation states and identification of the most pristine materials from within a sample set---critical information for assessing the quality of paleoproxy data generated from carbonates.
 
-We evaluated preservation and measured the clumped isotope composition of a suite of calcitic fossils (primarily brachiopods and corals), sedimentary grains, and cements from Silurian (ca.
+We measured the clumped isotope composition of a large suite of calcitic fossils (primarily brachiopods and corals), sedimentary grains, and cements from Silurian (ca.
 433 Ma) stratigraphic sections on the island of Gotland, Sweden.
 Substantial variability in clumped isotope temperatures suggests differential preservation with alteration largely tied to rock-buffered diagenesis, complicating the generation of a stratigraphically resolved climate history through these sections.
-Despite the generally high preservation quality of samples from these sections, micro-scale observations of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to only a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to that estimated for a ``modern'' ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$).
+Despite the generally high preservation quality of samples from these sections, micro-scale observations of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis suggest that only a subset of relatively pristine samples retain primary clumped isotope signatures.
+These samples indicate that Silurian tropical oceans were likely warm (31$\pm$2$^{\circ}$C) and similar in oxygen isotopic composition to that estimated for a ``modern'' ice-free world ($\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$).
 This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's ocean waters has remained broadly constant through time.
@@
 
-Despite more than half a century of investigations on the topic, the long-term temperature history of Earth's oceans remains poorly constrained.
+Despite more than half a century of study, the long-term temperature history of Earth's oceans remains poorly constrained.
 The most widely used proxy for past ocean temperature is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate $\delta^{18}$O thermometer \citep{Urey1947}.
@@
 The carbonate $\delta^{18}$O proxy has two significant complications.
-First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature.
-This is a fundamental challenge associated with carbonate minerals, which commonly undergo recrystallization during diagenesis and burial, altering primary geochemical signals and impacting the quality of proxy data.
+First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from a primary environmental signature.
+This is an acute challenge associated with carbonate minerals, which commonly undergo recrystallization during diagenesis and burial, altering primary geochemical signals and impacting the quality of proxy data.
 Second, the measured $\delta^{18}$O of carbonates depends not only on the temperature of the water from which the carbonates precipitated, but also on the $\delta^{18}$O of the fluid (seawater and/or pore fluids).
@@
 
-Based on observations of carbonates, it was hypothesized that the $\delta^{18}$O of the ocean increased gradually but substantially through Earth's history---by $\sim$8$\permil$ over Phanerozoic time \citep{Jaffres2007, Veizer1999}.
-On 100-million-year timescales, gradual changes in $\delta^{18}$O are, in principle, plausible given current knowledge of the fluxes impacting the seawater $\delta^{18}$O budget.
+Based largely on observations of carbonates, it was hypothesized that the $\delta^{18}$O of the ocean increased gradually but substantially through Earth's history---by $\sim8\permil$ over Phanerozoic time \citep{Jaffres2007, Veizer1999}.
+On 100-million-year timescales, gradual changes in $\delta^{18}$O are, in principle, possible given current knowledge of the fluxes impacting the seawater $\delta^{18}$O budget.
 The long-term average $\delta^{18}$O of all the water on Earth's surface is largely controlled by the balance between two classes of processes within the silicate Earth: exchange during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange during low-temperature weathering of oceanic and continental crust \citep{Muehlenbachs1998}.
@@
 
-The most widely cited evidence for gradual change in seawater $\delta^{18}$O, however, remains a longstanding controversy in geochemistry.
-This arises from the complications associated with the carbonate $\delta^{18}$O proxy, which allow conflicting interpretations of the observed $\sim$8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}.
+The most widely cited evidence for gradual change in seawater $\delta^{18}$O is an observed ca.
+8$\permil$ increase in the average $\delta^{18}$O of marine carbonates from the Cambrian to today \citep{Jaffres2007, Veizer1999}.
+However, the complications associated with the carbonate $\delta^{18}$O proxy allow conflicting interpretations of this trend in carbonate $\delta^{18}$O, giving rise to a longstanding controversy.
 Several studies have proposed that early Paleozoic carbonates may have low $\delta^{18}$O values because they are more likely to have undergone diagenetic alteration \citep{Clayton1959, Keith1964, Land1995}, which tends to decrease the $\delta^{18}$O of carbonate \citep{Marshall1992}.
@@
 These screened fossils also record increasing $\delta^{18}$O through the Phanerozoic, arguing against diagenetic alteration as a cause of the low $\delta^{18}$O values in the early Paleozoic.
-An alternative hypothesis holds that oceans could have been much warmer in early Paleozoic time \citep{Karhu1986, Knauth1976}.
-Assuming that the $\delta^{18}$O of the oceans was the same as today, Ordovician tropical oceans must have been $\sim$45$^{\circ}$C to produce carbonates with the observed average $\delta^{18}$O$_{\text{VPDB}}$ of -5.9$\permil$ \citep{Veizer1999}.
-Such warm temperatures present an apparent paradox with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading several authors to suggest that the increase in the $\delta^{18}$O of marine carbonates must therefore reflect a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.
+An alternative hypothesis holds that oceans were much warmer in early Paleozoic time \citep{Karhu1986, Knauth1976}.
+Assuming that the $\delta^{18}$O of the oceans was the same as today, Ordovician tropical oceans must have been $\sim$45$^{\circ}$C to produce carbonates with the observed average $\delta^{18}$O$_{\text{VPDB}}$ of -5.9$\permil$.
+Such warm temperature estimates present an apparent paradox with evidence for a large glaciation in the Late Ordovician \citep{Veizer1986}, leading several authors to suggest that the increase in the $\delta^{18}$O of marine carbonates over time must therefore reflect a concomitant increase in the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater from -8$\permil$ in the Cambrian to its present value of $\sim$0$\permil$ \citep{Jaffres2007, Veizer1999}.
 
-Carbonate clumped isotope thermometry supplies a unique insight into this problem because it provides data that can help resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration.
+Carbonate clumped isotope thermometry supplies a unique insight into this problem because it provides data that can help resolve the two main complications of the carbonate $\delta^{18}$O proxy: the clumped isotope composition of a carbonate is independent of the $\delta^{18}$O of water and highly sensitive to diagenetic alteration at temperatures different from those of original deposition.
 The extent to which $^{13}$C and $^{18}$O occur together, or ``clump'' within the same carbonate ion is a function of the temperature at which the carbonate precipitates \citep{Eiler2011}, and isotopic clumping in a wide diversity of shallow-water calcifiers has been shown to record ambient water temperatures \citep{Ghosh2006, Henkes2013, Saenger2012, Tripati2010}.
+There is some uncertainty regarding the uniformity of these calibrations, but all previous studies suggest this phenomenon is a reliable and reasonably precise thermometer when calibrated against carbonates of the same type as analyzed in the same lab.
 Clumped isotope temperature estimates can in turn be used, with the $\delta^{18}$O of shell calcite and the experimentally determined temperature dependence of the oxygen isotope fractionation between water and calcite \citep{Kim1997}, to calculate the $\delta^{18}$O of the water from which the calcite was precipitated.
 
-Shells that have been deeply buried in lithified sediments have often experienced high-temperature alteration, which can change the $\delta^{18}$O of the carbonate and is frequently difficult to detect.
-Clumped isotopes are sensitive to this type of alteration because the altered material will take on a new clumped isotope temperature that can be much higher than the temperature of the pristine fossil calcite.
-If the altered material is incorporated within pristine fossil material selected for sample analysis (e.g.
-as cements lining fossil void space), the clumped isotope temperature of the sample will be significantly higher than the temperature of the ocean water where the organism lived.
-We demonstrate here that carbonate clumped isotopes can provide a more sensitive indicator of diagenetic alteration than the $\delta^{18}$O or the trace metal content of a carbonate.
+Fossils that have been deeply buried in lithified sediments can experience alteration at temperatures greater than those of original deposition.
+This alteration is often associated with recrystallization, which can change the $\delta^{18}$O and trace-metal content of carbonate mineral phases but is frequently difficult to detect.
+Clumped isotopes are sensitive to this type of alteration because the recrystallized carbonate will take on a new clumped isotope temperature that can be much higher than the temperature of the pristine fossil calcite.
+If recrystallized material is sampled along with pristine fossil material during analysis (e.g.
+cements filling void space within a fossil), the clumped isotope temperature of the sample will be significantly higher than the temperature of the seawater where the organism lived.
+We demonstrate here that carbonate clumped isotopes can provide a more sensitive indicator of diagenetic recrystallization than the $\delta^{18}$O or the trace metal content of a given carbonate phase, particularly in rocks that undergo nearly rock-buffered diagenesis.
 
-In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope arrays are also diagnostic of diagenetic regimes, allowing identification of the most pristine end-member of a sample set.
-Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and fluid-buffered diagenesis, respectively) and can also identify fluid-buffered diagenesis in the presence of meteoric water.
-As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water.
+In carbonate strata that have been deeply buried and subjected to high temperatures for extended periods of time, diagenetic alteration can also take the form of closed-system diffusive reordering of atoms within the mineral lattice \citep{Dennis2013, Huntington2011, Quade2013, VanDeVelde2013}.
+During this process, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the measured temperature of the calcite.
+The amount of diffusion that occurs is controlled by both the burial temperature and the length of time for which the sample remains at this temperature.
+The effects of closed-system diffusive reordering can be difficult to recognize, because the reordering does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and re-precipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
+One way to recognize diffusive reordering is through its effects on the distribution of temperatures in a sample suite; diffusive reordering raises the mean temperature without changing the variance in temperature.
+In addition, the degree of diffusive reordering can be constrained by independent indicators of any given sedimentary succession’s time-temperature history, such as the conodont color alteration index (CAI) or rock evaluation parameters of organic matter maturity.
+Recent studies of closed-system diffusive reordering suggest the time-temperature conditions required to increase the clumped isotope temperature of a sample by more than the measurement precision of $\sim$1$^{\circ}$C are more than sufficient to change the color of conodont elements, raising the CAI above the pristine value of one \citep{Dennis2010, Epstein1976, Passey2012, Stolper2014}.
+Therefore, in sample suites with CAI=1 or pre-oil window catagenesis, elevated clumped isotope temperatures can be reasonably interpreted to be the result of diagenetic alteration due to recrystallization.
+
+In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotope data, when combined with bulk O and C isotope analyses of the same materials, can be diagnostic of specific diagenetic processes and regimes (recrystallization vs.
+closed-system diffusive reordering; high vs.
+low water/rock ratio conditions; meteoric vs.
+marine pore fluids), and can be used to identify the most pristine end-member of a sample set \citep{Eiler2011}.
+Of particular interest for this study, as carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water.
 During rock-buffered diagenesis, this causes the $\delta^{18}$O of the water to increase, while the $\delta^{18}$O of the altered carbonate is not affected.
-In a water-buffered diagenetic regime, increasing temperature instead causes the $\delta^{18}$O of the altered carbonate to decrease.
-In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the $\delta^{18}$O of both the carbonate and the water are diagnostically low.
-The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}.
+In contrast, during a water-buffered diagenetic regime, increasing temperature instead causes the $\delta^{18}$O of the altered carbonate to decrease.
+And, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the $\delta^{18}$O of both the carbonate and the water are diagnostically low.
 We demonstrate here that this distinction has important implications for the estimation of the sea surface temperature and $\delta^{18}$O of seawater in the past.
-
-\begin{figure*}[t]
-\centering
-\includegraphics[width=140mm]{Fig_01-01.png}
-\caption{Geologic map of the northern half of the island of Gotland, Sweden, modified from \cite{Calner2004a}.
-Locations of stratigraphic sections are marked and labeled with location number, which corresponds to the sample locations in Supplementary Table 1.}
-\label{map}
-\end{figure*}
 
@@
 
-Samples were collected from several sections exposed on the island of Gotland, Sweden that span the latest Llandovery through Wenlock epochs of the Silurian Period (Figure \ref{map}).
+Samples were collected from several sections exposed on the island of Gotland, Sweden that span the latest Llandovery through Wenlock epochs ($\sim$434 – 430 Ma) of the Silurian Period (Fig.
+1).
 Ages of the formations and correlations between outcrops are constrained by conodont and graptolite biostratigraphy \citep{Jeppsson2006}.
-The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was constrained to be 19$^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.
+The sections record shelf, carbonate platform and backreef lagoon sediments deposited on the margin of the Baltic basin, which was constrained to be $19\pm5.1^{\circ}$ south of the equator during the Silurian Period \citep{Torsvik1992}.
 
-The oldest exposed strata on Gotland comprise the Lower Visby Formation, which consists of marls interbedded with fine-grained limestones deposited below storm wave base (Figure \ref{column}).
+The oldest exposed strata on Gotland comprise the Lower Visby Formation, which consists of marls interbedded with fine-grained limestones deposited below storm wave base (Fig.
+2).
 Fossils are abundant throughout this unit \citep{Calner2004a, Samtleben1996}.
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=140mm]{Fig_02-01.png}
-\caption{Lithology and isotope stratigraphy of the sampled sequence.
-On the left, the stratigraphic column shows the generalized lithostratigraphic units sampled in this study.
-On the right are the carbon and oxygen isotopic compositions of brachiopods, collected from the same sections by \cite{Bickert1997}.
-Isotope data is from \cite{Bickert1997}.}
-\label{column}
-\end{figure*}
-
 Above the Upper Visby Beds, the H\"{o}gklint Formation consists of algal and crinoidal limestones with some reef mounds, deposited above storm wave base \citep{Riding1991, Samtleben1996, Watts2000}.
@@
 
-The Silurian section on Gotland is an attractive location to sample for a clumped isotope study for several reasons.
-
-First, this section has long been recognized for its generally excellent fossil preservation.
+The Silurian strata exposed on Gotland present attractive materials for a clumped isotope study.
+This section has long been recognized for its generally excellent fossil preservation.
 The strata on Gotland are very nearly flat lying, with a dip of less than 1$^{\circ}$, display only minimal evidence of tectonic disturbance and have not been deeply buried \citep{Calner2004a, Jeppsson1983}.
-Conodont fossils from this section yield a conodont alteration index (CAI) of one, which suggests a maximum temperature of $\sim$100$^{\circ}$C \citep{Epstein1976, Jeppsson1983, Wenzel2000}.
-
-The Silurian section on Gotland is also attractive because of the abundance of fossils, especially brachiopods, throughout the section.
-As with previous paleoclimate studies we sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements).
-Many studies have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}.
+Conodont fossils from this section yield a conodont color alteration index (CAI) of one \citep{Epstein1976, Jeppsson1983}.
+The Silurian strata on Gotland also have an abundance of fossils, especially brachiopods, throughout the section.
+We sampled fossil phases for clumped isotope measurements in order to target calcite that was first precipitated directly from seawater (as opposed to carbonate matrix, which may contain authigenic cements).
+Many studies of paleoclimate have targeted the secondary interior fibrous layer of the brachiopod shell, which is thought to precipitate in equilibrium with seawater and to be relatively resistant to diagenetic alteration \citep{Azmy1998, Samtleben2001}.
 The fibrous calcite fabric characteristic of the secondary layer has been well characterized in modern and fossil brachiopods \citep{PerezHuerta2007, Samtleben2001, Schmahl2004}, providing a standard of comparison for the samples in our dataset.
 The trace metal composition of brachiopods has also been extensively studied as an indicator of diagenetic alteration \citep{Azmy1998, Brand2003, Brand2012, Mii1994, Grossman1996, Shields2003}.
-The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often for paleoenvironmental studies than brachiopods, but have been examined by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}.
+The other fossils present in this section, including rugose corals, crinoids, bryozoans, and gastropods, are used less often than brachiopods for paleoenvironmental studies, but have been examined by other clumped isotope studies of paleotemperature \citep{Came2007, Dennis2013, Finnegan2011}.
 Measurements of several fossil taxa allow comparison of the relative preservation of different calcitic materials and can help to detect taxon-specific ``vital effects'' if they are present.
 
-A third attractive feature of the Silurian section on Gotland is the previously observed positive excursion in $\delta^{13}$C that initiates at the boundary between the Lower and Upper Visby Beds (Figure \ref{column}).
+A final attractive feature of the Silurian section on Gotland is the previously observed positive excursion in $\delta^{13}$C that initiates at the boundary between the Lower and Upper Visby Beds (Fig.
+2).
 This excursion, termed the Ireviken event, is now recognized globally \citep{Munnecke2003}.
-In Gotland, a 3$\permil$ excursion in $\delta^{13}$C is coupled to a poorly resolved 0.6$\permil$ positive excursion in $\delta^{18}$O \citep{Munnecke2003}.
-It is somewhat surprising that the excursions coincide with a maximum flooding interval, but it is difficult to distinguish the relative contributions of sediment supply and eustatic sea level rise to this facies change \citep{Calner2004a}.
+In Gotland, a 3$\permil$ excursion in $\delta^{13}$C is coupled to a weakly resolved 0.4$\permil$ positive excursion in $\delta^{18}$O \citep{Munnecke2003}.
+It is somewhat surprising that the positive excursions coincide with a maximum flooding interval, but it is difficult to distinguish the relative contributions of sediment supply and eustatic sea level rise to this facies change \citep{Calner2004a}.
 The excursions are also broadly coincident with minor extinctions in several benthic and pelagic marine taxa \citep{Munnecke2003}.
@@
 The similarity between the isotopic signatures of the Ireviken event and known glaciations, such as the Late Ordovician glaciation, have led several studies to hypothesize a glacial cause for the Ireviken event \citep{Azmy1998, Kaljo2003, Brand2006, Calner2008}.
-This hypothesis was thought to be supported by the presence of glacial tillites in Brazil that were at first interpreted to be early Wenlock in age \citep{Grahn1992}, but these diamictites were later re-interpreted as re-workings of earlier glacial deposits.
-Early Silurian glaciation in South America is now thought to have ended prior to the late Llandovery, before the Ireviken Event \citep{DiazMartinez2007a, DiazMartinez2007b}.
-Many of these observations could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Jeppsson1990, Munnecke2003, Samtleben1996}.
-According to this model, during the humid period recorded by the Lower Visby Beds, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves.
-Increased precipitation would also have caused intense weathering, which could explain the dominance of marls over carbonate reefs during this time.
-During the succeeding arid period beginning in the Upper Visby Beds, evaporation would have increased the $\delta^{18}$O of surface waters, and downwelling of saline water would have brought surface water with high $\delta^{13}$C onto the shelves.
+This hypothesis was supported by the presence of glacial tillites in Brazil that were once interpreted to be early Wenlock in age \citep{Grahn1992}, but these diamictites were later re-interpreted as re-workings of earlier glacial deposits.
+Early Silurian glaciation in South America is currently thought to have ended prior to the late Llandovery, before the Ireviken Event \citep{DiazMartinez2007a, DiazMartinez2007b}.
+Many of the characteristics of the Ireviken Event, although not necessarily its global nature, could also be explained by a shift from a humid period to an arid period in the tropics, which could accompany a glaciation or occur independently \citep{Bickert1997, Jeppsson1990, Munnecke2003, Samtleben1996}.
+According to this model, during the humid period recorded by the Lower Visby Formation, increased precipitation would have lowered the $\delta^{18}$O of surface waters and also stimulated the upwelling of deep water with low $\delta^{13}$C onto the shelves.
+Increased precipitation would have caused intense weathering, which could explain the dominance of marls over carbonate reefs during this time.
+During the succeeding arid period beginning in the Upper Visby Formation, evaporation would have increased the $\delta^{18}$O of surface waters, and downwelling of saline water would have brought surface water with high $\delta^{13}$C onto the shelves.
 Low terrigenous influxes would have allowed the growth of carbonate reefs as observed in the H\"{o}gklint Formation.
@@
 Here we test our ability to resolve possible changes in ocean temperature and/or seawater $\delta^{18}$O through the Ireviken event using stratigraphic samples from Gotland.
-Resolving a change this small with the uncertainties associated with carbonate clumped isotope thermometer is challenging but not impossible, because a shift of 0.6$\permil$ in $\delta^{18}$O corresponds to a change in temperature of approximately 2$^{\circ}$C.
-Though the uncertainties are commonly higher, single measurements of carbonate clumped isotope composition can attain a resolution of $\pm$1$^{\circ}$C, while running multiple replicates of a sample can improve the resolution to $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.
+Resolving a change this small with the uncertainties associated with the carbonate clumped isotope thermometer is challenging because a shift of 0.4$\permil$ in $\delta^{18}$O, if caused by change in temperature of carbonate precipitation at constant $\delta^{18}$O of water, corresponds to a change in temperature of approximately 2$^{\circ}$C.
+Though the uncertainties for a single carbonate clumped isotope temperature are commonly similar to or higher than this, replicate analyses can reduce these errors to $\pm$1$^{\circ}$C or even $\pm$0.4$^{\circ}$C \citep{Thiagarajan2011}.
 
@@
 Stratigraphic sections were measured and samples were collected from the island of Gotland, Sweden during the summer of 2011.
-Each sample was labeled with a location, as shown in Figure \ref{map}, and a stratigraphic height.
-For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby formations (Figure \ref{column}).
-Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}, as well as new bulk carbonate $\delta^{13}$C measurements (Fike et al.
-in prep).
+Each sample was labeled with a geographic location, as shown in Fig.
+1, and a stratigraphic height.
+For locations where the \textit{Phaulactis} layer is exposed, stratigraphic height was expressed relative to the boundary between the Lower and Upper Visby formations (Fig.
+2).
+Other locations were correlated on the basis of prior sequence-, bio- and chemostratigraphic work \citep{Calner2004a}, as well as new bulk carbonate $\delta^{13}$C measurements \citep{Fike2014}.
 
@@
 Brachiopods, bryozoans, crinoids, gastropods and ostracods were sampled by flaking the surface of the skeleton or shell using a dental chisel.
-Every flake was visually inspected under a stereo microscope at circa $25\times$ power.
+Every flake was visually inspected under a stereomicroscope at ca.
+$25\times$ magnification.
 Flakes with visible pyrite grains or rough-textured recrystallized regions were rejected and not included in the material for analysis.
-For brachiopods, only the flakes preserving the characteristic fabric of the fibrous secondary layer were chosen for analysis (see Figure \ref{flakes}).
+For brachiopods, only the flakes preserving the characteristic fabric of the fibrous secondary layer were chosen for analysis (see Fig.
+3).
 Flakes were then sonicated briefly, rinsed, dried in a 50$^{\circ}$C oven, and powdered using a mortar and pestle.
@@
 The apical ends of the corals were targeted because the coral septa are closely spaced together there, leaving less room for sediment and cement between the septa.
-However, this sampling technique is still not as selective as the technique used for brachiopods, and may include more cement and sediment along with skeletal calcite.
-
-\begin{figure}[tb]
-\centering
-\includegraphics[width=90mm]{Fig_03-01.png}
-\caption{Scanning electron microscope images of flakes of brachiopod shell.
-Images were taken in secondary electron mode.
-Panels A--C show flakes that were judged to be pristine and are representative of the material used for analysis.
-Panel D shows a flake that was judged to be partially recrystallized and would not have been chosen for analysis.
-Clumped isotope temperatures of the samples for which representative flakes are shown are: A) 35 $\pm$ 2$^{\circ}$C, B) 33 $\pm$ 2$^{\circ}$C, C) 35 $\pm$ 1$^{\circ}$C.}
-\label{flakes}
-\end{figure}
+However, this sampling technique is still not as selective as the technique used for brachiopods, and can include more cement and sediment along with skeletal calcite.
 
 All fossil samples shown in figures and used to estimate Silurian seawater temperature and $\delta^{18}$O were prepared as described above, with the aim of minimizing the inclusion of cement, sediment and altered material.
-However, in a few separate test cases we purposefully sampled cements, micrites and other textures that are independently known to be more susceptible to diagenetic alteration, as a way to assess possible diagenetic end-members.
+However, in a few separate test cases we purposefully sampled cements, micrite and other textures that are independently known to be more susceptible to diagenetic alteration, as a way to assess possible diagenetic end-members.
 Micrite and cement samples were drilled from cut surfaces and then powdered using a mortar and pestle.
-During the preparation of one brachiopod sample, flakes with large rough-textured recrystallized regions were picked and analyzed separately from the pristine flakes of fibrous secondary layer.
+During the preparation of one brachiopod sample, flakes with large rough-textured recrystallized regions were picked and analyzed separately from the pristine flakes of fibrous secondary layer as a measure of alteration.
 
-While all fossils were inspected under a microscope prior to sampling, we were not able to inspect all fossils in thin section, because carbonate clumped isotope measurements require large amounts of material, some of which is lost during preparation of thin sections.
+While all fossils were inspected under a microscope prior to sampling, we were not able to inspect all fossils in thin section, because carbonate clumped isotope measurements require significant amounts of material, some of which is lost during preparation of thin sections.
 However, 30 $\mu$m-thick uncovered, polished thin sections were obtained from cuts through four representative fossils, two brachiopods and two rugose corals.
@@
 
-Electron backscatter diffraction (EBSD) provides an independent measure of the degree and style of recrystallization by mapping the orientation of calcite crystals in a thin section \citep[e.g.][]{Bergmann2013}.
+Electron backscatter diffraction (EBSD) provides a measure of the degree and style of recrystallization by mapping the orientation of calcite crystals in a thin section \citep[e.g.][]{Bergmann2013}.
 This allows more detailed characterization of calcite fabrics than is possible using optical microscopy, and in the case of brachiopods, comparison to pristine modern examples \citep{PerezHuerta2007}.
-Two thin sections, one rugose coral and one brachiopod, were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under variable pressure vacuum, with an electron high tension of 20 kV.
+Two thin sections, one rugose coral and one brachiopod, were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscope (SEM) under variable pressure vacuum, with an accelerating voltage of 20 kV.
 The stage was tilted 70$^{\circ}$.
 The AZtec program was used in 4x4 binning mode, with frame averaging over one frame and a step size of 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod.
-The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite.
-The diffracted beam then interacts with the EBSD detector, depending on the orientation of the calcite crystal.
 The data were analyzed using software from Oxford instruments.
 
-Flakes of the secondary fibrous layer from twelve different brachiopods were also imaged via electron microscopy to further inspect the preservation of original fibrous calcite texture at a finer scale than available from the reflected light microscope.
+Flakes of the secondary fibrous layer from twelve different brachiopods were also imaged via electron microscopy to define their morphology (but not orientation; i.e., we did not perform EBSD on these flakes).
+This further inspection helped us assess the preservation of original fibrous calcite texture at a finer scale than available from the reflected light microscope.
 The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick.
 The coated flakes were imaged using the secondary electron detector in the same SEM as above, under high vacuum.
-The working distance was $\sim$6 mm and the electron high tension was 10 kV.
+The working distance was $\sim$6 mm and the accelerating voltage was 10 kV.
 
@@
 
-The use of wavelength-dispersive spectroscopy in combination with optical microscopy and EBSD allows characterization of the trace metal contents of specific calcite textures.
+The use of wavelength-dispersive spectroscopy on x-rays emitted during electron microprobe analysis, in combination with optical microscopy and EBSD, allows characterization of the trace metal contents of specific calcite textures.
 By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set.
@@
 Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nA, voltage of 15 kV, and a counting time of 80 milliseconds per pixel.
-Quantitative spot measurements were made using the JEOL quantitative analysis program.
+Quantitative spot measurements were made using the JEOL quantitative analysis program with the same beam conditions as used for the qualitative maps and counting times of 20 seconds on the elemental peak and 10 seconds on background.
 
@@
 
-Although we attempted to sample only well-preserved skeletal material for the carbonate clumped isotope measurements, the large mass of material required makes it very difficult to avoid including small amounts of other phases.
 The trace metal concentrations of bulk samples can be used as an indicator of the amount of sediment and recrystallized material included in the sample \citep{Azmy1998, Brand2012, Came2007, Finnegan2011}.
-Diagenetic recrystallization is thought to enrich calcite in iron (Fe) and manganese (Mn) and deplete strontium (Sr) from calcite \citep{Brand1980, Shields2003}, so we measured the bulk concentrations of these three elements in a large and representative subset of our samples.
+Most studies of the trace elements in fossil calcite measure concentrations of iron (Fe), manganese (Mn), and strontium (Sr), because diagenetic recrystallization is thought to enrich calcite in Fe and Mn and deplete Sr from calcite \citep{Brand1980, Shields2003}.
+We measured the bulk concentrations of these three elements in a large and representative subset of our samples.
 Between 1 and 5 milligrams of each powdered sample were dissolved in a known volume of 4\% nitric acid (HNO$_3$).
@@
 
-Each sample was measured between one and three times depending on the amount of available material.
-For each measurement approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90$^{\circ}$C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate.
-The subsequent purification of evolved CO$_2$ followed procedures outlined by \cite{Ghosh2006}.
+For each carbonate clumped isotope measurement approximately 10 milligrams of powdered sample were dissolved in phosphoric acid at 90$^{\circ}$C, which produces CO$_2$ with a known oxygen isotopic offset from the sample carbonate.
+The subsequent purification of evolved CO$_2$ followed procedures outlined by \cite{Passey2010}.
 Cleaned CO$_2$ was analyzed on a Finnigan MAT253 gas source mass spectrometer configured to collect masses 44 through 49.
@@
 The $\delta^{13}$C, $\delta^{18}$O, $\delta_{45}$, $\delta_{46}$, $\delta_{47}$, and $\delta_{48}$ were calculated for each sample relative to the reference CO$_2$.
-Acquisition-to-acquisition standard deviations were on average 0.03$\permil$ for $\delta_{47}$, 0.04$\permil$ for $\delta^{18}$O, and 0.02$\permil$ for $\delta^{13}$C.
 Raw measured values for $\Delta_{47}$ and $\Delta_{48}$ of each sample, defined as $\Delta_{47-[\text{SGvsWG}]}$ and $\Delta_{48-[\text{SGvsWG}]}$, were also calculated relative to the reference CO$_2$ following equations 3 and 4 in \cite{Huntington2009}.
-
-\begin{figure*}[t]
-\centering
-\includegraphics[width=\textwidth]{Fig_04_color-01.png}
-\caption{The $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of all the samples analyzed in this study, plotted against stratigraphic height.
-For the Lower \& Upper Visby and H\"{o}gklint formations, a height of zero meters is defined by the contact between the Lower Visby Formation and the Upper Visby Formation.
-For the Slite Formation, a height of zero meters is defined as an arbitrary bed near the base of the outcrop.
-Brachiopod $\delta^{13}$C and $\delta^{18}$O measurements from \cite{Munnecke2003} are also included for comparison.
-Error bars in clumped isotope temperature represent the one-sigma standard error of the mean.}
-\label{section}
-\end{figure*}
+Each sample was measured between one and three times depending on the amount of available material.
 
 All $\Delta_{47}$ values are projected into the absolute reference frame \citep{Dennis2011}.
-$\Delta_{47}$ values of two California Institute of Technology intralaboratory standards were converted to the absolute reference frame using a secondary transfer function (STF) constructed from the heated and equilibrated gases measured during the same week as the carbonate standards.
+The absolute reference frame is based on a thermodynamic equilibrium established in two types of CO$_2$ gases: heated gases and water-equilibrated gases.
 Heated gases are aliquots of CO$_2$ that were sealed in quartz tubes, heated at 1000$^{\circ}$C for three hours, and then cooled rapidly to room temperature.
-Equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25$^{\circ}$C for at least 24 hours.
-Heated gases were made from two separate reservoirs of CO$_2$ with different $\delta^{18}$O compositions, and equilibrated gases were equilibrated with either of two separate waters, also differing in their $\delta^{18}$O compositions.
-Heated and equilibrated gases were cleaned following the same procedure as the samples.
-The STF was constructed by plotting the accepted absolute reference frame $\Delta_{47}$ values of 0.0266$\permil$ for heated gases and 0.9252$\permil$ for equilibrated gases against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame \citep{Dennis2011}.
-The STF was used to convert the $\Delta_{47}$ values of the two intralaboratory standards from the \cite{Ghosh2006} reference frame to the absolute reference frame following equation C1 in \cite{Dennis2011}.
+Water-equilibrated gases are aliquots of CO$_2$ that were equilibrated with water at 25$^{\circ}$C for at least 24 hours.
+Heated gases were made from two separate reservoirs of CO$_2$ with different $\delta^{18}$O compositions, and water-equilibrated gases were equilibrated with either of two separate waters, also differing in their $\delta^{18}$O compositions.
+Heated and water-equilibrated gases were cleaned and measured following the same procedure as the samples.
 
-All of the samples in this study were projected into the absolute reference frame using a separate STF constructed for each week of measurements.
-To create the STF, the accepted $\Delta_{47}$ values for the two intralaboratory standards and heated gases were plotted against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
-Equilibrated gases were measured during some but not all weeks of sample measurements, so the accepted and observed $\Delta_{47}$ values of equilibrated gases were included in the weekly STF during the weeks when these measurements were available.
-$\Delta_{47}$ values of the samples in the \cite{Ghosh2006} reference frame were corrected using this STF following equation C1 in \cite{Dennis2011}.
+Before sample $\Delta_{47}$ values could be projected into the absolute reference frame, we first needed to establish accepted values in the absolute reference frame for two California Institute of Technology intralaboratory carbonate standards.
+$\Delta_{47}$ values of the two standards were converted to the absolute reference frame using only data from weeks during which heated gases and water-equilibrated gases were measured along with the standards.
+For each of these weeks, a transfer function was constructed by plotting the accepted absolute reference frame $\Delta_{47}$ values of 0.0266$\permil$ for heated gases and 0.9252$\permil$ for 25$^{\circ}$C water-equilibrated gases against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame \citep{Dennis2011}.
+This transfer function was used to convert the $\Delta_{47}$ values of the two intralaboratory carbonate standards from the \cite{Ghosh2006} reference frame to the absolute reference frame following equation C1 in \cite{Dennis2011}.
+
+The $\Delta_{47}$ values of all of the samples in this study were projected into the absolute reference frame using a separate transfer function constructed for each week of measurements.
+For weeks when water-equilibrated gases were measured, this transfer function was constructed by plotting the accepted absolute reference fame $\Delta_{47}$ values of the two intralaboratory carbonate standards, heated gases, and 25$^{\circ}$C water-equilibrated gases against their observed $\Delta_{47}$ as determined in the \cite{Ghosh2006} reference frame.
+For weeks when water-equilibrated gases were not measured, this transfer function included only measurements of the two intralaboratory carbonate standards and heated gases.
+$\Delta_{47}$ values of the samples in the \cite{Ghosh2006} reference frame were corrected using this transfer function following equation C1 in \cite{Dennis2011}.
 The resulting $\Delta_{47-\text{RF}}$ was corrected for the acid extraction at 90$^{\circ}$C by adding 0.092$\permil$ \citep{Henkes2013} to produce the final value ($\Delta_{47-\text{RF-AC}}$).
@@
 
+One-sigma standard errors based on all acquisitions in each measurement range from 0.0007$\permil$ to 0.0065$\permil$ (averaging 0.0017$\permil$) for $\delta^{13}$C, from 0.0008$\permil$ to 0.0069$\permil$ (averaging 0.003$\permil$) for $\delta^{18}$O, and from 0.005$\permil$ to 0.019$\permil$ (averaging 0.012$\permil$) for $\Delta_{47}$.
+These errors in $\Delta_{47}$ are consistent with counting-statistics limits of 0.013--0.016$\permil$ \citep{Huntington2009}.
+The errors in $\Delta_{47}$ propagate into temperature errors ranging from $\pm$1.2$^{\circ}$C to $\pm$4.4$^{\circ}$C, with an average of $\pm$2.8$^{\circ}$C.
+
 \section{Results and discussion}
@@
 
-The $\delta^{13}$C values of fossil material, micrite and cement are highly consistent within beds (Figure \ref{section}).
-We observe a 3$\permil$ positive excursion in $\delta^{13}$C, beginning just below the boundary between the Lower and Upper Visby formations, and continuing for $>$50 meters.
-The sample set does not capture the relaxation of this excursion, but samples from the Slite Formation all have consistently lower $\delta^{13}$C$_{\text{VPDB}}$ values of approximately -0.5$\permil$.
+The $\delta^{13}$C values of fossil material, micrite and cement are highly consistent within beds (Fig.
+4).
+We observe a $\sim$3.5$\permil$ positive excursion in $\delta^{13}$C, beginning just below the boundary between the Lower and Upper Visby formations, and continuing for $>$50 meters.
+Our samples do not capture the relaxation of this excursion, but samples from the Slite Formation all have consistently lower $\delta^{13}$C$_{\text{VPDB}}$ values of approximately -0.5$\permil$.
 The $\delta^{13}$C trend through the section is similar to the brachiopod $\delta^{13}$C measurements from \cite{Bickert1997} and \cite{Munnecke2003}.
 
-The $\delta^{18}$O values of fossil material are consistent within beds, but micrite and cements tend to have lower $\delta^{18}$O values than fossil material, often several permil lower than fossils in the same bed (Figure \ref{section}).
+The $\delta^{18}$O values of fossil material are consistent within beds, but micrite and cements tend to have lower $\delta^{18}$O values than fossil material, often several permil lower than fossils in the same bed (Fig.
+4).
 It is interesting that we were not able to resolve an excursion in carbonate $\delta^{18}$O.
 This may simply be because our dataset contains fewer samples than that of \cite{Munnecke2003}, although even with the additional samples, the previously reported excursion is poorly resolved.
-No $\delta^{18}$O excursion was observed in the isotopic data generated from bulk sediment (Fike et al.
-in prep.).
-The average $\delta^{18}$O$_{\text{VPDB}}$ values of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby Beds, making our Lower Visby samples extremely similar to those of \cite{Munnecke2003}, but our Upper Visby samples slightly more depleted in $^{18}$O.
+No $\delta^{18}$O excursion was observed in the isotopic data generated from bulk sediment collected from the same sections \citep{Fike2014}.
+The average $\delta^{18}$O$_{\text{VPDB}}$ value of the fossils remains fairly constant at -5.5$\permil$ through the Lower and Upper Visby formations.
+These values are extremely similar to those of \cite{Munnecke2003} for the Lower Visby Formation, but our data only overlap with the lower end of the $\delta^{18}$O$_{\text{VPDB}}$ values from \cite{Munnecke2003} for the Upper Visby Formation.
 The low $\delta^{18}$O values of Upper Visby fossils may be caused by increased contamination from secondary cements, which could be expected to accompany the transition from marl to limestone.
@@
 
-Clumped isotope temperatures of samples are highly variable within beds, often covering a range of 10$^{\circ}$C within a single bed (Figure \ref{section}).
-The lowest temperature sample, a brachiopod, is from the Slite Formation, deposited well after the hypothesized climate changes associated with the isotopic excursions in the Lower and Upper Visby formations.
-The large, within-bed scatter in clumped isotope temperatures is unlikely to be primary and suggests variable preservation quality among these samples.
-Even the secondary fibrous shell layer of brachiopods, which is commonly thought to be relatively resistant to diagenetic alteration, displays a range in temperature of more than 15$^{\circ}$C.
+Clumped isotope temperatures of samples are highly variable both between and within beds.
+Of the eight beds for which we measured more than one fossil sample, four beds have temperature ranges of more than 13$^{\circ}$C (Fig.
+4).
+Assuming an error of 5$^{\circ}$C to account for environmental variability and measurement error, we used order statistics to calculate that there is a probability of less than 1\% that one sample in these beds will have a temperature more than 13$^{\circ}$C higher than the true depositional temperature, if all samples record a primary temperature.
+Therefore, the within-bed scatter in clumped isotope temperatures suggests variable preservation quality among these samples.
+Even the secondary fibrous shell layer of brachiopods, which is commonly thought to be relatively resistant to diagenetic alteration, displays a range in temperature of more than 15$^{\circ}$C across the whole dataset.
 The variable preservation quality of these samples means that beds with fewer samples are less likely to contain a pristine fossil with a reliable temperature---introducing a sample bias.
-Because many of the beds in this study are only represented by a few samples, we are not able to stratigraphically resolve what we would interpret as a change in temperature with time.
+Because many of the beds in this study are only represented by a few samples, we are not able to stratigraphically resolve what we would interpret robustly as a change in seawater temperature with time.
 
-\begin{figure*}[b]
-\centering
-\includegraphics[width=\textwidth]{Fig_06_color-01.png}
-\captionsetup{labelformat=empty} 
-\caption{Figure \ref{metals}: Scatter plots showing the relationship between temperature and bulk trace metal concentrations.
-The negative correlation between temperature and the log of Sr concentration in rugose corals is weakly significant and strengthens when combined with the rest of the samples.
-There are also weakly significant positive correlations between temperature and the log of Fe concentration and between temperature and the log of Mn concentration among all the samples combined.
-These plots demonstrate clear differences in trace metal content between the two fossil taxa and the later carbonate phases.
-Brachiopods have the lowest average temperature of all the sample subsets, and also tend to have the most pristine trace metal signatures, with less Fe and Mn than the other samples.
-Micrites and cements have the highest average temperature and tend to have the most altered trace metal signatures, with high Fe and Mn concentrations.}
-\label{Brand}
-\end{figure*}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=90mm]{Fig_05_color-01.png}
-\captionsetup{labelformat=empty} 
-\caption{Figure \ref{Brand}: Bulk manganese and strontium concentrations of the brachiopods, rugose corals, micrites and cements in this study.
-The brachiopods in this study have more Sr on average than the brachiopods analyzed by \cite{Brand2012}, which are also shown in this figure.
-Lines represent limits suggested by different authors for the range of pristine concentrations defined for brachiopods.
-Rugose corals generally have more Fe (not shown) and more Mn than brachiopods, giving them more altered trace metal signatures, while cements and micrites tend to have the most highly altered trace metal signatures.
-Figure modified from \cite{Brand2012}.}
-\label{metals}
-\end{figure}
+\subsection{Preliminary considerations of the diagenetic regime}
 
 Although we are not able to measure a change in sea surface temperature through the Ireviken Event, we can use our data to constrain more broadly temperature and $\delta^{18}$O of seawater by uncovering the mechanism of diagenetic alteration to identify the most pristine samples.
-All of the fossil samples have fairly high $\delta^{18}$O values, making low-temperature alteration by meteoric water unlikely.
-This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent diagenetic alteration during burial at elevated temperatures.
-These higher clumped isotope temperatures could be caused by incorporation of late-stage cements or wholesale recrystallization.
-We tested this hypothesis using independent methods, described above, to assess the preservation quality of representative samples.
+As described above, diagenetic regimes can vary in their fluid composition (meteoric vs.
+marine pore fluids) and in their mechanism (closed-system diffusive reordering vs.
+recrystallization and incorporation of late-stage cements).
+We suggest that alteration likely occurred in the presence of marine pore fluids rather than meteoric fluids, because all of the fossil samples have fairly high $\delta^{18}$O values, making low-temperature alteration by meteoric water unlikely.
+We also argue that closed-system diffusive reordering did not strongly affect any of our samples.
+Gotland strata have a conodont color alteration index of one \citep{Epstein1976, Jeppsson1983}, which constrains the possible effects of diffusive reordering to an increase in the clumped isotope temperature by a maximum of $\sim$1$^{\circ}$C \citep{Dennis2010, Passey2012, Stolper2014}.
+Closed-system diffusive reordering is also incompatible with the observed variability in clumped isotope temperatures because this process is expected to preserve the small variance resulting from environmental variability and measurement uncertainty (Fig.
+5).
+The variance of our sample set is much larger than the maximum measurement uncertainty of 4.4$^{\circ}$C, which is consistent with recrystallization as the dominant diagenetic process.
+This leads us to hypothesize that the lowest temperature samples are the most pristine, while the higher temperature samples likely represent recrystallization and incorporation of late-stage cements during burial at elevated temperatures.
+
+Comparisons of the clumped isotope temperature of different calcite textures within samples help diagnose the diagenetic regime that acted to increase the apparent temperature of more altered samples.
+We measured the clumped isotope composition of micrites adjacent to four brachiopods, and in all four cases, the temperature of the micrite was higher than the temperature of the neighboring brachiopod (Fig.
+6).
+Micrite is expected to be less pristine than the brachiopod samples because it has high surface areas, far more commonly recrystallizes during burial, and often includes several generations of cements.
+Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
+Interestingly, while the $\delta^{13}$C values of micrites are similar to those of adjacent brachiopods, three of the four micrites tested are depleted in $\delta^{18}$O relative to the adjacent brachiopods, suggesting that the micrites reflect recrystallization with a higher water-to-rock ratio.
+
+In a similar experiment, recrystallized skeletal material was also shown to be associated with higher clumped isotope temperatures.
+One brachiopod sample was split into two fractions during picking.
+The first fraction consisted of pristine flakes of fibrous secondary layer and had a temperature of 38$\pm$3$^{\circ}$C, while the other fraction was comprised of flakes with large rough-textured recrystallized regions and had a temperature of 43$\pm$3$^{\circ}$C, where the error estimates represent the one-sigma standard error of the mean over all acquisitions.
+The results of this experiment support for the hypothesis that the lowest-temperature samples are the most pristine.
 
@@
 
+We tested the hypothesis that the lowest-temperature samples are the most pristine using independent methods, described above, to assess the preservation quality of representative samples.
 Bulk trace metal measurements, which have been widely used as indicators of diagenetic alteration \citep{Azmy1998, Brand2012, Came2007, Finnegan2011}, suggest that all the brachiopods in this study are comparatively pristine.
@@
 Then we consider EBSD and SEM evidence for the inclusion of diagenetic material in the typical fossil sample.
-Finally, we turn to clumped isotopes to distinguish the best-preserved samples from among those that would ordinarily all be identified as pristine based on their trace metal contents.
 
@@
 Diagenetic recrystallization is also thought to deplete Sr because brachiopod skeletons concentrate Sr relative to its abundance in seawater and meteoric waters \citep{Brand1980, DePaolo2011, Schrag1995}.
-Many previous studies have measured the concentrations of these three elements in fossil brachiopods, and some have defined limits for the Mn and Sr concentrations of brachiopods that can be considered pristine.
-The bulk Mn and Sr concentrations of all brachiopods in this study fall within these proposed limits (Figure \ref{Brand}), the small variations in trace metal concentrations among the brachiopods would not ordinarily be considered as evidence for variable preservation.
+Previous studies have defined limits for the Mn and Sr concentrations of brachiopods that can be considered pristine \citep{Azmy1998, vanGeldern2006, Korte2005, Mii1999}.
+The bulk Mn and Sr concentrations of all brachiopods in this study fall within these proposed limits (Fig.
+7); the small variations in trace metal concentrations among the brachiopods would not ordinarily be interpreted as variable preservation.
 
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=140mm]{Fig_07_color-01.png}
-\caption{Calcite fabric and trace metal concentrations in a rugose coral.
-The four images show the same area of a rugose coral thin section, all at the same magnification.
-The black bar at the bottom of panel B is 500 $\mu$m long for scale.
-A) Transmitted light.
-The dark triangle labeled ``RC'' was identified as a recrystallized domain of the coral septum.
-Cement in the void space between coral septa appears as blocky calcite crystals.
-B) Electron backscatter diffraction (EBSD) analysis.
-The orientation of calcite crystals is indicated by different shades.
-The recrystallized domain is clearly marked by many calcite crystals with the same orientation.
-C) Iron and D) manganese concentrations measured using the E-probe.
-The maps are qualitative, with darker shades representing higher concentrations.
-For both metals, the recrystallized domain is indistinguishable from the rest of the septum.
-The black lines show the location of an array of 62 spots where trace metal concentrations were measured quantitatively.
-The concentration of Fe and Mn at these spots is plotted to the right of the respective map.}
-\label{EBSD_coral}
-\end{figure*}
-
-Under the hypothesis that the lowest-temperature samples are the most pristine, we also expected to find a correlation between clumped isotope temperatures and trace metal evidence for diagenetic alteration, as observed for similar-aged fossils by \cite{Finnegan2011}.
-We tested brachiopods, rugose corals, and micrites/cements separately and combined for the robustness of a linear correlation between temperature and the log of trace metal concentrations (Figure \ref{metals}).
+We also expected to find a correlation between clumped isotope temperatures and trace metal evidence for diagenetic alteration, as observed for fossils by \cite{Finnegan2011}.
+We tested brachiopods, rugose corals, and micrites/cements separately and combined for the robustness of a linear correlation between temperature and the log of trace metal concentrations (Fig.
+8).
 We performed a hypothesis test on each pair of variables to test the probability (\textit{p}) that the observed correlation is due to random variation when the true correlation is zero.
@@
 There are weakly significant positive correlations between temperature and Fe concentration and also between temperature and Mn concentration when all samples are combined.
-These correlations are largely due to differences between fossil taxa and later phases, brachiopods have the lowest average temperature, Fe, and Mn concentrations, while micrites and cements combined have the highest average temperature, Fe and Mn concentrations.
+These correlations are largely due to differences between fossil taxa and later phases; brachiopods have the lowest average temperature, Fe, and Mn concentrations, while micrites and cements combined have the highest average temperature, Fe and Mn concentrations.
 There is also a weakly significant negative correlation between temperature and Sr concentration among rugose corals, which strengthens (to 99\% confidence) when corals are combined with brachiopods, and with micrites and cements.
-Trace metal concentrations reflect broad differences between taxa and textures, however none of the other variable pairs are significantly correlated within individual taxa (Supplementary Table 2).
+Trace metal concentrations reflect broad differences between taxa and textures, however none of the other variable pairs are significantly correlated within individual taxa (Supplementary table 2).
 These tests may be complicated by contamination from early void-filling cements, which have low temperatures but ``altered'' trace metal signatures because they precipitate from pore fluids.
@@
 
-\begin{figure*}[tb]
-\centering
-\includegraphics[width=\textwidth]{Fig_08_color-01.png}
-\caption{Calcite fabric in a brachiopod.
-A) Transmitted polarized light image of a thin section of brachiopod G3 LVF B1.
-This sample has a clumped isotope temperature of 35$^{\circ}$C.
-The boxed overlay is an electron backscatter diffraction (EBSD) image of the same sample.
-The orientation of calcite crystals is indicated by different shades.
-B) Euler pole figure generated from the EBSD analysis.
-The central band represents the preferential orientation of calcite crystals parallel to the outer surface of the brachiopod shell.}
-\label{EBSD_brach}
-\end{figure*}
-
-There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration.
+There are several possible explanations for the relative insensitivity of trace metal concentrations to diagenetic alteration of these materials.
 Our sample set likely does not include a sufficient number of highly altered samples to capture the full range in severity of alteration, because visibly altered fossils were rejected during visual screening of samples and not included in the sample set.
 Variability in primary trace metal concentrations, which is large among modern brachiopods \citep{Brand2003}, may also mask any trends superimposed by diagenetic alteration.
-However, E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization may not strongly affect Fe and Mn concentrations.
-A recrystallized region of a rugose coral septum is clearly visible as a dark area in transmitted light (Figure \ref{EBSD_coral}A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Figure \ref{EBSD_coral}B).
-However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Figure \ref{EBSD_coral}C \& D).
-This is only one example, but it highlights relatively large calcite domains within samples that were recrystallized but this is not reflected in the trace metal concentrations.
-The trace metal signature associated with recrystallization likely depends on conditions during recrystallization, suggesting that this coral underwent closed-system diagenesis.
+However, E-probe analysis suggests a more compelling reason for the insensitivity of trace metal concentrations: diagenetic recrystallization need not strongly affect Fe and Mn concentrations.
+A recrystallized region of a rugose coral septum is clearly visible as a dark area in transmitted light (Fig.
+9A) and an area of aligned calcite crystals in an electron backscatter diffraction (EBSD) image (Fig.
+9B).
+However, the Fe and Mn concentrations of the recrystallized region are indistinguishable from those of the pristine coral skeleton (Fig.
+9C \& D).
+This is only one example, but it highlights relatively large calcite domains within samples that were recrystallized but not strongly modified in trace metal concentrations.
+The trace metal signature associated with recrystallization likely depends on conditions during recrystallization, suggesting that this coral underwent closed-system diagenesis, or recrystallization in the presence of trace-metal-poor waters.
 If closed-system recrystallization is an important contributor to the elevated temperatures of fossil material in our sample set, trace metals may not be expected to be reliable indicators of diagenetic alteration.
@@
 We therefore turn to micro-scale studies of calcite textures as indicators of diagenetic alteration.
-Reflected-light and EBSD images of rugose coral fossil skeletons reveal substantial domains of late-phase diagenetic calcite, including recrystallized regions and cements (Figure \ref{EBSD_coral}A \& B).
+Reflected-light and EBSD images of rugose coral fossil skeletons reveal substantial domains of late-phase diagenetic calcite, including recrystallized regions and cements (Fig.
+9A \& B).
 As discussed above, recrystallized regions are visible as dark colors in reflected light and aligned calcite crystals in EBSD images.
-Cements in Figure \ref{EBSD_coral}A \& B are characterized by large, subequant, interlocking calcite crystals.
+Cements in Fig.
+9A \& B are characterized by large, subequant, interlocking calcite crystals.
 
-In contrast to the rugose corals, the EBSD image of a low-temperature (35$^{\circ}$C) brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Figure \ref{EBSD_brach}A).
+In contrast to the rugose corals, the EBSD image of a low-temperature (35$^{\circ}$C) brachiopod thin section demonstrates excellent preservation of original fibrous calcite textures (Fig.
+10A).
 Within the brachiopod shell, calcite crystals are elongated approximately parallel to the surface of the shell, as expected for the fibrous secondary layer.
-This preferential orientation is represented by the vertical band in the stereographic projection of crystallographic planes (Figure \ref{EBSD_brach}B).
+This preferential orientation is represented by the vertical band in the stereographic projection of crystallographic planes (Fig.
+10B).
 No large regions of recrystallization are visible within the shell.
@@
 
-Scanning electron microscope (SEM) images of low-temperature brachiopod flakes also demonstrate excellent preservation of original fibrous calcite textures (Figure \ref{flakes}A, B \& C).
-The original fibrous texture clearly contrasts with the rough recrystallized texture covering part of a flake that would not have been chosen for analysis (Figure \ref{flakes}D).
+Scanning electron microscope (SEM) images of low-temperature brachiopod flakes also demonstrate excellent preservation of original fibrous calcite textures (Fig.
+3A, B \& C).
+The original fibrous texture clearly contrasts with the rough recrystallized texture covering part of a flake that would not have been chosen for analysis (Fig.
+3D).
 The difference between these two textures is visible under the microscope used for picking brachiopod flakes, so flakes with large recrystallized regions were easily identified and excluded from the sample.
-However, small recrystallized regions, especially if they occurred between layers of fibrous texture, may not have been identified during picking.
+However, small recrystallized regions, especially if they occurred between layers of fibrous texture, may elude identification during picking.
 Therefore, although the representative flakes chosen for inspection with the SEM appear pristine, this does not guarantee that all recrystallized material was excluded from the samples.
 
-The micro-scale studies of calcite textures in Figures  \ref{flakes}, \ref{EBSD_coral} and \ref{EBSD_brach} reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite.
+The micro-scale studies of calcite textures in Figs.
+3, 9 and 10 reveal relatively well-preserved samples, but do not rule out the inclusion of small amounts of late-phase calcite.
 Variable fractions of late-phase calcite likely contribute to the range in temperature among fossil samples.
-Due to constraints on the amount of available material for analysis, we only performed SEM and EBSD analysis on a small subset of samples, but an important conclusion that can be drawn from the electron microscopy and spectroscopy results is that the low temperature brachiopods imaged in Figures \ref{flakes} and \ref{EBSD_brach} probably contain at most only trace amounts of late-phase calcite, variable but likely less than 5\% of the sample.
+Due to constraints on the amount of available material for analysis, we only performed SEM and EBSD analysis on a small subset of samples, but an important conclusion that can be drawn from the electron microscopy and spectroscopy results is that the low temperature brachiopods imaged in Figs.
+3 and 10 probably contain at most only trace amounts of late-phase calcite, variable but likely less than 5\% of the sample.
 We therefore look to where these samples fall on diagenetic trajectories to estimate Silurian ocean temperature and the value of seawater $\delta^{18}$O.
 
-\subsection{Constraining the temperature and $\mathbf{\delta^{18}}$O of Silurian oceans}
-
-\begin{figure}[b]
-\centering
-\includegraphics[width=90mm]{Fig_09-01.png}
-\caption{The clumped isotope temperature and calculated water $\delta^{18}$O$_{\text{VSMOW}}$ for all samples analyzed in this study.
-The contours represent lines of constant carbonate $\delta^{18}$O$_{\text{VPDB}}$.
-We approximate the correlated error as symmetric using ellipses that represent one standard deviation of the mean of all measurement acquisitions for each sample.
-The positive correlation between temperature and water $\delta^{18}$O$_{\text{VSMOW}}$ is diagnostic of rock-buffered diagenesis.
-Under this regime, the lowest-temperature samples are the most pristine.
-Note that some cement and micrite samples fall off the trend, suggesting the effects of a more open system diagenetic process.}
-\label{ellipses}
-\end{figure}
+\subsection{Constraining the temperature and $\mathit{\delta^{18}}\textit{O}$ of Silurian oceans}
 
@@
 As outlined above, different diagenetic processes have characteristic effects on the relationship between temperature and the $\delta^{18}$O of interstitial water.
-When the clumped isotope temperatures of the samples in this study are plotted versus the $\delta^{18}$O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}).
+When the clumped isotope temperatures of the samples in this study are plotted versus the $\delta^{18}$O of water calculated for each sample, there is a clear positive relationship (Fig.
+11).
 This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is characteristic of rock-buffered diagenesis.
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=90mm]{Fig_10-01.png}
-\caption{Comparisons between the $\delta^{13}$C, $\delta^{18}$O and clumped isotope temperature of brachiopod samples and adjacent micrite.
-Micrite is more easily altered than brachiopod shell material, so the consistently higher temperature of micrite supports the hypothesis that the lowest temperature samples comprise the more pristine end of the diagenetic trajectory shown in Figure \ref{ellipses}.
-The generally depleted $\delta^{18}$O values of micrites compared to brachiopods suggest some influence from fluid-buffered diagenesis.}
-\label{comparisons}
-\end{figure}
-
-Additional evidence for a diagenetic regime that acted to increase the apparent temperature of more altered samples is provided by comparisons of the clumped isotope temperature of different calcite textures within samples.
-We measured the clumped isotope composition of micrites adjacent to four brachiopods, and in all four cases, the temperature of the micrite was higher than the temperature of the neighboring brachiopod (Figure \ref{comparisons}).
-Micrite is expected to be less pristine than the brachiopod samples because it far more commonly recrystallizes during burial and can include different generations of cements.
-Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.
-Interestingly, while the $\delta^{13}$C values of micrites are similar to those of adjacent brachiopods, three of the four micrites tested are depleted in $\delta^{18}$O relative to the adjacent brachiopods, suggesting that the micrites may have undergone diagenesis in the presence of a higher water-to-rock ratio.
-
-In a similar experiment, recrystallized skeletal material was also shown to be associated with higher clumped isotope temperatures.
-One brachiopod sample was split into two fractions during picking.
-The first fraction consisted of pristine flakes of fibrous secondary layer and had a temperature of 38$\pm$3$^{\circ}$C, while the other fraction was comprised of flakes with large rough-textured recrystallized regions and had a temperature of 43$\pm$3$^{\circ}$C, where the error estimates represent the one-sigma standard error of the mean over all acquisitions.
-The results of this experiment provide further support for the hypothesis that the lowest-temperature samples are the most pristine, and also demonstrate the importance of careful picking when sampling brachiopods.
-
-Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis \citep{Dennis2013, Huntington2011, Quade2013, VanDeVelde2013}, but we suggest that solid-state diffusion is not a major contributor to our results.
-During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite.
-This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and re-precipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}.
-We argue that solid-state diffusion did not strongly affect any of our samples because Gotland strata were exposed to a maximum burial temperature of only 100$^{\circ}$C, as constrained by the conodont alteration index \citep{Jeppsson1983, Wenzel2000}, Temperatures above 100$^{\circ}$C are thought to be required for solid-state diffusion to increase the clumped isotope temperature of a calcite \citep{Dennis2010}.
-We cannot rule out a minor influence from solid-state diffusion, but due to the shallow burial history and low conodont alteration index of Gotland strata, we are confident that the lowest-temperature, most pristine samples in our dataset record a temperature that is very close to the primary seawater temperature.
-
-The lowest-temperature sample within our sample set, a brachiopod (G13 10.5 to 10.8 B1, Slite Fm.), suggests that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$ (Figure \ref{ellipses}).
-The two next-most pristine samples, also brachiopods (G11 0.2 to 0.4 B6, stratigraphic height 1.3 m, G2 0.5 to 0.6 B2, stratigraphic height 0.5 m), have temperatures of 33$\pm$2$^{\circ}$C and 34$\pm$1$^{\circ}$C.
+The lowest-temperature sample within our sample set, a brachiopod (G13 10.5 to 10.8 B1, Slite Fm.), suggests that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$ (Fig.
+11).
+The two next-most pristine samples, also brachiopods (G11 0.2 to 0.4 B6, stratigraphic height 1.3 m; G2 0.5 to 0.6 B2, stratigraphic height 0.5 m), have temperatures of 33$\pm$2$^{\circ}$C and 34$\pm$1$^{\circ}$C.
 Together with the $\delta^{18}$O$_{\text{VPDB}}$ of each brachiopod, these temperatures can be used to calculate seawater $\delta^{18}$O$_{\text{VSMOW}}$ values of -2.2$\pm$0.4$\permil$ and -1.2$\pm$0.2$\permil$, respectively.
@@
 
-Due to the increase of observed sample temperatures during diagenesis, our estimates of the temperature and $\delta^{18}$O of Silurian tropical oceans are technically upper limits.
+Due to the increase of observed sample temperatures during diagenesis, our estimates of the temperature and $\delta^{18}$O of Silurian tropical oceans are upper limits.
 We are not able to exclude the possibility that small amounts of diagenesis affected all the samples in our study, even the most pristine samples.
@@
 Assuming closed-system diagenesis so that the late-phase calcite is similar in bulk $\delta^{13}$C and $\delta^{18}$O to the pristine fossil calcite, $\Delta_{47}$ values will mix linearly \citep{Affek2006}.
-In this case, if the measured temperature of the sample is 31$^{\circ}$C and 5\% of the sample material was recrystallized at 100$^{\circ}$C, the primary temperature of the pristine fossil material would be 28.3$^{\circ}$C.
-Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 3$^{\circ}$C.
+In this case, if the measured temperature of the sample is 31$^{\circ}$C and 5\% of the sample material was recrystallized at a maximum of 80$^{\circ}$C (constrained by CAI=1), the primary temperature of the pristine fossil material would be 29$^{\circ}$C.
+Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 2$^{\circ}$C---similar to our measurement uncertainty.
 
-Our clumped isotope measurements indicate that Silurian tropical oceans were broadly similar to modern tropical oceans in their temperature and $\delta^{18}$O.
-Our estimate of the temperature at a Silurian paleolatitude of approximately 19$^{\circ}$ south \citep{Torsvik1992} falls within the range of observed temperatures for the modern ocean at a similar latitude \citep{Reynolds1994}.
+These clumped isotope results indicate that Silurian tropical oceans were broadly similar to modern tropical oceans in their temperature and $\delta^{18}$O.
+Our estimate of the temperature at a Silurian paleolatitude of $19\pm5.1^{\circ}$ south \citep{Torsvik1992} is within 1$^{\circ}$C of the maximum observed temperature for the modern ocean at a similar latitude \citep{Reynolds1994}.
 The average $\delta^{18}$O$_{\text{VSMOW}}$ of the modern ocean is approximately -0.3$\permil$ \citep{Shackleton1974}, but without ice, the average $\delta^{18}$O$_{\text{VSMOW}}$ would decrease to approximately -1.4$\permil$ \citep{Lhomme2005}.
-The volume of continental glaciers is poorly constrained for the early Silurian, but assuming Silurian ice volume was less than or equal to modern ice volume, our findings suggest that the $\delta^{18}$O of Silurian oceans was within 2$\permil$ of the modern value.
+As stated above, the volume of continental ice is not known with certainty for all of early Silurian time, but assuming Silurian ice volume was less than or equal to modern ice volume, our findings suggest that the $\delta^{18}$O of Silurian oceans was within 2$\permil$ of the modern value.
 
@@
 
-Our results rule out large ($>$2$\permil$) changes in the $\delta^{18}$O of seawater and extremely hot ($>$40$^{\circ}$C) ocean temperatures as causes of the depleted $\delta^{18}$O values of Silurian carbonates.
-Instead, clumped isotopes demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality.
-This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition.
+Our results rule out large ($>2\permil$) long-term changes in the $\delta^{18}$O of seawater and extremely hot ($>40^{\circ}$C) ocean temperatures as causes of the depleted $\delta^{18}$O values of Silurian carbonates.
+Instead, clumped isotopes demonstrate that even among samples with well-preserved calcite fabrics and pristine trace metal signatures, there is a large range in preservation quality.
+This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition in an asymmetric way.
 We note that it is common to take the average $\delta^{18}$O value of carbonate samples when producing time series records of paleoclimate \citep{Bickert1997, Jaffres2007, Munnecke2003, Shields2003, Veizer2000}.
 If a carbonate sample set has undergone high-temperature fluid-buffered diagenesis, causing more altered samples to have lower $\delta^{18}$O values, the average $\delta^{18}$O of all the samples would not accurately reflect primary ocean conditions.
-The best-preserved sample in our dataset has a $\delta^{18}$O$_{\text{VPDB}}$ of -5.0$\permil$, which is close to the average $\delta^{18}$O$_{\text{VPDB}}$ value (-4.6$\permil$) of compiled low-magnesium calcites from the Silurian Period \citep{Veizer1999}.
-We expect that the same would not be true if clumped isotopes were used to identify the most pristine sample within a sample set from the Ordovician or Cambrian.
-Indeed, the most pristine sample from the upper Mississippi Valley (pre-glacial Ordovician) measured by \cite{Finnegan2011} has a $\delta^{18}$O$_{\text{VPDB}}$ of -4.9$\permil$, while the average of compiled low-magnesium calcites from the Ordovician Period is -5.9$\permil$ \citep{Veizer1999}.
+The best-preserved sample in our dataset has a $\delta^{18}$O$_{\text{VPDB}}$ of -5.0$\permil$, which is close to the average $\delta^{18}$O$_{\text{VPDB}}$ value (-4.6$\permil$) of compiled low-magnesium calcites from the Silurian Period \citep{Veizer1999}; however the same need not be true of all sample sets or intervals, particularly as one moves into older and more deeply buried strata.
+Indeed, the most pristine sample from the upper Mississippi Valley (pre-glacial Ordovician) measured by \cite{Finnegan2011} has a $\delta^{18}$O$_{\text{VPDB}}$ of -4.9$\permil$, while the average of compiled low-magnesium calcites measured globally from the Ordovician Period is -5.9$\permil$ \citep{Veizer1999}.
 
-\begin{figure}[tb]
-\centering
-\includegraphics[width=90mm]{Fig_11-01.png}
-\caption{Measurements and estimates of seawater $\delta^{18}$O through time.
-In gray, the $\delta^{18}$O$_{\text{VPDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999} and updated in 2004.
-The data are available online: $\mathtt{http://mysite.science.uottawa.ca/jveizer/isotope_data/index.html}$.
-Values from benthic foraminifera are not included because they tend to record much colder water than the other fossils.
-The black solid line represents one hypothesized time series for the $\delta^{18}$O$_{\text{VSMOW}}$ of the oceans through time based on a linear least-squares fit \citep{Veizer2000} through the skeletal calcite data \citep{Veizer1999}.
-The straight dashed black line represents the hypothesis of \cite{Gregory1981} that seawater $\delta^{18}$O$_{\text{VSMOW}}$ has remained constant through time.
-The curved dashed black line represents the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater through time as simulated by the box model in \cite{Wallmann2001}.
-All three black lines ignore short-term perturbations to seawater $\delta^{18}$O caused by glaciations.
-Finally, the black shapes are estimates of seawater $\delta^{18}$O from clumped isotope studies, measured at five different times by four separate studies.
-These clumped isotope estimates suggest that seawater $\delta^{18}$O has not changed by more than 2$\permil$ in the last 430 million years.
-These estimates also highlight that the average $\delta^{18}$O$_{\text{VPDB}}$ in a given interval offers a biased estimate of the seawater composition.}
-\label{Veizer}
-\end{figure}
-
-The results from this study join a growing set of clumped isotope studies \citep{Came2007, Dennis2013, Finnegan2011} which suggest that the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater has been fairly constant throughout much of Phanerozoic time, at around -1$\pm$1$\permil$ (Figure \ref{Veizer}).
-Several studies have suggested that high- and low-temperature alteration of the seafloor acts as a buffer for the $\delta^{18}$O of seawater \citep{Gregory1981, Gregory1991, Muehlenbachs1976, Muehlenbachs1998}.
+The results from this study join a growing set of clumped isotope studies \citep{Came2007, Dennis2013, Finnegan2011} which suggest that the $\delta^{18}$O$_{\text{VSMOW}}$ of seawater has been broadly constant throughout much of Phanerozoic time, at around -1$\pm$1$\permil$ (Fig.
+12).
+Several studies have hypothesized that high- and low-temperature alteration of the seafloor acts as a buffer for the $\delta^{18}$O of seawater \citep{Gregory1981, Gregory1991, Muehlenbachs1976, Muehlenbachs1998}.
 Observations from modern ocean drill cores suggest that the average temperature of exchange between the seafloor and seawater is 250 to 350$^{\circ}$C \citep{Muehlenbachs1998}.
@@
 However, this buffering capacity rests on the assumption that the flux of $^{18}$O exchange due to continental weathering has remained insignificant in the overall $^{18}$O budget throughout Earth’s history, and also that the average temperature of exchange between the seafloor and seawater has not changed through time.
-More recent theoretical efforts challenged these assumptions and suggested that this buffer may not be strong enough to overcome slow changes in the $\delta^{18}$O of seawater \citep{Jaffres2007}.
+More recent theoretical efforts challenged these assumptions and suggested that this buffer may not be strong enough to prevent slow changes in the $\delta^{18}$O of seawater \citep{Jaffres2007}.
 If this buffer cannot explain the relative constancy of seawater $\delta^{18}$O values through time, then the two competing controls on seawater $\delta^{18}$O, alteration of rocks at high and low temperatures, must have remained in balance over the last 500 million years.
-This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through as-yet unidentified negative feedbacks to balance changes in the other.
+This suggests that the magnitudes of high-temperature hydrothermal alteration and low-temperature weathering have not changed substantially, or that any changes in the magnitude of one process are tied through as-yet unidentified feedbacks to balance changes in the other.
 
@@
 
-We used clumped isotopes to constrain both the temperature and $\delta^{18}$O of Silurian oceans because they provide a temperature that is independent of the $\delta^{18}$O of the water from which the carbonate precipitated.
+We used clumped isotopes to constrain both the temperature and $\delta^{18}$O of Silurian oceans because this method provides a temperature that is independent of the $\delta^{18}$O of the water from which the carbonate precipitated.
 Samples were collected from well-preserved flat-lying sections on the island of Gotland, Sweden.
 By sampling well-preserved, fossil-rich rocks that potentially record a climate phenomenon associated with changing seawater temperatures, we tested the ability of clumped isotopes to resolve potentially small climate change phenomena associated with the Ireviken Event.
-Despite the overall high quality of fossiliferous materials in our Gotland sections, we were not able to resolve a secular change in temperature through the section, due to the variable preservation quality of the samples.
-We observed that trace metal concentrations have only a limited ability to detect variations in preservation among brachiopods, despite a range in brachiopod clumped isotope temperatures, suggesting that clumped isotopes are a more sensitive indicator of diagenetic alteration than trace metal compositions.
-This conclusion is supported by EBSD observations.
+Despite the overall high quality of fossiliferous materials in our Gotland sections, we were not able to resolve a robust secular change in temperature through the section, due to the variable preservation quality of the samples.
+We observed that trace metal concentrations have only a limited ability to detect variations in preservation among Gotland brachiopods, despite a range in brachiopod clumped isotope temperatures, suggesting that clumped isotopes are a more sensitive indicator of diagenetic alteration than trace metal compositions.
+This conclusion is supported by EBSD observations of fossil calcite that was recrystallized without a detectable change in trace metal concentrations.
 Micro-scale analysis of calcite fabrics, comparisons between brachiopods and micrite, and the positive relationship between temperature and $\delta^{18}$O of water calculated from the samples all support the hypothesis that the lowest-temperature samples are the most pristine with alteration of fossil material largely tied to rock-buffered diagenesis.
 Based on the most pristine materials in our dataset, we estimate that tropical Silurian oceans had a temperature of 31$\pm$2$^{\circ}$C and a $\delta^{18}$O$_{\text{VSMOW}}$ of -1.4$\pm$0.5$\permil$.
-Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the $\delta^{18}$O of seawater has remained broadly constant over time.
+Our results, along with those from other clumped isotope studies of paleoenvironment, suggest that the $\delta^{18}$O of seawater (considering the fraction stored as ice during the Pleistocene) has remained broadly constant over Phanerozoic time.
 
@@
 We thank Chi Ma for his assistance with the SEM and the E-probe, and Joel Hurowitz for help with trace element geochemistry.
-We also thank Mark Garcia and David Mann for their help with sample processing.
+We also thank Mark Garcia and David Mann for their help with sample processing and Clint Cummins for help with statistics.
 This work was supported by NSF grant EAR-1053523 and the Agouron Institute.
-
-\nocite{*}
 
